---
title: "Untitled"
author: "Finn Lindgren"
date: "2023-01-28 and 2025-10-30"
output:
  pdf_document: default
  html_document: default
header-includes:
- \newcommand{\wt}[1]{\widetilde{#1}}
- \DeclareMathOperator{\argmax}{arg\,max}
- \DeclareMathOperator{\dist}{dist}
- \newcommand{\VV}{\mathcal{V}}
- \newcommand{\BB}{\mathcal{B}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all("~/git/inlabru/")
```

## Hierarchical basis


Let $\dist(a, b)$ denote the distance between two nodes, $a$, and $b$.
Let $V_K=\{v_k;k=1,\dots,K\}$ be the set of root nodes included after step $K$
of the algorithm.

1. Let $V_1$ be a set of initial nodes, one for each disconnected component
   of the graph, and and let $\VV_1=V_1$
2. Let $d_1(v)=\min_{v_k\in V_1}\dist(v_k, v)$ and $D_1=\max_v d_1(v)$
3. Let $\wt{b}^{D_1}_{v_k}(v)= b^{D_1}_{v_k}(v)= \mathbb{I}[\text{$v$ is in the same component as $v_k$}]$,
   for each $v_k\in V_1$.
4. Let $\BB_1 = \{\wt{b}^{D_1}_{v_k};\, v_k\in V_1\}$
5. For $K=2,\dots,$ until $|\VV_{K}|=N_V$:
    a. Let $D_K=\max_v d_{K-1}(v)$
    b. Find $V_{K} = \{v;\, d_{K-1}(v) =D_K\}$ and let $\VV_K=\VV_{K-1}\cup V_K$.
    c. Let $b^{D_K}_{v_k}(v) = \max(0, D_K - \dist(v_k, v))$ for each $v_k\in\VV_K$.
    d. Normalise $\wt{b}^{D_K}_{v_k}(v) = b^{D_K}_{v_k}(v) / \sum_{v_j\in\VV_K} b^{D_k}_{v_j}(v)$
       for each $v_k\in V_K$.
    e. Add $\wt{b}^{D_K}_{v_k}$, for all $v_k\in V_K$, to the basis set:
       $\BB_{K} = \BB_{K-1} \cup \{\wt{b}^{D_K}_{v_k};\, v_k\in V_K\}$
    f. Let $d_{K}(v)=\min_{v_j\in V_K} \min(d_{K-1}(v), \dist(v_k,v))$
    
Step 5b and the subsequent steps need to handle ties, i.e. when there is no unique $v_K$.
If there are multiple candidates for $v_K$, one could consider $=\min_v \{v;d_{K-1}(v)=D_K\}$,
but this may result in extremal points for the other candidates, such that some of the
$\wt{b}^{D_K}_k(v)$ become ill-behaved at or near those points, when no basis function is non-zero.
Handling all the ties at the same time seems like a reasonable solution to this problem.

In step 5c, for $v_k\in\VV_{K-1}$,
$$
b^{D_K}_{v_k}(v) = \max(0, b^{D_{K-1}}_{v_k}(v) + D_{K} - D_{K-1}),
$$
and for $v_k\in V_K$,
$$
b^{D_K}_{v_k}(v) = \max(0, D_K - \dist(v_k,v)) .
$$
```{r}
library(ggplot2)
xy <- as.data.frame(expand.grid(x = seq(0, 1, length.out = 50), y = seq(0, 1, length.out = 50)))
B <- data.frame(
  B1 = pmax(0, 1 - sqrt(xy$x^2 + xy$y^2)),
  B2 = pmax(0, 1 - sqrt((xy$x - 1)^2 + xy$y^2)),
  B3 = pmax(0, 1 - sqrt(xy$x^2 + (xy$y - 1)^2)),
  B4 = pmax(0, 1 - sqrt((xy$x - 1)^2 + (xy$y - 1)^2))
)
ggplot(cbind(xy, B)) +
  geom_contour_filled(aes(x, y, z = (B2 + 1e-6) / (B1 + B2 + B3 + 3e-6))) +
  coord_fixed()
```



### Misc

```{r}
hole <- fm_segm(rbind(c(-2,-2), c(0,4), c(3,0)),is.bnd = TRUE)
mesh <- fmesher::fm_mesh_intersection(fmesher::fmexample$mesh, hole)
plot(mesh)

m <- fm_subdivide(mesh, 1)
B <- list()
for (method in c("graphdistance", "laplace", "graphlaplace")) {
 B[[method]] <- make_hierarchical_mesh_basis(m, method = method)
}

require("ggplot2") && require("patchwork")
print(
  ggplot(
    data =
      data.frame(
        idx = seq_len(ncol(B$graphdistance)),
        nnz = as.vector(Matrix::colSums(0 != B$graphdistance)),
        distance = as.vector(Matrix::colSums(B$graphdistance)),
        laplace = as.vector(Matrix::colSums(B$laplace)),
        graph = as.vector(Matrix::colSums(B$graphlaplace))
      )
  ) +
    geom_point(aes(x = idx, y = nnz, color = "nnz")) +
    geom_point(aes(x = idx, y = distance, color = "distance")) +
    geom_point(aes(x = idx, y = laplace, color = "laplace")) +
    geom_point(aes(x = idx, y = graph, color = "graph")) +
    scale_y_log10() +
    scale_x_log10()
)
   
idx <- seq_len(200)
theta <- qr.solve(B$laplace[, idx, drop = FALSE], m$loc[, 1])
print(
  ggplot() +
    gg(m, col = B$laplace[, idx, drop = FALSE] %*% theta, nx = 60, ny = 60) +
    geom_fm(data = m, color = ggplot2::alpha("black", 0.1), alpha = 0) +
    scale_fill_distiller(palette = "RdBu")
)
idx0 <- 1
idx <- 1:2
theta <- rep(1, length(idx))
print(
  (
    ggplot() +
      gg(m, col = B$graphdistance[, idx0, drop = TRUE], nx = 60, ny = 60) +
      geom_fm(data = m, color = ggplot2::alpha("black", 0.1), alpha = 0) +
      scale_fill_distiller(palette = "RdBu")
  ) |
    (
      ggplot() +
        gg(m, col = B$graphdistance[, idx0, drop = FALSE] / (B$graphdistance[, idx, drop = FALSE] %*% theta), nx = 60, ny = 60) +
        geom_fm(data = m, color = ggplot2::alpha("black", 0.1), alpha = 0) +
        scale_fill_distiller(palette = "RdBu")
    )
)
```



## Equivalent variational adjustment 

$$
\begin{aligned}
u &\sim N(\mu_u,Q_u^{-1}) \\
u &= Bv \\
v &\sim N(B^{-1}\mu_u,(B^\top Q_u B)^{-1}) \\
\eta &= A u = ABv \\
Q_{u|y} &= Q_u + A^\top Q_{y|u} A \\
Q_{u|y} \mu_{u|y} &= Q_u\mu_u + A^\top Q_{y|u} y = b_u \\
Q_{v|y} &= B^\top Q_u B + B^\top A^\top Q_{y|u} A B \\
Q_{v|y} \mu_{v|y} &= Q_v\mu_v + B^\top A^\top Q_{y|u} y = b_v = B^\top b_u
\end{aligned}
$$
Variational adjustment equations:
$$
\begin{aligned}
Q_{v|y} \mu_{v|y} &= b_v + \delta \\
B^\top Q_{u|y} B B^{-1} \mu_{u|y} &= B^\top b_u + \delta \\
Q_{u|y} \mu_{u|y} &= b_u + B^{-\top} \delta
\end{aligned}
$$

## Fractional approximation

Require $p\geq\alpha$.
Require $-2p-1-\lambda > -1$ for integrability at 0, i.e. $\lambda > -2p$.
Require $2p-2p-1-\lambda < -1$ for integrability at $\infty$, i.e. $\lambda < 0$.
\begin{align*}
\int_0^\infty (z^\alpha - \sum_{i=0}^p b_i z^i)^2 z^{-2p-1-\lambda} dz
&= \int_0^\infty z^{2\alpha - 2p -1 - \lambda} dz
- 2 \sum_{i=0}^p b_i \int_0^\infty z^{\alpha + i - 2p -1 - \lambda} dz
+ \sum_{i=0}^p \sum_{j=0}^p b_i b_j \int_0^\infty z^{i + j - 2p -1 - \lambda} dz \\
&= \frac{1}{2(p + \lambda) - 2\alpha}
- 2 \sum_{i=0}^p b_i \frac{1}{2(p + \lambda) - (\alpha + i)}
+ \sum_{i=0}^p \sum_{j=0}^p b_i b_j \frac{1}{2(p + \lambda) - (i + j)} \\
&= \frac{1}{2(p + \lambda) - 2\alpha}
- 2 \sum_{i=0}^p b_i \frac{1}{2(p + \lambda) - (\alpha + i)}
+ b^\top H b,
\end{align*}
where $H = \left(\frac{1}{2(p + \lambda) - (i + j)}\right)_{i,j=0}^p$.
Parsimonius choice for $\kappa>0$ was $\lambda=\alpha-\lfloor\alpha\rfloor$, but
we need $-2p<\lambda<0$ here.
Minimising wrt. $b$ gives
\begin{align*}
H b = \left(\frac{1}{2(p + \lambda) - (\alpha + i)}\right)_{i=0}^p
\end{align*}
