---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# inlabru

<!-- badges: start -->
[![CRAN Status](http://www.r-pkg.org/badges/version-last-release/inlabru)](https://cran.r-project.org/package=inlabru)
[![inlabru status badge](https://inlabru-org.r-universe.dev/badges/inlabru)](https://inlabru-org.r-universe.dev)
[![R build status](https://github.com/inlabru-org/inlabru/workflows/R-CMD-check/badge.svg)](https://github.com/inlabru-org/inlabru/actions)
[![R code coverage status](https://github.com/inlabru-org/inlabru/workflows/test-coverage/badge.svg)](https://github.com/inlabru-org/inlabru/actions)
[![lintr status](https://github.com/inlabru-org/inlabru/workflows/lint/badge.svg)](https://github.com/inlabru-org/inlabru/actions)
[![Codecov test coverage](https://codecov.io/gh/inlabru-org/inlabru/branch/devel/graph/badge.svg)](https://app.codecov.io/gh/inlabru-org/inlabru?branch=devel)
<!-- badges: end -->

The goal of [inlabru](http://inlabru.org) is to facilitate spatial modeling using integrated nested Laplace approximation via the [R-INLA package](https://www.r-inla.org).
Additionally, extends the GAM-like model class to more general nonlinear predictor
expressions, and implements a log Gaussian Cox process likelihood for 
modeling univariate and spatial point processes based on ecological survey data.
Model components are specified with general inputs and mapping methods to the
latent variables, and the predictors are specified via general R expressions,
with separate expressions for each observation likelihood model in
multi-likelihood models. A prediction method based on fast Monte Carlo sampling
allows posterior prediction of general expressions of the latent variables.
See
Fabian E. Bachl, Finn Lindgren, David L. Borchers, and Janine B. Illian (2019), inlabru: an R package for Bayesian
spatial modelling from ecological survey data, Methods in Ecology and Evolution, British Ecological Society, 10,
760--766, [doi:10.1111/2041-210X.13168](https://doi.org/10.1111/2041-210X.13168), and `citation("inlabru")`.

The [inlabru.org](http://inlabru.org) website has links to old tutorials with code examples for versions up to 2.1.13.
For later versions, updated versions of these tutorials, as well as new examples, can be found at [https://inlabru-org.github.io/inlabru/articles/](https://inlabru-org.github.io/inlabru/articles/)

## Installation

You can install the current [CRAN](https://CRAN.R-project.org) version of inlabru:
```{r cran-installation, eval = FALSE}
install.packages("inlabru")
```

You can install the latest bugfix release of inlabru from [GitHub](https://github.com/inlabru-org/inlabru) with:
```{r gh-bugfix-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("inlabru-org/inlabru", ref = "stable")
```

You can install the development version of inlabru from [GitHub](https://github.com/inlabru-org/inlabru) with
```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("inlabru-org/inlabru", ref = "devel")
```
or track the development version builds via [inlabru-org.r-universe.dev](https://inlabru-org.r-universe.dev/builds):
```{r universe-installation, eval = FALSE}
# Enable universe(s) by inlabru-org
options(repos = c(
  inlabruorg = "https://inlabru-org.r-universe.dev",
  INLA = "https://inla.r-inla-download.org/R/testing",
  CRAN = "https://cloud.r-project.org"
))

# Install some packages
install.packages("inlabru")
```

## Example

This is a basic example which shows you how fit a simple spatial Log Gaussian Cox Process (LGCP)
and predicts its intensity:
```{r special-setup, echo=FALSE}
suppressPackageStartupMessages(library(sp))
```
```{r example}
# Load libraries
library(inlabru)
library(INLA)
library(ggplot2)
bru_safe_sp(force = TRUE) # Ensures sp works without rgdal installed

# Load the data
data(gorillas, package = "inlabru")

# Construct latent model components
matern <- inla.spde2.pcmatern(gorillas$mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(0.01, 0.01)
)
cmp <- ~ mySmooth(coordinates, model = matern) + Intercept(1)
# Fit LGCP model
# This particular bru/like combination has a shortcut function lgcp() as well
fit <- bru(
  components = cmp,
  like(
    formula = coordinates ~ .,
    family = "cp",
    data = gorillas$nests,
    samplers = gorillas$boundary,
    domain = list(coordinates = gorillas$mesh)
  ),
  options = list(control.inla = list(int.strategy = "eb"))
)

# Predict Gorilla nest intensity
lambda <- predict(
  fit,
  fm_pixels(gorillas$mesh, mask = gorillas$boundary, format = "sp"),
  ~ exp(mySmooth + Intercept)
)

# Plot the result
ggplot() +
  gg(lambda) +
  gg(gorillas$nests, color = "red", size = 0.2) +
  coord_equal() +
  ggtitle("Nest intensity per km squared")
```
