---
title: "Iterative INLA method"
output:
  - rmarkdown::html_vignette
  - rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Iterative INLA method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
  - \newcommand{\wt}[1]{\widetilde{#1}}
  - \newcommand{\ol}[1]{\overline{#1}}
  - \newcommand{\wh}[1]{\widehat{#1}}
  - \newcommand{\E}{\mathbb{E}}
  - \DeclareMathOperator*{\argmax}{arg\,max}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(inlabru)
library(ggplot2)
```

## The INLA method for linear predictors

The INLA method is used to compute fast approximative posterior distribution
for Bayesian generalised additive models. The hierarchical structure of such a 
model with latent Gaussian components $\bm{u}$, covariance parameters
$\bm{\theta}$, and measured response variables $\bm{y}$, can be written as
$$
\begin{aligned}
\bm{\theta} &\sim p(\bm{\theta}) \\
\bm{u}|\bm{\theta} &\sim \mathcal{N}\!\left(\bm{\mu}_u, \bm{Q}(\bm{\theta})^{-1}\right) \\
\bm{\eta}(\bm{u}) &= \bm{A}\bm{u} \\
\bm{y}|\bm{u},\bm{\theta} & \sim p(\bm{y}|\bm{\eta}(\bm{u}),\bm{\theta})
\end{aligned}
$$
where typically each linear predictor element, $\eta_i(\bm{u})$, is linked to a
location parameter of the distribution for observation $y_i$, for each $i$,
via a (non-linear) link function $g^{-1}(\cdot)$. In the R-INLA implementation,
the observations are assumed to be conditionally independent, given $\bm{\eta}$
and $\bm{\theta}$.

## Approximate INLA for non-linear predictors

The premise for the inlabru method for non-linear predictors is to build on
the existing implementation, and only add a linearisation step. The properties of
the resulting approximation will depend on the nature of the non-linearity.

Let $\wt{\bm{\eta}}(\bm{u})$ be a non-linear predictor, i.e. a deterministic function of $\bm{u}$, $$ \wt{\bm{\eta}} (\bm{u}) = \textsf{fcn} (\bm{u}), $$
and let
$\ol{\bm{\eta}}(\bm{u})$ be the 1st order Taylor approximation at $\bm{u}_0$,
$$
\ol{\bm{\eta}}(\bm{u})
= \wt{\bm{\eta}}(\bm{u}_0) + \bm{B}(\bm{u} - \bm{u}_0)
= \left[\wt{\bm{\eta}}(\bm{u}_0) - \bm{B}\bm{u}_0\right] + \bm{B}\bm{u}
,
$$
where $\bm{B}$ is the derivative matrix for the non-linear predictor, evaluated
at $\bm{u}_0$.
Hence, we define
$$
\begin{aligned}
\bm{y} | \bm{u}, {\bm{\theta}} &\overset{d}{=} \bm{y} | \wt{\bm{\eta}}(\bm{u}), {\bm{\theta}} \\
&\sim p (\bm{y} | g^{-1}[\wt{\bm{\eta}}(\bm{u})], {\bm{\theta}})\\
\end{aligned}
$$
The non-linear observation model
$p(\bm{y}|g^{-1}[\wt{\bm{\eta}}(\bm{u})],\bm{\theta})$
is approximated by replacing the non-linear predictor with its linearisation,
so that the linearised model is defined by

$$
\ol{p}(\bm{y}|\bm{u},\bm{\theta})
=
p(\bm{y}|\ol{\bm{\eta}}(\bm{u}),\bm{\theta})
=
p(\bm{y}|g^{-1}[\ol{\bm{\eta}}(\bm{u})],\bm{\theta})
\approx
p(\bm{y}|g^{-1}[\wt{\bm{\eta}}(\bm{u})],\bm{\theta})
=
p(\bm{y}|\wt{\bm{\eta}}(\bm{u}),\bm{\theta})
=
\wt{p}(\bm{y}|\bm{u},\bm{\theta})
$$
The non-linear model posterior is factorised as
$$
\wt{p}(\bm{\theta},\bm{u}|\bm{y}) = \wt{p}(\bm{\theta}|\bm{y})\wt{p}(\bm{u}|\bm{y},\bm{\theta}),
$$
and the linear model approximation is factorised as
$$
\ol{p}(\bm{\theta},\bm{u}|\bm{y}) = \ol{p}(\bm{\theta}|\bm{y})\ol{p}(\bm{u}|\bm{y},\bm{\theta}).
$$

### Fixed point iteration

The remaining step of the approximation is how to choose the linearisation
point $\bm{u}_*$.
For a given linearisation point $\bm{v}$, INLA will compute the posterior mode for $\bm{\theta}$,
$$
\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_\bm{v} ( {\bm{\theta}} | \bm{y} ),
$$
and the joint conditional posterior mode for $\bm{u}$,
$$
\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_\bm{v} ( \bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}} ) .
$$
[FL note: need to show that when $\bm{u}_*$ is the fixed point of this functional, then the corresponding $\bm{\theta}_*$ is equal to the true posterior mode for $\bm{\theta}$, and $\wh{\bm{u}}_*$ is the true conditional posterior mode for $\bm{u}$.]

Define the Bayesian estimation functional^[
Potential other choices for $f(\cdot)$ include the posterior expectation
$\ol{E}(\bm{u}|\bm{y})$ and the marginal conditional modes, 
$$
\left\{\argmax_{u_i} \ol{p}_{\bm{v}}(u_i|\bm{y}),\,i=1,\dots,n\right\},
$$
which was used in `inlabru` up to version 2.1.15, which caused problems for some nonlinear models.
From version 2.2.3, the joint conditional mode is used,
$$
\argmax_{\bm{u}} \ol{p}_{\bm{v}}(\bm{u}|\bm{y},\wh{\bm{\theta}}_{\bm{v}}),
$$
where $\wh{\bm{\theta}}=\argmax_{\bm{\theta}} \ol{p}_{\bm{v}}(\bm{\theta}|\bm{y})$.
]
$$
f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
$$
and let $f(p)=(\wh{\bm{\theta}},\wh{\bm{u}})$ denote the corresponding posterior modes for the true posterior distribution,
$$
\begin{aligned}
    \wh{{\bm{\theta}}} &= \argmax_{\bm{\theta}} p ( {\bm{\theta}} | \bm{y} ), \\
    \wh{\bm{u}} &= \argmax_{\bm{u}} p (\bm{u} | \bm{y}, \wh{{\bm{\theta}}}).
\end{aligned}
$$

The fixed point $(\bm{\theta}_*,\bm{u}_*)=f(\ol{p}_{\bm{u}_*})$ should ideally be close to $(\wh{\bm{\theta}},\wh{\bm{u}})$, i.e. close to the true marginal/conditional posterior mode. We can achieve this for the conditional
latent mode, so that $\bm{u}_*=\argmax_{\bm{u}} p (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{u}_*})$.

We therefore seek the latent vector $\bm{u}_*$ that generates the fixed point of the
functional, so that $(\bm{\theta}_*,\bm{u}_*)=f(\ol{p}_{\bm{u}_*})$.


One key to the fixed point iteration is that the observation model is linked to
$\bm{u}$ only through the non-linear predictor $\wt{\bm{\eta}}(\bm{u})$, since
this leads to a simplified line search method below.

<!-- intentional to start with 0 ?-->
0. Let $\bm{u}_0$ be an initial linearisation point for the latent variables 
   obtained from the initial INLA call. Iterate the following steps for $k=0,1,2,...$
1. Compute the predictor linearisation at $\bm{u}_0$.
2. Compute the linearised INLA posterior $\ol{p}_{\bm{u}_0}(\bm{\theta}|\bm{y})$.
3. Let $(\bm{\theta}_1,\bm{u}_1)=(\wh{\bm{\theta}}_{\bm{u}_0},\wh{\bm{u}}_{\bm{u}_0})=f(\ol{p}_{\bm{u}_0})$ be the initial candidate for new
   linearisation point.
4. Let $\bm{v}_\alpha=(1-\alpha)\bm{u}_1+\alpha\bm{u}_0$, and find the value
   $\alpha$ minimises $\|\wt{\eta}(\bm{v}_\alpha)-\ol{\eta}(\bm{u}_1)\|$.
5. Set the new linearisation point $\bm{u}_0$ equal to $\bm{v}_\alpha$ and repeat from step 1,
   unless the iteration has converged to a given tolerance.

#### Line search

In step 4, we would ideally want an $\alpha$ to be  
$$ \argmax_{\alpha} [\ln p(\bm{u}|\bm{y},\bm{\theta}_1)].$$
As a simpler alternative, we will consider loss function such as $\|\wt{\eta}(\bm{u}_\alpha)-\ol{\eta}(\bm{u}_1)\|$
that only depend on the nonlinear and linearised predictor expressions (and other known quantities, given $\bm{u}_0$)
instead as part of an approximate line search. This avoids many potentially
expensive evaluations of the true posterior conditional log-density. We evaluate
$\wt{\bm{\eta}}_1=\wt{\bm{\eta}}(\bm{u}_1)$ and make use of the linearised predictor
information. Let $\wt{\bm{\eta}}_\alpha=\wt{\bm{\eta}}(\bm{u}_\alpha)$ and $\ol{\bm{\eta}}_\alpha=\ol{\bm{\eta}}(\bm{u}_\alpha)=(1-\alpha)\wt{\bm{\eta}}(\bm{u}_0)+\alpha\ol{\bm{\eta}}(\bm{u}_1)$.
In other words, $\alpha=0$ 
An exact line search would minimise $\|\wt{\bm{\eta}}_\alpha-\ol{\bm{\eta}}_1\|$. 
Instead, we define a quadratic approximation to the
non-linear predictor as a function of $\alpha$,
$$
\breve{\bm{\eta}}_\alpha =
\ol{\bm{\eta}}_\alpha + \alpha^2 (\wt{\bm{\eta}}_1 - \ol{\bm{\eta}}_1)
$$
and minimise the quartic polynomial in $\alpha$,
$$
\begin{aligned}
\|\breve{\bm{\eta}}_\alpha-\ol{\bm{\eta}}_1\|^2
&=
\| (\alpha-1)(\ol{\bm{\eta}}_1 - \ol{\bm{\eta}}_0) + \alpha^2 (\wt{\bm{\eta}}_1 - \ol{\bm{\eta}}_1) \|^2
.
\end{aligned}
$$
If initial expansion and contraction steps are carried out, leading to an initial
guess of $\alpha=\gamma^k$, where $\gamma>1$ is a scaling factor (see `?bru_options`, `bru_method$factor`) and $k$ is the
(signed) number of expansions and contractions, the quadratic expression is replaced by
$$
\begin{aligned}
\|\breve{\bm{\eta}}_\alpha-\ol{\bm{\eta}}_1\|^2
&=
\| (\alpha-1)(\ol{\bm{\eta}}_1 - \ol{\bm{\eta}}_0) + \frac{\alpha^2}{\gamma^{2k}} (\wt{\bm{\eta}}_{\gamma^k} - \ol{\bm{\eta}}_{\gamma^k}) \|^2
,
\end{aligned}
$$
which is minimised on the interval $\alpha\in[\gamma^{k-1},\gamma^{k+1}]$.

We are working in progress on the user-defined minimisation of the loss function 
term, where inlabru currently uses 
$\| \wt{\bm{\eta}} (\bm{u}_\alpha) - \ol{\bm{\eta}}(\bm{u}_1)\|$. We hope to 
improve the the accuracy but at the very least objectify this term as diagnostics. 
There are two potential options. First option is 

$$ \sum_i E_{\bm{u}\sim \ol{p}(\bm{u}|\bm{y})} \frac{(|\ol{\bm{\eta}}_i-\wt{\bm{\eta}}_i|^2)}{\mathrm{Var}_{\bm{u}\sim \ol{p}(\bm{u}|\bm{y})}(\ol{\bm{\eta}}_i)},$$ 

this is scaling with the variance and somewhat similar to Dawid-Sebastiani score. 
Another option is 
$$ E_{\bm{u}\sim \ol{p}(\bm{u}|\bm{y})} (\log \frac{\ol{p}(\bm{u} |\bm{y},{\bm{\theta}})}{\wt{p}(\bm{u}|\bm{y},{\bm{\theta}})}),$$ 

this is basically the approximate version of Kullback–Leibler divergence term. 

A potential improvement of step 4 might be to also take into account the prior
distribution for $\bm{u}$ as a minimisation penalty, to avoid moving further than
would be indicated by a full likelihood optimisation.

#### Accuracy
We wish to assess how accurate the approximation is. Thus, we compare $({\bm{\theta}}, \bm{u} | \bm{y} )$ and $\overline{({\bm{\theta}}, \bm{u} |\bm{y})}$. 
With Bayes' theorem, 
$$
\begin{aligned}
    p(\bm{u}|\bm{y},{\bm{\theta}}) &= \frac{p(\bm{u},\bm{y},{\bm{\theta}})}{p(\bm{y},{\bm{\theta}})} \\
    &= \frac{p(\bm{y}|\bm{u},{\bm{\theta}}) p(\bm{u}|{\bm{\theta}})}{p(\bm{y},{\bm{\theta}})},
\end{aligned}
$$
and by chain rule, 
$$ \nabla_{\bm{u}} \ln p (\bm{y} | {\bm{\theta}}, \wt{\bm{\eta}}(\bm{u}_*)) \biggr\rvert_{\bm{u}_*} (\bm{u}-\bm{u}_*) = \nabla_{\bm{\eta}} \ln p (\bm{y} | {\bm{\theta}}, \bm{\eta}) \cdot \frac{d\bm{\eta}}{d\bm{u}} \biggr\rvert_{\bm{u}_*} (\bm{u}-\bm{u}_*). $$

Hence, using Taylor's expansion,
$$
\begin{aligned}
    \ln p(\bm{y}|\bm{u},{\bm{\theta}}) &= \ln p (\bm{y}| {\bm{\theta}}, \wt{\bm{\eta}}(\bm{u})) \\
    &= \ln p (\bm{y}|{\bm{\theta}},\wt{\bm{\eta}}(\bm{u}_*))  \\ 
    &\qquad +\nabla_{\wt{\bm{\eta}}} \ln p (\bm{y} | {\bm{\theta}}, \wt{\bm{\eta}}(\bm{u}_*)) \cdot (\wt{\bm{\eta}}(\bm{u}) - \wt{\bm{\eta}}(\bm{u}_*)) \\
    &\qquad + \frac{1}{2!} (\wt{\bm{\eta}}(\bm{u}) - \wt{\bm{\eta}}(\bm{u}_*))^\intercal [\nabla^2_{\wt{\bm{\eta}}} \ln p (\bm{y} | {\bm{\theta}})](\wt{\bm{\eta}}(\bm{u}) - \wt{\bm{\eta}}(\bm{u}_*))  + \mathsf{h.o.t},
\end{aligned}
$$

where $\nabla^2$ is the Hessian. 
We want to compare this expression with the Taylor's expansion of $\ln p (\bm{y} | {\bm{\theta}}, \ol{\bm{\eta}}(\bm{u}))$. Basically, we are looking for the expectation of the 2nd order term, i.e. $E(\textrm{Hessian})$, between $\ln p(\bm{y}|{\bm{\theta}},\wt{\bm{\eta}}(\bm{u}))$ and $\ln p(\bm{y}|{\bm{\theta}},\ol{\bm{\eta}}(\bm{u}))$. It may require to look into the 3rd order term to have some control over all the rest higher order terms. We would tell how different the terms are.

#### Prior, Data and Posterior
Here are some thoughts on the spectrum of models, namely, weak, somethings in between, and strong models. For the weak models, prior is strong, given the Gaussian in INLA, data is weak, then the posterior ends up Gaussian. Strong model is Gaussian because of a lot of data, then prior is weak, posterior ends up Gaussian again. What lies in between is problematic, which we have to investigate. 

Most of the time, we only consider asymptoticity but the data in real world are most likely lying in between. In particular, low count models and binary models are the worst. For example, t-distribution in INLA, Hessian not positive definite, inlabru forces it to be positive definite for linearisation model. The question is how wrong it can be to force it in a point process setting. 

#### Initialisation
On a side note, one might be concerned about initalisation at, or convergence to, a saddle point. Although it is not implemented in inlabru, we want to talk about the technicality how we define the initial linearisation point $u_0$.

Generally speaking, any values of $u_0$ work except the case that the gradient evaluated at $u_0$ is $\vec{0}$ because the linearisation point will never moves away. For multiple likelihood models, it can be tackled with another likelihood. For single likelihood, it is a saddle point problem. What even worse is the condition of multimodal property, it is difficult to obtain an unique solution. Therefore, we have to get around with and re-parameterise it. Here are some choices for reparameterisation, 

$$ 
\begin{aligned}
\bm{\eta}_1 &= \bm{u} \\
\bm{\eta}_2 &= \beta \bm{u} \\
\bm{\eta}_3 &= e^\beta \bm{u} \\
\bm{\eta}_4 &= F_\beta^{-1} ( \Phi(z_\beta)) \bm{u}, \quad z_{\beta} \sim \mathsf{N}(0,1) \\
\end{aligned}
$$


