---
title: "Installation of INLA with SingularityCE on HPC"
output:
  rmarkdown::html_vignette:
  rmarkdown::pdf_document:
vignette: >
  %\VignetteIndexEntry{SingularityCE}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
  - \newcommand{\wt}[1]{\widetilde{#1}}
  - \newcommand{\ol}[1]{\overline{#1}}
  - \newcommand{\wh}[1]{\widehat{#1}}
  - \DeclareMathOperator*{\argmax}{arg\,max}
  - \newcommand{\proper}[1]{\mathsf{#1}}
  - \newcommand{\pExp}{\proper{Exp}}
  - \newcommand{\pN}{\proper{N}}
  - \newcommand{\pPo}{\proper{Po}}
  - \newcommand{\pGa}{\proper{Ga}}
  - \newcommand{\pE}{\proper{E}}
  - \newcommand{\pP}{\proper{P}}
  - \newcommand{\pVar}{\proper{Var}}
---
High Performance Computing (HPC) is surging in popularity. Building INLA package
on the HPC systems with the Anaconda environment sometimes can be tricky due to 
the Linux distributions and versions. This article is to suggest a workaround 
with container images using [SingularityCE](https://github.com/sylabs/singularity)
(formerly known as Singularity) because [Docker](https://www.docker.com/) 
requires root access and is usually not available on HPC. SingularityCE is a 
workaround to pull Docker image on HPC. Despite its convenience, the image size 
can be large. 

## Installation Procedures 

1. Check if [SingularityCE](https://github.com/sylabs/singularity/blob/main/INSTALL.md)
   is available on your HPC. If not, ask the administrator to install it. Type 
   on your HPC terminal to check if it is installed. 
```{r singularity check, eval=FALSE}
$ singularity --version
```

2. Connect to the HPC Terminal and type the code below to pull the 
[rocker/geospatial](https://rocker-project.org/images/versioned/rstudio.html) 
sif file with the latest R version.

```{r singularity, eval=FALSE}
# To pull the docker image 
$ singularity pull name_your_containter.sif docker://rocker/geospatial:latest
# once the sif is downloaded, get into an interactive shell 
$ singularity shell name_your_containter.sif
# Inside the interactive shell, one can install INLA on a personal library path
$ R --verbose 
# Now, in a R environment
> install.packages("INLA") 
# One will be asked to choose your CRAN Mirror and create a personal library 
# path.
> install.packages("inlabru") 
# quit R
> q()
```
Note: Using `inla. setOption(num. threads=ncpu)` to set the number of cpu (ncpu) since 
INLA does not set it automatically. The ncpu also directly links to RAM 
requirement.

