---
title: "LGCPs - An example in space and time"
author: "Fabian E. Bachl and Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - An example in space and time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
```

Introduction
----------------
For this vignette we are going to be working with a dataset obtained from 
the `R` package `MRSea`. We will set up a LGCP with a spatio-temporal SPDE model to estimate species distribution.


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(ggplot2)
```

Get the data
-----------------------------------
Load the dataset, that has coordinates in UTM in kilometres:

```{r }
mrsea <- inlabru::mrsea
```

The points (representing animals) and the sampling regions of this dataset are associated with a season. Let's have a look at the observed points and sampling regions for all seasons:

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() +
  geom_fm(data = mrsea$mesh) +
  gg(mrsea$boundary) +
  gg(mrsea$samplers) +
  gg(mrsea$points, size = 0.5) +
  facet_wrap(~season) +
  ggtitle("MRSea observation seasons")
```


Fitting the model
-----------------------------------

Fit an LGCP model to the locations of the animals. In this example we will employ a spatio-temporal SPDE. Note how the `group` and `ngroup` parameters are employed to let the SPDE model know about the name of the time dimension (season) and the total number of distinct points in time. The point process likelihood is constructed by the `lgcp()` function, which is a wrapper around the `bru_obs(..., model = "cp")` and `bru()` functions. Like for ordinary spatial point process
models, the `samplers` argument specifies the observation region/set, in this case
combinations spatial lines and seasons. The `domain` argument specifies the
spatio-temporal function space to use when constructing the numerical integration
scheme needed for the point process likelihood evaluation.

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
matern <- inla.spde2.pcmatern(mrsea$mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(10, 0.01)
)

cmp <- ~ Intercept(1) +
  mySmooth(
    geometry,
    model = matern,
    group = season,
    ngroup = 4
  )

fit <- bru(
  cmp,
  bru_obs(
    geometry + season ~ .,
    family = "cp",
    data = mrsea$points,
    samplers = mrsea$samplers,
    domain = list(
      geometry = mrsea$mesh,
      season = seq_len(4)
    )
  )
)
```

Predict and plot the intensity for all seasons:

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ppxl <- fm_pixels(mrsea$mesh, mask = mrsea$boundary, format = "sf")
ppxl_all <- fm_cprod(ppxl, data.frame(season = seq_len(4)))

lambda1 <- predict(
  fit,
  ppxl_all,
  ~ data.frame(season = season, lambda = exp(mySmooth + Intercept))
)
```
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
pl1 <- ggplot() +
  gg(lambda1, geom = "tile", aes(fill = q0.5)) +
  gg(mrsea$points, size = 0.3) +
  facet_wrap(~season) +
  coord_sf()
pl1
```

Integration points
-----------------------------------

The `inlabru` point process model, `lgcp()` or `bru_obs(..., model = "cp")`,
knows how to construct the numerical integration
scheme for the LGCP likelihood.  To see what happens internally, we can also
call the internal functions directly to see
what the integration scheme will look like,
using the `fm_int()` function with the same `domain` and `samplers` arguments as
in the previous `lgcp()` call.
Note that omitting
the `season` dimension from `domain` would lead to aggregation of all sampling
regions over time.

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ips <- fm_int(
  domain = list(geometry = mrsea$mesh, season = 1:4),
  samplers = mrsea$samplers
)
```
Plot the integration points:
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ggplot() +
  geom_fm(data = mrsea$mesh) +
  gg(ips, aes(size = weight)) +
  scale_size_area(max_size = 1) +
  facet_wrap(~season)
```
