---
title: "LGCPs - Multiple Likelihoods"
author: "Fabian E. Bachl"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Multiple Likelihoods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  cache = TRUE
)
```

Introduction
----------------
For this vignette we are going to be working with the inlabru's ´gorillas_sf´ dataset which was originally obtained from 
the `R` package `spatstat`. The data set contains two types of gorillas nests which are marked as either major or minor. We will set up a multi-likelihood model for these nests which creates two spatial LGCPs that share a common intercept but have employ different spatial smoothers.


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(ggplot2)
bru_safe_sp(force = TRUE)
```

Get the data
-----------------------------------
For the next few practicals we are going to be working with a dataset obtained from 
the `R` package `spatstat`, which contains the locations of 647 gorilla nests. We load the 
dataset like this:

```{r }
data(gorillas_sf, package = "inlabru")
```

Plot the nests and visualize the group membership (major/minor) by color:

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() +
  gg(gorillas_sf$mesh) +
  gg(gorillas_sf$nests, aes(color = group)) +
  gg(gorillas_sf$boundary, alpha = 0) +
  ggtitle("Gorillas nests and group membership")
```


Fiting the model
-----------------------------------

First, we define all components that enter the joint model. That is, the intercept that is common to both LGCPs and
the two different spatial smoothers, one for each nest group.

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
matern <- inla.spde2.pcmatern(gorillas_sf$mesh,
  prior.range = c(0.1, 0.01),
  prior.sigma = c(1, 0.01)
)

cmp <- ~
  Common(geometry, model = matern) +
    Difference(geometry, model = matern) +
    Intercept(1)
```

Given these components we define the linear predictor for each of the likelihoods.
(Using "." indicates a pure additive model, and one can use include/exclude
options for `like()` to indicate which components are actively involved in each model.)
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
fml.major <- geometry ~ Intercept + Common + Difference / 2
fml.minor <- geometry ~ Intercept + Common - Difference / 2
```

Setting up the Cox process likelihoods is easy in this example. Both nest types were observed within the same window:
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
lik_minor <- like("cp",
  formula = fml.major,
  data = gorillas_sf$nests[gorillas_sf$nests$group == "major", ],
  samplers = gorillas_sf$boundary,
  domain = list(geometry = gorillas_sf$mesh)
)
lik_major <- like("cp",
  formula = fml.minor,
  data = gorillas_sf$nests[gorillas_sf$nests$group == "minor", ],
  samplers = gorillas_sf$boundary,
  domain = list(geometry = gorillas_sf$mesh)
)
```
... which we provide to the ´bru´ function.
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
jfit <- bru(cmp, lik_major, lik_minor,
  options = list(
    control.inla = list(int.strategy = "eb"),
    bru_max_iter = 1
  )
)
```

```{r, warning=FALSE,fig.width=8}
library(patchwork)
pl.major <- ggplot() +
  gg(gorillas_sf$mesh,
    mask = gorillas_sf$boundary,
    col = exp(jfit$summary.random$Common$mean)
  )
pl.minor <- ggplot() +
  gg(gorillas_sf$mesh,
    mask = gorillas_sf$boundary,
    col = exp(jfit$summary.random$Difference$mean)
  )
(pl.major + scale_fill_continuous(trans = "log")) +
  (pl.minor + scale_fill_continuous(trans = "log")) &
  theme(legend.position = "right")
```


## Rerunning

Rerunning with the previous estimate as starting point sometimes improves the
accuracy of the posterior distribution estimation.

```{r, warning=FALSE}
jfit0 <- jfit
jfit <- bru_rerun(jfit)
```


```{r, warning=FALSE,fig.width=8}
library(patchwork)
pl.major <- ggplot() +
  gg(gorillas_sf$mesh,
    mask = gorillas_sf$boundary,
    col = exp(jfit$summary.random$Common$mean)
  )
pl.minor <- ggplot() +
  gg(gorillas_sf$mesh,
    mask = gorillas_sf$boundary,
    col = exp(jfit$summary.random$Difference$mean)
  )
(pl.major + scale_fill_continuous(trans = "log")) +
  (pl.minor + scale_fill_continuous(trans = "log")) &
  theme(legend.position = "right")
```

```{r}
summary(jfit0)
```

```{r}
summary(jfit)
```

## Single-likelihood version

In this particular model, we can also rewrite the problem as a single point process
over a product domain over space and `group`.  In versions `<= 2.7.0`, the integration
domain had to be numeric, so we convert the group variable to a 0/1 variable,
`group_major <- group == "major"`, which is also useful in the predictor expression:
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
fml.joint <-
  geometry + group_major ~ Intercept + Common + (group_major - 0.5) * Difference
```

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
gorillas_sf$nests$group_major <- gorillas_sf$nests$group == "major"
lik_joint <- like("cp",
  formula = fml.joint,
  data = gorillas_sf$nests,
  samplers = gorillas_sf$boundary,
  domain = list(
    geometry = gorillas_sf$mesh,
    group_major = c(0, 1)
  )
)
```

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
jfit_joint <- bru(cmp, lik_joint,
  options = list(
    control.inla = list(int.strategy = "eb") # Approximate for faster vignette
  )
)
```

Plotting the ratios of `exp(Common)` and `exp(Difference)` between the new fit and
the old confirms that the results are the same up to small numerical differences.
```{r, warning=FALSE,fig.width=8}
library(patchwork)
pl.major <- ggplot() +
  gg(gorillas_sf$mesh,
    mask = gorillas_sf$boundary,
    col = exp(jfit_joint$summary.random$Common$mean -
      jfit$summary.random$Common$mean)
  )
pl.minor <- ggplot() +
  gg(gorillas_sf$mesh,
    mask = gorillas_sf$boundary,
    col = exp(jfit_joint$summary.random$Difference$mean -
      jfit$summary.random$Difference$mean)
  )
(pl.major + scale_fill_continuous(trans = "log")) +
  (pl.minor + scale_fill_continuous(trans = "log")) &
  theme(legend.position = "right")
```
