---
title: "LGCPs - Spatial covariates"
author: "David Borchers and Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Spatial covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
.cache_dir <- file.path("2d_lgcp_covars_cache")
.cache_recompute <- !file.exists(.cache_dir)
if (.cache_recompute) {
  dir.create(.cache_dir, recursive = TRUE)
}
.vignette_cache <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```
Set things up

```{r results="hide",warning=FALSE,message=FALSE}
library(INLA)
library(inlabru)
library(fmesher)
library(RColorBrewer)
library(ggplot2)
library(patchwork)
bru_options_set(control.compute = list(dic = TRUE)) # Activate DIC output
```

Introduction
----------------------------------

We are going to fit spatial models to the gorilla data, using factor and continuous explanatory
variables in this practical. We will fit one using the factor variable `vegetation`, the other using
the continuous covariate `elevation`

(Jump to the bottom of the practical if you want to start gently with a 1D example!)

Get the data
-----------------------------------

```{r }
data(gorillas_sf, package = "inlabru")
```

This dataset is a list (see `help(gorillas_sf)` for details. Extract the objects
you need from the list, for convenience:

```{r }
nests <- gorillas_sf$nests
mesh <- gorillas_sf$mesh
boundary <- gorillas_sf$boundary
gcov <- gorillas_sf_gcov()
```

Factor covariates
-----------------------------------

Look at the vegetation type, nests and boundary:


```{r warning=FALSE,message=FALSE}
ggplot() +
  gg(gcov$vegetation) +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "white", cex = 0.5)
```

Or, with the mesh:


```{r warning=FALSE,message=FALSE}
ggplot() +
  gg(gcov$vegetation) +
  gg(mesh) +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "white", cex = 0.5)
```

#### A model with vegetation type only

It seems that vegetation type might be a good predictor because nearly all the nests fall in 
vegetation type `Primary`. So we construct a model with vegetation type as a fixed effect.
To do this, we need to tell 'lgcp' how to find the vegetation type at any point in
space, and we do this by creating model components with a fixed effect that we
call `vegetation` (we could call it
anything), as follows:


```{r warning=FALSE,message=FALSE}
comp_veg <- geometry ~ vegetation(gcov$vegetation, model = "factor_full") - 1
```

Notes:
* We need to tell 'lgcp' that this is a factor fixed effect, which we do 
with `model="factor_full"`, giving one coefficient for each factor level.
* We need to be careful about overparameterisation when using factors.
Unlike regression models like 'lm()', 'glm()' or 'gam()', 'lgcp()',
`inlabru` does not automatically remove the first level and absorb it into
an intercept. Instead, we can either use `model="factor_full"` without an
intercept, or `model="factor_contrast"`, which does remove the first level.

```{r warning=FALSE,message=FALSE}
comp_veg_alt <-
  geometry ~ vegetation(gcov$vegetation, model = "factor_contrast") +
  Intercept(1)
```

Fit the model as usual, and increasing the integration scheme resolution to
better match the covariate:
```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
mesh_int <- fm_subdivide(mesh, 2)
fit_veg <- lgcp(
  comp_veg,
  nests,
  samplers = boundary,
  domain = list(geometry = mesh_int)
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
mesh_int <- fm_subdivide(mesh, 2)
.cache_file <- file.path(.cache_dir, "fit_veg.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_veg <- readRDS(file = .cache_file)
} else {
  fit_veg <- lgcp(
    comp_veg,
    nests,
    samplers = boundary,
    domain = list(geometry = mesh_int)
  )
  saveRDS(fit_veg, .cache_file)
}
```

Predict the intensity, and plot the median intensity surface. (In older
versions, predicting takes some time because we did not have vegetation values
outside the mesh so 'inlabru' needed to predict these first. Since v2.0.0, the
vegetation has been pre-extended.)

The `predict` function of `inlabru` takes into its `data` argument an `sf`
object or other object supported by the predictor evaluation code (for
non-geographical data, typically a `data.frame`). We can use the `inlabru`
function `pixels` to generate an `sf` object with points only within the
boundary, using its `mask` argument, as shown below.


```{r warning=FALSE,message=FALSE}
pred.df <- fm_pixels(mesh, mask = boundary)
intensity_veg <- predict(fit_veg, pred.df, ~ exp(vegetation))

# gg() with sf points and geom = "tile" plots a raster
ggplot() +
  gg(intensity_veg, geom = "tile") +
  gg(boundary, alpha = 0, lwd = 2) +
  gg(nests, color = "DarkGreen")
```

Not surprisingly, given that most nests are in `Primary` vegetation, the high
density is in this vegetation. But there are substantial patches of predicted
high density that have no nests, and some areas of predicted low density that
have nests. What about the estimated abundance (there are really 647 nests
there):

```{r warning=FALSE,message=FALSE}
ips <- fm_int(mesh_int, boundary)
Lambda_veg <- predict(fit_veg, ips, ~ sum(weight * exp(vegetation)))
Lambda_veg
```

#### A model with vegetation type and a SPDE type smoother
Lets try to `explain` the pattern in nest distribution that is not captured by 
the vegetation covariate, using an SPDE:


```{r results='hide',warning=FALSE,message=FALSE}
pcmatern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(1, 0.01),
  prior.range = c(0.1, 0.01)
)

comp_veg_field <- geometry ~
  -1 +
  vegetation(gcov$vegetation, model = "factor_full") +
  field(geometry, model = pcmatern)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit_veg_field <- lgcp(comp_veg_field,
  nests,
  samplers = boundary,
  domain = list(geometry = mesh_int)
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_veg_field.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_veg_field <- readRDS(file = .cache_file)
} else {
  fit_veg_field <- lgcp(comp_veg_field,
    nests,
    samplers = boundary,
    domain = list(geometry = mesh_int)
  )
  saveRDS(fit_veg_field, .cache_file)
}
```

And plot the posterior median intensity surface

```{r warning=FALSE,message=FALSE}
intensity_veg_field <-
  predict(
    fit_veg_field,
    pred.df,
    ~ exp(field + vegetation),
    n.samples = 1000
  )

ggplot() +
  gg(intensity_veg_field, aes(fill = q0.5), geom = "tile") +
  gg(boundary, alpha = 0, lwd = 2) +
  gg(nests)
```

... and the expected integrated intensity (mean of abundance)


```{r warning=FALSE,message=FALSE}
Lambda_veg_field <- predict(
  fit_veg_field,
  fm_int(mesh_int, boundary),
  ~ sum(weight * exp(field + vegetation))
)
Lambda_veg_field
```

Look at the contributions to the linear predictor from the SPDE and from
vegetation:
```{r warning=FALSE,message=FALSE}
lp_veg_field <- predict(fit_veg_field, pred.df, ~ list(
  smooth_veg = field + vegetation,
  smooth = field,
  veg = vegetation
))
```

The function `scale_fill_gradientn` sets the scale for the plot legend. Here we
set it to span the range of the three linear predictor components being plotted
(medians are plotted by default).


```{r warning=FALSE,message=FALSE}
lprange <- range(
  lp_veg_field$smooth_veg$median,
  lp_veg_field$smooth$median,
  lp_veg_field$veg$median
)
csc <- scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), limits = lprange)

plot.lp_veg_field <- ggplot() +
  gg(lp_veg_field$smooth_veg, geom = "tile") +
  csc +
  gg(boundary, alpha = 0) +
  ggtitle("field + vegetation")

plot.lp_veg_field.spde <- ggplot() +
  gg(lp_veg_field$smooth, geom = "tile") +
  csc +
  gg(boundary, alpha = 0) +
  ggtitle("field")

plot.lp_veg_field.veg <- ggplot() +
  gg(lp_veg_field$veg, geom = "tile") +
  csc +
  gg(boundary, alpha = 0) +
  ggtitle("vegetation")

(plot.lp_veg_field + plot.lp_veg_field.spde + plot.lp_veg_field.veg) +
  plot_layout(ncol = 3, guides = "collect") &
  guides(fill = guide_legend("value")) &
  theme(legend.position = "right")
```

#### A model with SPDE only

Do we need vegetation at all? Fit a model with only an SPDE + Intercept, and
choose between models on the basis of DIC, using `deltaIC()`.

```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
comp_field <- geometry ~ field(geometry, model = pcmatern) + Intercept(1)
fit_field <- lgcp(comp_field,
  data = nests,
  samplers = boundary,
  domain = list(geometry = mesh_int),
  options = list(control.compute(dic = TRUE))
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
comp_field <- geometry ~ field(geometry, model = pcmatern) + Intercept(1)
.cache_file <- file.path(.cache_dir, "fit_field.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_field <- readRDS(file = .cache_file)
} else {
  fit_field <- lgcp(comp_field,
    nests,
    samplers = boundary,
    domain = list(geometry = mesh_int),
    options = list(control.compute(dic = TRUE))
  )
  saveRDS(fit_field, .cache_file)
}
```

```{r results='hide',warning=FALSE,message=FALSE}
intensity_field <- predict(fit_field, pred.df, ~ exp(field + Intercept))

ggplot() +
  gg(intensity_field, geom = "tile") +
  gg(boundary, alpha = 0) +
  gg(nests)
```
```{r warning=FALSE,message=FALSE}
Lambda_field <- predict(
  fit_field,
  fm_int(mesh_int, boundary),
  ~ sum(weight * exp(field + Intercept))
)
Lambda_field
```
```{r warning=FALSE,message=FALSE}
knitr::kable(deltaIC(fit_veg, fit_veg_field, fit_field, criterion = c("DIC")))
```

#### CV and SPDE parameters for Model 2

We are going with Model `fit_veg_field`. Lets look at the spatial distribution of the 
coefficient of variation

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() +
  gg(intensity_veg_field, aes(fill = sd / mean), geom = "tile") +
  gg(boundary, alpha = 0) +
  gg(nests)
```

Plot the vegetation "fixed effect" posteriors. First get their names - from
`$marginals.random$vegetation` of the fitted object, which contains the fixed
effect marginal distribution data

```{r warning=FALSE,message=FALSE}
flist <- vector("list", NROW(fit_veg_field$summary.random$vegetation))
for (i in seq_along(flist)) {
  flist[[i]] <- plot(fit_veg_field, "vegetation", index = i)
  flist_combined <- if (i == 1) flist[[i]] else flist_combined + flist[[i]]
}
flist_combined +
  plot_layout(ncol = 3, guides = "collect")
```

Use `spde.posterior( )` to obtain and then plot the SPDE parameter posteriors
and the Matern correlation and covariance functions for this model.

```{r results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
spde.range <- spde.posterior(fit_veg_field, "field", what = "range")
spde.logvar <- spde.posterior(fit_veg_field, "field", what = "log.variance")
range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)
(range.plot / var.plot)

corplot <-
  plot(spde.posterior(fit_veg_field, "field", what = "matern.correlation"))
covplot <-
  plot(spde.posterior(fit_veg_field, "field", what = "matern.covariance"))
(covplot / corplot)
```

Continuous covariates
-----------------------------------

Now lets try a model with elevation as a (continuous) explanatory variable.
(First centre elevations for more stable fitting.)

```{r results="hide",warning=FALSE,message=FALSE}
elev <- gcov$elevation
elev <- elev - mean(terra::values(elev), na.rm = TRUE)

ggplot() +
  gg(elev, geom = "tile") +
  gg(boundary, alpha = 0)
```


The elevation variable here is of class 'SpatRaster', that can be handled in the
same way as the vegetation covariate, with automatic evaluation via an
`eval_spatial()` method. However, since in some cases data may be stored
differently, other methods are needed to access the stored values, or there's
some post-processing to be done. In such cases, we can define a function that
knows how to evaluate the covariate at arbitrary points in the survey region,
and call that function in the component definition. The method `eval_spatial()`
is the method that handles this automatically, and supports `terra` `SpatRaster`
and `sf` geometry points objects, and mismatching coordinate systems as well.
In the following evaluator example function, we only add infilling of missing
values as a post-processing step.

```{r results="hide",warning=FALSE,message=FALSE}
# Note: this method is usually not needed; the automatic invocation of
# `eval_spatial()` method by the component input evaluator is usually
# sufficient.
f.elev <- function(where) {
  # Extract the values
  v <- eval_spatial(elev, where, layer = "elevation")
  # Fill in missing values; this example would work for
  # SpatialPixelsDataFrame data
  # if (any(is.na(v))) {
  #   v <- bru_fill_missing(elev, where, v)
  # }
  v
}
```

For brevity we are not going to consider models with elevation only, with
elevation and a SPDE, and with SPDE only. We will just fit one with elevation
and SPDE.
We create our model to pass to lgcp thus:


```{r results="hide",warning=FALSE,message=FALSE}
matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(1, 0.01),
  prior.range = c(0.1, 0.01)
)

comp_elev_field <- geometry ~ elev(f.elev(.data.), model = "linear") +
  field(geometry, model = matern) + Intercept(1)
```

Note how the elevation effect is defined. We could alternatively use the `terra`
grid object directly (causing `inlabru` to automatically call `eval_spatial()`),
like in the vegetation case:
we specified it like
```{r results="hide",warning=FALSE,message=FALSE,eval=FALSE}
elev(elev, model = "factor_full")
```
whereas with the special function method we specify the covariate like this:
```{r results="hide",warning=FALSE,message=FALSE,eval=FALSE}
elev(f.elev(.data.), model = "linear")
```
Most applications can use the automatic method, and the special function method
is included only as an example of how to handle more complex cases.

We also now include an intercept term in the model.

The model is fitted in the usual way:
```{r results="hide",warning=FALSE,message=FALSE,eval=FALSE}
fit_elev_field <- lgcp(comp_elev_field, nests,
  samplers = boundary,
  domain = list(geometry = mesh_int)
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_elev_field.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_elev_field <- readRDS(file = .cache_file)
} else {
  fit_elev_field <- lgcp(comp_elev_field,
    nests,
    samplers = boundary,
    domain = list(geometry = mesh_int)
  )
  saveRDS(fit_elev_field, .cache_file)
}
```

Summary and model selection


```{r warning=FALSE,message=FALSE}
summary(fit_elev_field)
deltaIC(fit_veg, fit_veg_field, fit_field, fit_elev_field)
```

Predict and plot the density


```{r results="hide",warning=FALSE,message=FALSE}
pred_elev_field <- predict(
  fit_elev_field,
  pred.df,
  ~ list(
    int = exp(field + elev + Intercept),
    int.log = field + elev + Intercept
  )
)

p1 <- ggplot() +
  gg(pred_elev_field$int, aes(fill = log(sd)), geom = "tile") +
  gg(boundary, alpha = 0) +
  gg(nests, shape = "+")
p2 <- ggplot() +
  gg(pred_elev_field$int.log, aes(fill = exp(mean + sd^2 / 2)), geom = "tile") +
  gg(boundary, alpha = 0) +
  gg(nests, shape = "+")
(p1 | p2)
```

Now look at the elevation and SPDE effects in space. Leave out the Intercept
because it swamps the spatial effects of elevation and the SPDE in the 
plots and we are interested in comparing the effects of elevation and the SPDE.

First we need to predict on the linear predictor scale.
```{r warning=FALSE,message=FALSE,echo=TRUE}
lp_elev_field <- predict(
  fit_elev_field,
  pred.df,
  ~ list(
    smooth_elev = field + elev,
    elev = elev,
    smooth = field
  )
)
```

The code below, which is very similar to that used for the vegetation factor
variable, produces the plots we want.


```{r warning=FALSE,message=FALSE, fig.height=5, fig.width=13}
lprange <- range(
  lp_elev_field$smooth_elev$mean,
  lp_elev_field$elev$mean,
  lp_elev_field$smooth$mean
)

library(RColorBrewer)
csc <- scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), limits = lprange)

plot.lp_elev_field <- ggplot() +
  gg(lp_elev_field$smooth_elev, mask = boundary, geom = "tile") +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("SPDE + elevation")

plot.lp_elev_field.spde <- ggplot() +
  gg(lp_elev_field$smooth, mask = boundary, geom = "tile") +
  csc +
  gg(boundary, alpha = 0) +
  ggtitle("SPDE")

plot.lp_elev_field.elev <- ggplot() +
  gg(lp_elev_field$elev, mask = boundary, geom = "tile") +
  csc +
  gg(boundary, alpha = 0) +
  ggtitle("elevation")

(plot.lp_elev_field + plot.lp_elev_field.spde + plot.lp_elev_field.elev) +
  plot_layout(ncol = 3, guides = "collect") &
  guides(fill = guide_legend("value")) &
  theme(legend.position = "right")
```

You might also want to look at the posteriors of the fixed effects and of the
SPDE. Adapt the code used for the vegetation factor to do this.

```{r warning=FALSE,message=FALSE}
Lambda_elev_field <- predict(
  fit_elev_field,
  fm_int(mesh_int, boundary),
  ~ sum(weight * exp(Intercept + elev + field))
)
Lambda_elev_field
```


```{r results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
flist <- vector("list", NROW(fit_elev_field$summary.fixed))
for (i in seq_along(flist)) {
  flist[[i]] <- plot(fit_elev_field, rownames(fit_elev_field$summary.fixed)[i])
  flist_combined <- if (i == 1) flist[[i]] else flist_combined + flist[[i]]
}
flist_combined +
  plot_layout(ncol = 2, guides = "collect")
```

Plot the SPDE parameter posteriors and the Matern correlation and covariance
functions for this model.

```{r results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
spde.range <- spde.posterior(fit_elev_field, "field", what = "range")
spde.logvar <- spde.posterior(fit_elev_field, "field", what = "log.variance")
range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)
(range.plot / var.plot)

corplot <-
  plot(spde.posterior(fit_elev_field, "field", what = "matern.correlation"))
covplot <-
  plot(spde.posterior(fit_elev_field, "field", what = "matern.covariance"))
(covplot / corplot)
```

Also estimate abundance. The `data.frame` in the second call leads to inclusion
of `N` in the prediction object, for easier plotting.

```{r fig.show="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
Lambda_elev_field <- predict(
  fit_elev_field, fm_int(mesh_int, boundary),
  ~ sum(weight * exp(field + elev + Intercept))
)
Lambda_elev_field

Nest_elev_field <- predict(
  fit_elev_field,
  fm_int(mesh_int, boundary),
  ~ data.frame(
    N = 200:1000,
    density = dpois(200:1000,
      lambda = sum(weight * exp(field + elev + Intercept))
    )
  ),
  n.samples = 2000
)
```

Plot in the same way as in previous practicals

```{r }
Nest_elev_field$plugin_estimate <-
  dpois(Nest_elev_field$N, lambda = Lambda$median)
ggplot(data = Nest_elev_field) +
  geom_line(aes(x = N, y = mean, colour = "Posterior")) +
  geom_line(aes(x = N, y = plugin_estimate, colour = "Plugin"))
```

### Non-spatial evaluation of the covariate effect

The previous examples of posterior prediction focused on spatial prediction.
It is also possible to evaluate the effect of a covariate directly, by bypassing
the component definition input. This is done by adding the suffix `_eval()` to
the end of the component name in the predictor expression, and supplying data
that is sufficient for evaluating , and supplying data needed by the input
arguments to this function, see `bru_comp_eval()`.

Note on backwards version support, starting with version `2.2.8`
Prior to version `2.8.0`, this feature required setting the `include`
argument to `character(0)` to disable normal component evaluation for all
components.
From version `2.8.0`, `inlabru` automatically detects which model components
are used by the prediction expression, including the use of component `_eval()`
calls, making the `include` argument unnecessary. From version `2.11.0`,
the `include`, `exclude`, and `include_latent` arguments have been deprecated in
favour of the `used` argument that takes input generated by `bru_used()`, with
a deprecation warning being generated from version `2.12.0.9003`.

Since the elevation effect in this model is linear, the resulting plot isn't
very interesting, but the same method can be applied to non-linear effects
(e.g. "rw2") as well, and combined into general R expressions.

```{r}
elev.pred <- predict(
  fit_elev_field,
  data.frame(elevation = seq(0, 100, length.out = 1000)),
  formula = ~ elev_eval(elevation)
  # include = character(0) # Not needed from version 2.8.0
)

ggplot(elev.pred) +
  geom_line(aes(elevation, mean)) +
  geom_ribbon(
    aes(elevation,
      ymin = q0.025,
      ymax = q0.975
    ),
    alpha = 0.2
  ) +
  geom_ribbon(
    aes(elevation,
      ymin = mean - 1 * sd,
      ymax = mean + 1 * sd
    ),
    alpha = 0.2
  )
```


A 1D Example
----------------
Try fitting a 1-dimensional model to the point data in the `inlabru` dataset
`Poisson2_1D`. This comes with a covariate function called `cov2_1D`. Try to
reproduce the plot below (used in lectures) showing the effects of the
`Intercept + z` and the `SPDE`. (You may find it helpful to build on the model
you fitted in the previous practical, adding the covariate to the model
specification.)

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
data(Poisson2_1D)
x <- seq(0, 55, length.out = 50)
mesh <- fm_mesh_1d(x, degree = 2, boundary = "free")

process_model <- inla.spde2.pcmatern(
  mesh,
  prior.sigma = c(1, 0.01),
  prior.range = c(5, 0.01),
  constr = TRUE
)
comp <- x ~
  z(cov2_1D(x), model = "linear") +
  smooth(x, model = process_model) +
  Intercept(1)

fitcov1D <- lgcp(comp,
  pts2,
  domain = list(x = mesh),
  samplers = tibble::tibble(x = cbind(0, 55))
)
pr.df <- data.frame(x = x)
prcov1D <- predict(
  fitcov1D,
  pr.df,
  ~ list(
    Intensity = exp(z + smooth + Intercept),
    z = exp(z + Intercept),
    smooth = exp(smooth)
  ),
  n.samples = 2000
)

ggplot() +
  gg(prcov1D$Intensity,
    aes(colour = "Intensity", fill = "Intensity"),
    alpha = 0.2,
    lwd = 1.25
  ) +
  gg(prcov1D$smooth,
    aes(colour = "Smooth", fill = "Smooth"),
    alpha = 0.2,
    lwd = 1.25
  ) +
  gg(prcov1D$z,
    aes(colour = "z-effect", fill = "z-effect"),
    alpha = 0.2,
    lwd = 1.25
  ) +
  geom_line(aes(x, lambda2_1D(x), colour = "True intensity"), lwd = 1.25) +
  geom_point(data = pts2, aes(x = x), y = 0.2, shape = "|", cex = 4) +
  xlab(quote(bold(s))) +
  ylab(quote(hat(lambda)(bold(s)) ~ ~"and its components")) +
  guides(colour = guide_legend("Quantity"), fill = "none")
```
