---
title: "LGCPs - Distance sampling and per-transect covariates"
author: "Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Distance sampling and per-transect covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
.cache_dir <- file.path("2d_lgcp_per_transect_covars_cache")
.cache_recompute <- !file.exists(.cache_dir)
if (.cache_recompute) {
  dir.create(.cache_dir, recursive = TRUE)
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
library(INLA)
library(inlabru)
library(tibble)
library(sf)
library(fmesher)
library(dplyr)
library(ggplot2)
library(fmesher)
library(RColorBrewer)
```
Set things up

```{r results="hide",warning=FALSE,message=FALSE}
library(INLA)
library(inlabru)
library(tibble)
library(sf)
library(fmesher)
library(dplyr)
library(ggplot2)
library(fmesher)
library(RColorBrewer)
```

Construct sampler transects, with a per-transect covariate, `visibility`:
```{r}
samplers <- st_sf(
  bind_rows(
    tibble(
      transect = 1L,
      geometry = fm_as_sfc(fm_segm(rbind(c(0, 0), c(100, 0)),
        is.bnd = FALSE
      )),
      visibility = 1
    ),
    tibble(
      transect = 2L,
      geometry = fm_as_sfc(fm_segm(rbind(c(100, 0), c(100, 50)),
        is.bnd = FALSE
      )),
      visibility = 0.5
    )
  )
)
```
Point observations, with marks `distance` and `transect`:
```{r}
intensity <- 0.1
set.seed(1324L)
N_expected <- c(100 * 10 * intensity * 1, 50 * 10 * intensity * 0.5)
N <- c(rpois(1, N_expected[1]), rpois(1, N_expected[2]))
obs_coord <- rbind(
  tibble(
    x = runif(N[1], 0, 100),
    y = 0,
    transect = 1L,
    distance = abs(runif(N[1], -5, 5))
  ),
  tibble(
    x = 100,
    y = runif(N[2], 0, 50),
    transect = 2L,
    distance = abs(runif(N[2], -5, 5))
  )
)
observations <- st_as_sf(obs_coord, coords = c("x", "y"))
```

Domain of interest:
```{r}
bnd <- fm_as_sfc(fm_segm(
  rbind(c(-30, -40), c(125, -20), c(135, 70), c(-20, 40), c(-30, -40)),
  is.bnd = TRUE
))
```

Computational meshes:
```{r}
mesh <- fm_mesh_2d(
  loc = fm_hexagon_lattice(bnd, edge_len = 5),
  boundary = bnd,
  max.edge = c(10, 25)
)

# Distance mesh for integration scheme
distance_mesh <- fm_mesh_1d(
  seq(0, 5, length.out = 21),
  boundary = "free"
)
```

Transect integration scheme, including per-transect covariates:
```{r}
ips <- fm_int(
  domain = list(
    geometry = fm_subdivide(mesh, 1),
    distance = distance_mesh
  ),
  samplers = samplers
)
```

Join per-transect covariates into the point observations, so they are available
for model fitting:
```{r}
obs_with_transect_info <- observations |>
  left_join(as_tibble(samplers),
    by = "transect",
    suffix = c("", ".sampler")
  )


ggplot() +
  geom_fm(data = mesh) +
  geom_sf(data = ips, aes(size = weight, color = visibility), alpha = 0.2, stroke = 0) +
  scale_size_area(max_size = 3) +
  geom_sf(data = samplers, color = "blue") +
  geom_sf(
    data = obs_with_transect_info, aes(shape = as.factor(transect)),
    color = "red"
  ) +
  theme_minimal()
```


## Model fitting

Fitting model with visibility.
```{r results="hide"}
cmp <- ~ Intercept(1) +
  visibility(log(visibility), model = "linear") +
  twosided(log(2), model = "const")
obs <- bru_obs(
  formula = geometry + distance ~ Intercept + visibility + twosided,
  family = "cp",
  domain = list(
    geometry = fm_subdivide(mesh, 1),
    distance = distance_mesh
  ),
  samplers = samplers,
  data = obs_with_transect_info,
  options = list(control.inla = list(int.strategy = "eb"))
)
fit_vis <- bru(
  cmp,
  obs,
  options = list(
    bru_verbose = FALSE
  )
)
```
```{r}
summary(fit_vis)
```
Posterior predictive interval and median for the intensity
(true value `r intensity`)
```{r}
rbind(
  True = c(NA, intensity, NA),
  Estimated = exp(fit_vis$summary.fixed[1, c(
    "0.025quant",
    "0.5quant",
    "0.975quant"
  )])
)
```
Note: The true expected point counts for the two transects, taking the visibility into account are
`r N_expected[1]` and `r N_expected[2]`, respectively. The observed point counts
were `r N[1]` and `r N[2]`, respectively.
