---
title: "LGCPs - Plot sampling"
author: "David Borchers and Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Plot sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
```

Introduction
----------

This practical demonstrates use of the `samplers` argument in `lgcp`, which you
need to use when you have observed points from only a sample of plots in the survey 
region. 


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(mgcv)
library(ggplot2)
bru_safe_sp(force = TRUE)
```

Get the data
-----------------------------------

```{r }
data(gorillas_sf, package = "inlabru")
```

This dataset is a list (see `help(gorillas)` for details. Extract the the objects 
you need from the list, for convenience:

```{r }
nests <- gorillas_sf$nests
mesh <- gorillas_sf$mesh
boundary <- gorillas_sf$boundary
gcov <- gorillas_sf_gcov()
```

The `gorillas` data also contains a plot sample subset which covers 60% of the survey region.

```{r }
sample <- gorillas_sf$plotsample
```

```{r results="hide",warning=FALSE,message=FALSE}
plotdets <- ggplot() +
  gg(boundary) +
  gg(sample$plots) +
  gg(sample$nests, pch = "+", cex = 4, color = "red") +
  geom_text(aes(
    label = sample$counts$count,
    x = sf::st_coordinates(sample$counts)[, 1],
    y = sf::st_coordinates(sample$counts)[, 2]
  )) +
  labs(x = "Easting", y = "Northing")
plot(plotdets)
```

On this plot survey, only points within the rectangles are detected, but it is also 
informative to plot all the points here (which if it was a real plot survey you could
not do, because you would not have seen them all).


```{r results="hide",warning=FALSE,message=FALSE}
plotwithall <- ggplot() +
  gg(boundary) +
  gg(sample$plots) +
  gg(nests, pch = "+", cex = 4, color = "blue") +
  geom_text(aes(
    label = sample$counts$count,
    x = sf::st_coordinates(sample$counts)[, 1],
    y = sf::st_coordinates(sample$counts)[, 2]
  )) +
  gg(sample$nests, pch = "+", cex = 4, color = "red") +
  labs(x = "Easting", y = "Northing")
plot(plotwithall)
```

Inference
------------------------------------
The observed nest locations are in the SpatialPointsDataFrame `sample$nests`, and the 
plots are in the SpatialPolygonsDataFrame `sample$plots`. Again, we are using the following SPDE
setup:


```{r }
matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(0.05, 0.01)
)
```


Fit an LGCP model with SPDE only to these data by using the `samplers=` argument of
the function `lgcp( )`:


```{r results="hide",warning=FALSE,message=FALSE}
cmp <- geometry ~ my.spde(geometry, model = matern)

fit <- lgcp(cmp, sample$nests, samplers = sample$plots, domain = list(geometry = mesh))
```


Plot the density surface from your fitted model

```{r results="hide",warning=FALSE,message=FALSE}
pxl <- fm_pixels(mesh, mask = boundary)
lambda.sample <- predict(fit, pxl, ~ exp(my.spde + Intercept))
```

```{r warning=FALSE,message=FALSE}
lambda.sample.plot <- ggplot() +
  gg(lambda.sample, geom = "tile") +
  gg(sample$plots, alpha = 0) +
  gg(boundary, col = "yellow", alpha = 0)

lambda.sample.plot
```


Estimate the integrated intensity lambda. We compute both the overall integrated
intensity, representative of an imagined new realisation of the point process,
and the conditional expectation that takes the actually observed nests into account,
by recognising that we have complete information in the surveyed plots. 
```{r results="hide",warning=FALSE,message=FALSE}
Lambda <- predict(fit, fm_int(mesh, boundary), ~ sum(weight * exp(my.spde + Intercept)))
Lambda.empirical <- predict(
  fit,
  rbind(
    cbind(fm_int(mesh, boundary), data.frame(all = TRUE)),
    cbind(fm_int(mesh, sample$plots), data.frame(all = FALSE))
  ),
  ~ (sum(weight * exp(my.spde + Intercept) * all) -
    sum(weight * exp(my.spde + Intercept) * !all) +
    nrow(sample$nests))
)
rbind(
  Lambda,
  Lambda.empirical
)
```


Fit the same model to the full dataset (the points in `gorillas_sf$nests`), or get your previous 
fit, if you kept it. Plot the intensity surface and estimate the integrated intensity

```{r results="hide",warning=FALSE,message=FALSE}
fit.all <- lgcp(cmp, nests,
  samplers = boundary,
  domain = list(geometry = mesh)
)
lambda.all <- predict(fit.all, pxl, ~ exp(my.spde + Intercept))
Lambda.all <- predict(fit.all, fm_int(mesh, boundary), ~ sum(weight * exp(my.spde + Intercept)))
```

Your plot should look like this:


```{r results="hide",warning=FALSE,message=FALSE, echo=FALSE}
lambda.all.plot <- ggplot() +
  gg(lambda.all, geom = "tile") +
  gg(sample$plots, alpha = 0) +
  gg(boundary, col = "yellow", alpha = 0)

lambda.all.plot
```

The values `Lambda.empirical`, `Lambda`, and `Lambda.all` should be close
to each other if the plot samples
gave sufficient information for the overall prediction:

```{r }
rbind(
  Plots = Lambda,
  PlotsEmp = Lambda.empirical,
  All = Lambda.all,
  AllEmp = c(nrow(gorillas$nests), 0, rep(nrow(gorillas$nests), 3), rep(NA, 3))
)
```

Now, let's compare the results


```{r results="hide",warning=FALSE,message=FALSE,fig.width=7}
library(patchwork)
lambda.sample.plot + lambda.all.plot +
  plot_layout(guides = "collect") &
  theme(legend.position = "left") &
  scale_fill_continuous(limits = range(c(0, 340)))
```

Do you understand the reason for the differences in the posteriors of the abundance estimates?
