---
title: "LGCPs - Model comparison using Cross Validation"
author: "David Borchers, Finn Lindgren and Hans Montcho"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Model comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
.cache_dir <- file.path("2d_lgcp_covars_gcpo_cache")
.cache_recompute <- !file.exists(.cache_dir)
if (.cache_recompute) {
  dir.create(.cache_dir, recursive = TRUE)
}
.vignette_cache <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```
Set things up

```{r results="hide",warning=FALSE,message=FALSE}
library(INLA)
library(inlabru)
library(fmesher)
library(RColorBrewer)
library(ggplot2)
library(patchwork)
```

Introduction
----------------------------------

We are going to fit some spatial models to the gorilla data, and compare them using  cross validation.  For illustration, we consider the same models from the [vignette on spatial covariates](https://inlabru-org.github.io/inlabru/articles/2d_lgcp_covars.html), and we  only focus on the practical aspects of LGCPs comparison.  More specifically, we consider the following models:

| Models  | Components |
| -------- | -------- | 
| `Fit_hpp`    | Intercept ( a homogeneous PP) | 
| `Fit_veg`    | Intercept and factor covariate `vegetation`| 
| `Fit_field`    | Intercept and a spatial  `SPDE field`| 
| `Fit_veg_field`    | Intercept, `vegetation` and the  `SPDE field`| 
| `Fit_veg_elev_field`    | Intercept,  `vegetation`,  `elevation`, and the `SPDE field`| 


Leave-Region-Out Cross Validation
-----------------------------------

In point pattern analysis,  the available information comes from  *events locations* and *empty regions*,  and this should be taken into consideration for models comparison. Here, we use a  *leave-region-out cross validation* approach for model comparison,  in which the domain is partitioned into subregions, and the model is used the predict the likelihood contribution from  holdout subregions.

Let $Y$ be LGCP with intensity $\lambda(.)$ defined on a bounded domain $\Omega$, and we observe a point pattern $\mathbf{y} = (y_1,\cdots, y_n)$. Then the log likelihood is defined as 

$$
\ell(\lambda(.);\mathbf{y}, \Omega) = |\Omega| - \int_{\Omega}\lambda(s)ds + \sum_{y_i \in \mathbf{y}\cap\Omega } \ln \lambda(y_i). 
$$

Now, if we  consider a partition $\Omega = \Cup_{j=1}^J B_j$ of the domain; we can split the log likelihood into regional log likelihood contributions as  $\ell(\lambda(.);\mathbf{y}, \Omega) = \sum_{j=1}^J \ell_j(\lambda(.);\mathbf{y}, B_j)$  where

$$
\ell_j(\lambda(.);\mathbf{y}, B_j) = |B_j| - \int_{B_j}\lambda(s)ds + \sum_{y_i \in \mathbf{y}\cap B_j } \ln \lambda(y_i),
$$
and we note that, this representation *is not equivalent to the aggregated count or grid* approach. 

To fit the model, we use the complete information from $\ell(\lambda(.);\mathbf{y}, \Omega)$. For model comparison, however,  the  likelihood contribution, $\ell_j(\lambda(.);\mathbf{y}, B_j)$ from  each $B_j$, is left out and after fitting the model, we evaluate its predictive performance on the holdout subregion using the logarithmic scoring rule. 

The aggregated Group Conditional Predictive Ordinate (GCPO)  is defined as sum of  the logarithm of the joint leave-region-out predictive distribution 

$$\text{GCPO} = \sum_{j=1}^J \log \pi(\mathbf{y}_{B_j} | \mathbf{y}_{-B_j}) $$
where we define  

$$
\pi(\mathbf{y}_{B_j} | \mathbf{y}_{-B_j}) = \int_{B_j}\exp(\ell_j(\lambda(.);\mathbf{y}, B_j)) \pi(\lambda(s)|\mathbf{y}, \Omega\setminus B_j)) ds.
$$ 

In principle, the model should be refitted $J$ times, however, a fast and accurate 
approximation to the leave-region-out predictive distribution, accessible via the `control.gcpo` argument, allows us to compute the approximate GCPO score from a single model fit using the complete point pattern. 


Get the data
-----------------------------------

```{r }
data(gorillas_sf, package = "inlabru")
```

This dataset is a list (see `help(gorillas_sf)` for details. Extract the objects
you need from the list, for convenience:

```{r }
nests <- gorillas_sf$nests
mesh <- gorillas_sf$mesh
boundary <- gorillas_sf$boundary
gcov <- gorillas_sf_gcov()
ggplot() +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "red", cex = 0.5)
```



Covariates: vegetation and elevation
-----------------------------------

Let's take a look at the vegetation type, nests and boundary:

```{r warning=FALSE,message=FALSE}
ggplot() +
  gg(gcov$vegetation) +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "white", cex = 0.5)
```

Also, let's do the same for  elevation:

```{r results="hide",warning=FALSE,message=FALSE}
elev <- gcov$elevation
elev <- elev - mean(terra::values(elev), na.rm = TRUE)

ggplot() +
  gg(elev, geom = "tile") +
  gg(boundary, alpha = 0) +
  gg(nests, color = "white", cex = 0.5)
```

For the spatial component, we use a SPDE type smoother, with the mesh, boundary and nests illustrated below:

```{r warning=FALSE,message=FALSE}
ggplot() +
  gg(mesh) +
  gg(boundary, alpha = 0.2) +
  gg(nests, color = "red", cex = 0.5)
```

The models
----------------------------------

### Defining the models 
Below, we prepare the formulas for  the models under consideration.

```{r warning=FALSE,message=FALSE}
# fit_hpp
comp_hpp <- geometry ~ Intercept(1)

# fit_veg
comp_veg <- geometry ~ -1 +
  vegetation(gcov$vegetation, model = "factor_full")

# fit_field
pcmatern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(1, 0.01),
  prior.range = c(0.1, 0.01)
)
comp_field <- geometry ~ Intercept(1) +
  field(geometry, model = pcmatern)

# fit_veg_field
comp_veg_field <- geometry ~
  -1 +
  vegetation(gcov$vegetation, model = "factor_full") +
  field(geometry, model = pcmatern)

# fit_veg_elev_field
f.elev <- function(where) {
  v <- eval_spatial(elev, where, layer = "elevation")
  v
}
comp_veg_elev_field <- geometry ~
  -1 +
  vegetation(gcov$vegetation, model = "factor_full") +
  elev(f.elev(.data.), model = "linear") +
  field(geometry, model = pcmatern)
```

### A hexagonal partition for cross validation

To enable the block or regional  `group.cv` calculations, we need to split the domain into spatial
regions, and compute the association of each observed point to these subregions.
The `cv_hex` function allows to create hexagonal partitions, but if meaningful subregions of interest are available, these must be used to define the likelihood contribution.  

```{r warning=FALSE,message=FALSE}
cvpart <- cv_hex(boundary, cellsize = 0.15, n_group = 1)

ggplot() +
  gg(cvpart, aes(fill = group), alpha = 0.5) +
  gg(boundary, alpha = 0) +
  gg(nests, color = "white", cex = 0.5) +
  theme(legend.position = "none")
```


It is important check that *every event belongs to a region*, to ensure  the regional likelihood contribution is properly calculated:

```{r warning=FALSE,message=FALSE}
cvpart$block_ID <- seq_len(nrow(cvpart))
cvpart$group <- NULL
a <- sf::st_intersects(nests, cvpart)
if (any(lengths(a) != 1)) {
  stop("Point in none or multiple polygons")
}
nests$.block <- unlist(a)
```


### Fitting the models

Fit the models as usual, and increase the integration scheme resolution to
better match the covariate. In the code below,`enable = TRUE` tells `INLA` that the `Cross validation score` must be calculated,  `type.cv = "joint"` ensures that we are doing a joint prediction for the entire region. For speed, we set `control.compute(config=FALSE)` to compute the `group.cv` *score conditional on posterior mode of the hyperparameters*. 

```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
mesh_int <- fm_subdivide(mesh, 2)
mesh_int <- mesh

gcpo_config <- list(enable = TRUE, type.cv = "joint")
options_config <- list(
  bru_verbose = 0,
  control.compute(config = FALSE)
)
```


```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit_hpp <- lgcp(
  comp_hpp,
  nests,
  samplers = cvpart,
  domain = list(geometry = mesh_int),
  control.gcpo = gcpo_config,
  options = options_config
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_hpp.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_hpp <- readRDS(file = .cache_file)
} else {
  fit_hpp <- lgcp(
    comp_hpp,
    nests,
    samplers = cvpart,
    domain = list(geometry = mesh_int),
    control.gcpo = gcpo_config,
    options = options_config
  )

  saveRDS(fit_hpp, .cache_file)
}
```


```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit_veg <- lgcp(
  comp_veg,
  nests,
  samplers = cvpart,
  domain = list(geometry = mesh_int),
  control.gcpo = gcpo_config,
  options = options_config
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_veg.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_veg <- readRDS(file = .cache_file)
} else {
  fit_veg <- lgcp(
    comp_veg,
    nests,
    samplers = cvpart,
    domain = list(geometry = mesh_int),
    control.gcpo = gcpo_config,
    options = options_config
  )

  saveRDS(fit_veg, .cache_file)
}
```


```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit_field <- lgcp(
  comp_field,
  nests,
  samplers = cvpart,
  domain = list(geometry = mesh_int),
  control.gcpo = gcpo_config,
  options = options_config
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_field.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_field <- readRDS(file = .cache_file)
} else {
  fit_field <- lgcp(
    comp_field,
    nests,
    samplers = cvpart,
    domain = list(geometry = mesh_int),
    control.gcpo = gcpo_config,
    options = options_config
  )
  saveRDS(fit_field, .cache_file)
}
```


```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit_veg_field <- lgcp(
  comp_veg_field,
  nests,
  samplers = cvpart,
  domain = list(geometry = mesh_int),
  control.gcpo = gcpo_config,
  options = options_config
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_veg_field.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_veg_field <- readRDS(file = .cache_file)
} else {
  fit_veg_field <- lgcp(
    comp_veg_field,
    nests,
    samplers = cvpart,
    domain = list(geometry = mesh_int),
    control.gcpo = gcpo_config,
    options = options_config
  )
  saveRDS(fit_veg_field, .cache_file)
}
```



```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit_veg_elev_field <- lgcp(
  comp_veg_elev_field,
  nests,
  samplers = cvpart,
  domain = list(geometry = mesh_int),
  control.gcpo = gcpo_config,
  options = options_config
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit_veg_elev_field.rds")
if ((!.cache_recompute) &&
  file.exists(.cache_file)) {
  fit_veg_elev_field <- readRDS(file = .cache_file)
} else {
  fit_veg_elev_field <- lgcp(
    comp_veg_elev_field,
    nests,
    samplers = cvpart,
    domain = list(geometry = mesh_int),
    control.gcpo = gcpo_config,
    options = options_config
  )
  saveRDS(fit_veg_elev_field, .cache_file)
}
```

### Comparing the models with group CV

There are mainly two functions `bru_block_gcpo` and `bru_gcpo_table` for models comparison using cross validation. `bru_block_gcpo` is used to  access  individual scores for a *single* model, and its output is a list containing the index for each region and the corresponding CV score.  On the other hand, `bru_gcpo_table` is used for *multiple* models comparison.


```{r}
gcpo_hpp <- bru_block_gcpo(fit_hpp)
str(gcpo_hpp)
```


From the table below, the  `vegetation` covariate is a strong predictor, as expected, but also the `SPDE smoother`.  Including `elevation` in the model already containing `vegetation` and the `SPDE smoother` does not provide any considerable gain in terms of predictive score.  These conclusions are similar to the ones obtained from the [vignette on spatial covariates](https://inlabru-org.github.io/inlabru/articles/2d_lgcp_covars.html), although the latter was mainly based on in-sample analysis.
```{r}
gcpo_df <- bru_gcpo_table(
  list(
    Fit_hpp = fit_hpp,
    Fit_veg = fit_veg,
    Fit_field = fit_field,
    Fit_veg_field = fit_veg_field,
    Fit_veg_elev_field = fit_veg_elev_field
  )
)

gcpo_df[, -1] <- log(gcpo_df[, -1])
knitr::kable(tibble(
  Models = names(gcpo_df)[-1],
  GCPO = colSums(gcpo_df[, -1])
))
```

### Plotting  regional scores

A valuable feature of the regional CV approach, is to assess regions where models might provide better fits or to detect some missing covariate. As an illustration, by evaluating the difference in  the log predictive density per region between models `Fit_veg` and `Fit_hpp`, we can see that including `vegetation` improves the prediction in regions with high number of nests,  which was expected, since  most of the nests falls in the primary vegetation type.


```{r}
names(gcpo_df)[1] <- "block_ID"
gcpo_block_ID <- cvpart |> dplyr::left_join(gcpo_df, "block_ID")
ggplot() +
  gg(gcpo_block_ID, aes(fill = Fit_veg - Fit_hpp), alpha = 1) +
  gg(boundary, alpha = 0) +
  gg(nests, color = "white", cex = 0.2)
```


A similar analysis could be done to assess the gain from the `SPDE smoother` when added to the model with `vegetation`. Indeed, most of the subregions with increased GCPO are nearby regions, probably as a result of the  spatial component. 

```{r}
ggplot() +
  gg(gcpo_block_ID, aes(fill = Fit_veg_field - Fit_veg), alpha = 1) +
  gg(boundary, alpha = 0) +
  gg(nests, color = "white", cex = 0.2)
```


We can also look at the contribution from the `elevation` covariate, which slightly increased the log score in few subregions,  but overall, it seems to not be relevant. This was also expected from its posterior distribution.   

```{r}
plt_gcpo_elev <- ggplot() +
  gg(gcpo_block_ID, aes(fill = Fit_veg_elev_field - Fit_veg_field),
    alpha = 1
  ) +
  gg(boundary, alpha = 0) +
  gg(nests, color = "white", cex = 0.2)

plt_post_elev <- plot(fit_veg_elev_field, "elev")

plt_gcpo_elev + plt_post_elev
```


Therefore, based on the previous analysis,  we would select the model  containing `vegetation` and the `SPDE smoother`. We note the small difference in the GCPOs between some of the models, making model selection less straightforward. This illustrates the need for uncertainty quantification, that we are currently investigating. 
