---
title: "Random Fields in One Dimension"
author: "Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random Fields in One Dimension}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
library(inlabru)
inla_available <- inlabru::bru_safe_inla(
  multicore = interactive(),
  quietly = TRUE
)
```


Setting things up
----------------

```{r results="hide",warning=FALSE,message=FALSE,eval=inla_available}
library(INLA)
library(inlabru)
library(mgcv)
library(ggplot2)
library(fmesher)
library(patchwork)
```

Make a shortcut to a nicer colour scale:
```{r results="hide",warning=FALSE,message=FALSE,eval=inla_available}
colsc <- function(...) {
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlBu")),
    limits = range(..., na.rm = TRUE)
  )
}
```


Get the data
-----------------------------------

Load the data and rename the countdata object to `cd` (just because '`cd`'
is less to type than '`countdata2`'.):

```{r results="hide",warning=FALSE,message=FALSE,eval=inla_available}
data(Poisson2_1D)
cd <- countdata2
```

Take a look at the count data.

```{r warning=FALSE,message=FALSE,eval=inla_available}
cd
ggplot(cd) +
  geom_point(aes(x, y = count)) +
  ylim(0, max(cd$count))
```

_Tip_: `RStudio > Help > Cheatsheets > Data visualisation with ggplot2` is a useful 
reference for `ggplot2` syntax.

Fitting a Generalised Additive Model (GAM)
-----------------------------------
If you're not familiar with GAMs and the syntax of `gam` don't worry, the point of 
this is just to provide something to which we can compare the `inlabru` model fit.

```{r results="hide",warning=FALSE,message=FALSE,eval=inla_available}
fit2.gam <- gam(count ~ s(x, k = 10) + offset(log(exposure)),
  family = poisson(),
  data = cd
)
```

The term `s(x,k=10)` just specifies that as nonparametric smooth function is to
be fitted to the data, with _no more than_ 10 degrees of freedom (df). (The
larger the df, the more wiggly the fitted curve (recall from the lecture that
this is an effect of how some spline methods are defined, without discretisation
dependent penalty); `gam` selects the 'best' df.)  Notice the use of `offset=`
to incorporate the variable `exposure` in `cd` is the size of the bin in which
each count was made.

You can look at the fitted model using `summary( )` as below if you want to, but
you do not need to understand this output, or the code that makes the
predictions immediately below it if you are not familiar with GAMs.

```{r results="hide",warning=FALSE,message=FALSE,eval=inla_available}
summary(fit2.gam)
```

Make a prediction data frame, get predictions and add them to the data frame
First make vectors of x-values and associated (equal) exposures:

```{r,eval=inla_available}
xs <- seq(0, 55, length = 100)
exposures <- rep(cd$exposure[1], 100)
```

and put them in a data frame:

```{r,eval=inla_available}
dat4pred <- data.frame(x = xs, exposure = exposures)
```

Then predict 

```{r,eval=inla_available}
pred2.gam <- predict(fit2.gam, newdata = dat4pred, type = "response")
# add column for prediction in data frame:
dat4pred2 <- cbind(dat4pred, gam = pred2.gam)
```

Plotting the fit and the data using the `ggplot2` commands below should give you
the plot shown below

```{r,eval=inla_available}
ggplot(dat4pred2) +
  geom_line(aes(x = x, y = gam), lty = 2) +
  ylim(0, max(dat4pred2$gam, cd$count)) +
  geom_point(aes(x = x, y = count), cd)
```

Fitting an SPDE model with inlabru
-----------------------------------
Make mesh. To avoid boundary effects in the region of interest, let the
mesh extend outside the data range.

```{r,eval=inla_available}
x <- seq(-10, 65, by = 5) # this sets mesh points - try others if you like
(mesh1D <- fm_mesh_1d(x, degree = 2, boundary = "free"))
```

... and see where the mesh knots are:


```{r fig.show="hide",eval=inla_available}
ggplot() +
  gg(mesh1D)
```

We can also draw the basis functions:

```{r,eval=inla_available}
ggplot() +
  geom_fm(data = mesh1D)
```

For special cases, one can control the knots and boundary conditions. Normally,
we just need to ensure that there is enough flexibility to represent the intended Gaussian
process dependence structure, while not adding needless computations.

```{r,eval=inla_available}
(mesh <- fm_mesh_1d(c(1, 2, 3, 4, 6),
  boundary = c("neumann", "free"),
  degree = 2
))
ggplot() +
  geom_fm(data = mesh, xlim = c(0, 7))
```

### Using function `bru( )` to fit to count data

We need to specify model components and a model formula in order to fit it.
This can be done inside the call to `bru( )` but that is a bit messy, so we'll
store it in `comp` first and then pass that to `bru( )`.

Our response variable in the data frame `cd` is called `count` so the model
specification needs to have that on the left of the `~`. We add an intercept
component with `+ Intercept(1)` on the right hand side (all the models we use
have intercepts), and because we want to fit a Gaussian random field (GRF), it
must have a GRF specification. In `inlabru` the GRF specification is a function,
which allows the GRF to be calculated at any point in space while `inlabru` is
doing its calculations.

The user gets to name the GRF function. The syntax is
`myname(input, model= ...)`, where:

* 'myname' is whatever you want to call the GRF (we called it `field` below);
* `input` specifies the coordinates in which the GRF or SPDE 'lives'. Here we
  are working in one dimension, and we called that dimension `x` when we set
  up the data set.
* `model=` designates the type of effect, here an SPDE model object from the
  `INLA` function `inla.spde2.pcmatern( )`, which requires a mesh to be passed
  to it, so we pass it the 1D mesh that we created above, `mesh1D`.

For models that only sums all the model components, we don't need to specify the
full predictor formula. Instead, we can provide the name of the output to the
left of the `~` in the component specification, and "." on the right hand side,
which will cause it to add all components (unless a subset is selected via the
`used` argument to `bru_obs()`).

```{r,eval=inla_available}
the_spde <- inla.spde2.pcmatern(mesh1D,
  prior.range = c(1, 0.01),
  prior.sigma = c(1, 0.01)
)

comp <- ~ field(x, model = the_spde) + Intercept(1, prec.linear = 1 / 2^2)

fit2.bru <- bru(
  comp,
  bru_obs(
    count ~ .,
    data = cd,
    family = "poisson",
    E = exposure
  )
)

summary(fit2.bru)
```

Predict the values at the x points used for mesh
(the data argument must be a data frame, see `?predict.bru`):

```{r,eval=inla_available}
x4pred <- data.frame(x = xs)
pred2.bru <- predict(fit2.bru,
  x4pred,
  x ~ exp(field + Intercept),
  n.samples = 1000
)
```

Let's do a plot to compare the fitted model to the true model. The expected
counts of the true model are stored in the variable `E_nc2` which comes with the
dataset `Poisson2_1D`. For ease of use in plotting with `ggplot2` (which needs a
data frame), we create a data frame which we call `true.lambda`, containing `x`-
and `y` variables as shown below.

Given that `inlabru` predictions are always on the intensity function scale, do
you understand why we divide the count by `cd$exposure`? (We will in due course
allow predictions on the count scale as well.)


```{r,eval=inla_available}
true.lambda <- data.frame(x = cd$x, y = E_nc2 / cd$exposure)
```

These `ggplot2` commands should generate the plot shown below. It shows the true
intensities as short horizontal blue lines, the observed intensities as black
dots, and the fitted intensity function as a red curve, with 95% credible
intervals shown as a light red band about the curve.

```{r,eval=inla_available}
ggplot() +
  gg(pred2.bru) +
  geom_point(data = cd, aes(x = x, y = count / exposure), cex = 2) +
  geom_point(data = true.lambda, aes(x, y), pch = "_", cex = 9, col = "blue") +
  coord_cartesian(xlim = c(0, 55), ylim = c(0, 6)) +
  xlab("x") +
  ylab("Intensity")
```

Compare the `inlabru` fit to the `gam` fit:

```{r fig.show="hide",eval=inla_available}
ggplot() +
  gg(pred2.bru) +
  geom_point(data = cd, aes(x = x, y = count / exposure), cex = 2) +
  geom_line(data = dat4pred2, aes(x, gam / exposure), lty = 2) +
  coord_cartesian(xlim = c(0, 55), ylim = c(0, 6)) +
  xlab("x") +
  ylab("Intensity")
```

### Looking at the posterior distributions
We can look at the Intercept posterior using the function 
`plot( )`, as below. 

```{r,eval=inla_available}
plot(fit2.bru, "Intercept")
```

You have to know that there is a variable called `Intercept` in order to use
this function. To see what fixed effect parameters' posterior distributions are
available to be plotted, you can type

```{r results="hide",warning=FALSE,message=FALSE,eval=inla_available}
names(fit2.bru$marginals.fixed)
```

This does not tell you about the SPDE parameters, and if you type

```{r warning=FALSE,message=FALSE,eval=inla_available}
names(fit2.bru$marginals.random)
```

this just tells you that there is an SPDE in fit2.bru called 'field', it does
not tell you what the associated parameter names are. The parameters that are
used in estimation are cryptic -- what we are interested in is the range and
variance of the Matern covariance funcion, that are functions of the internal
parameters.  We can look at the posterior distributions of the range parameter
and the log of the variance parameters as follows. (We look at the posterior of
the *log* of the variance because the variance posterior is very skewed and so
it is easier to view the log of the variance)


```{r warning=FALSE,message=FALSE,eval=inla_available}
spde.range <- spde.posterior(fit2.bru, "field", what = "range")
spde.logvar <- spde.posterior(fit2.bru, "field", what = "log.variance")

range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)

(range.plot | var.plot)
```

We can look at the posterior distributions of the Matern correlation and
covariance functions as follows:

```{r warning=FALSE,message=FALSE,eval=inla_available}
(plot(spde.posterior(fit2.bru, "field", what = "matern.correlation")) /
  plot(spde.posterior(fit2.bru, "field", what = "matern.covariance")))
```
