---
title: "Tidy model output with broom-style tidiers"
author: "inlabru authors"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tidy model output with broom-style tidiers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
library(inlabru)
inla_available <- inlabru::bru_safe_inla(
  multicore = interactive(),
  quietly = TRUE
)
```

inlabru provides [broom](https://broom.tidymodels.org/)-style tidiers for `bru` model
objects through three generics from the
[generics](https://cran.r-project.org/package=generics) package:

- `tidy()` — one row per model term (fixed effects or hyperparameters)
- `glance()` — one row of model-level fit summaries
- `augment()` — original data extended with posterior fitted values

These make it straightforward to work with inlabru results inside tidy
workflows (dplyr, ggplot2, etc.) without manually unpacking the `bru` object.

Note that the column names in the broom framework are standardised in a way that
in most cases are inaccurate in the Bayesian context (e.g. `std.error` in the
frequentist context is the standard error of an estimator, but to allow these
methods to work in a Bayesian context, the value used is the posterior standard
deviation of the estimated quantity). However, the tidy output is still useful
for quick summaries and plotting, and the column names are chosen to integrate
with tidy tools that expect them.  Only fixed effects and hyperparameters are
initially supported `tidy()`, not random effects; those may be added in a
future version.

Introduced in version `2.14.1.9005`.

## Set up

```{r libraries, eval=inla_available, message=FALSE, warning=FALSE}
library(INLA)
library(inlabru)
library(ggplot2)
bru_options_set(control.compute = list(dic = TRUE, waic = TRUE))
```

We use a simple simulated Gaussian dataset throughout.

```{r simulate, eval=inla_available}
set.seed(42L)
n <- 100
x <- runif(n, 0, 5)
y <- 2 + 1.5 * x + rnorm(n, sd = 0.8)
dat <- data.frame(x = x, y = y)
```

## Fit a model

```{r fit, eval=inla_available, results="hide"}
fit <- bru(
  y ~ Intercept(1) + slope(x, model = "linear"),
  family = "gaussian",
  data = dat
)
```

## `tidy()` — fixed effects

`tidy()` returns a tibble with one row per fixed-effect term, including the
posterior mean, standard deviation, and 95 % credible interval.

```{r tidy-fixed, eval=inla_available}
library(generics)
tidy(fit)
```

The column names follow the broom convention so the output integrates with
standard tidy tools.  For example, a quick coefficient plot:

```{r coef-plot, eval=inla_available}
tidy(fit) |>
  ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(title = "Fixed-effect posterior summaries", x = NULL, y = "Estimate")
```

## `tidy()` — hyperparameters

Pass `effects = "hyperpar"` to extract the precision (or other hyperparameters)
instead.

```{r tidy-hyperpar, eval=inla_available}
tidy(fit, effects = "hyperpar")
```

The Gaussian family has one hyperparameter — the precision of the likelihood.
More complex models (e.g. with SPDE random fields) will have additional rows.

## `glance()` — model-level summaries

`glance()` returns a single-row tibble with the DIC, WAIC, marginal log-likelihood,
the number of observations, and total wall-clock time.

```{r glance, eval=inla_available}
glance(fit)
```

This is useful when comparing competing models:

```{r compare, eval=inla_available, results="hide"}
fit_null <- bru(
  y ~ Intercept(1),
  family = "gaussian",
  data = dat
)
```

```{r compare-show, eval=inla_available}
library(dplyr)
bind_rows(
  glance(fit_null) |> mutate(model = "intercept only"),
  glance(fit)      |> mutate(model = "intercept + slope")
) |>
  select(model, dic, waic, marginal_loglik, nobs)
```

The model with `slope` should have a substantially lower DIC and WAIC.

## `augment()` — fitted values on new data

`augment()` adds posterior summaries of the linear predictor to a data frame.
Because inlabru prediction expressions are arbitrary R code, you must supply a
`pred_formula` that names what to predict.

```{r augment, eval=inla_available, results="hide"}
grid <- data.frame(x = seq(0, 5, length.out = 50))

augmented <- augment(
  fit,
  data = grid,
  pred_formula = ~ Intercept + slope,
  n_samples = 500L,
  seed = 1L
)
```

```{r augment-show, eval=inla_available}
head(augmented)
```

The four added columns are `.fitted` (posterior mean), `.fitted_low` and
`.fitted_high` (2.5 % and 97.5 % quantiles), and `.fitted_sd` (posterior
standard deviation).

Plot the fitted line with a credible band on top of the raw data:

```{r augment-plot, eval=inla_available}
ggplot() +
  geom_point(data = dat, aes(x, y), alpha = 0.4, size = 0.8) +
  geom_ribbon(
    data = augmented,
    aes(x, ymin = .fitted_low, ymax = .fitted_high),
    fill = "steelblue", alpha = 0.3
  ) +
  geom_line(data = augmented,
            aes(x, .fitted),
            colour = "steelblue", linewidth = 1) +
  labs(
    title = "Posterior fitted line with 95 % credible band",
    x = "x", y = "y"
  )
```

## Naming convention for `pred_formula`

The latent variables in `pred_formula` follow inlabru's naming rule: the effect
of a component named `foo` is available as `foo`, and the corresponding
latent variables (model coefficients, spline coefficients, etc) can be
accessed as `foo_latent` inside prediction
expressions.  In classic frequentist terminology, these two distinct quantities
are usually conflated, making it difficult to know whether one is talking about
`beta_x` (the latent variable/coefficient), `x` (the covariate value), or
`beta_x * x` (the effect of a model component that multiplies a latent variable
with a covariate). In `inlabru` expressions, the covariate can be accessed as
`.data$x`, the effect can be accessed as `x` or `.effect$x`, and the latent
variable can be accessed as `x_latent` or `.latent$x`.

Components with `model = "linear"` (including the implicit
intercept) appear in `summary.fixed`; components with a random-effect model
(iid, rw1, SPDE, …) appear in `summary.random`.

```{r names, eval=inla_available}
# fixed-effect component names → use <name>_latent in pred_formula to access
# the coefficients, and <name> to access the effect itself.
rownames(fit$summary.fixed)

# random-effect component names (none in this model)
names(fit$summary.random)
```

## Calling tidiers with the `broom` package

The tidiers are registered as S3 methods on `bru` objects through the
[generics](https://cran.r-project.org/package=generics) package, so they work
automatically when you load either `generics` or `broom`.

```{r broom, eval=inla_available && requireNamespace("broom", quietly = TRUE)}
library(broom)
tidy(fit)
glance(fit)
```
