---
title: "Tidy model output with broom-style tidiers"
author: "inlabru authors"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tidy model output with broom-style tidiers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
library(inlabru)
inla_available <- inlabru::bru_safe_inla(
  multicore = interactive(),
  quietly = TRUE
)
```

inlabru provides [broom](https://broom.tidymodels.org/)-style tidiers for `bru` model
objects through three generics from the
[generics](https://cran.r-project.org/package=generics) package:

- `tidy()` — one row per model term (fixed effects or hyperparameters)
- `glance()` — one row of model-level fit summaries
- `augment()` — original data extended with posterior fitted values

These make it straightforward to work with inlabru results inside tidy
workflows (dplyr, ggplot2, etc.) without manually unpacking the `bru` object.

## Set up

```{r libraries, eval=inla_available, message=FALSE, warning=FALSE}
library(INLA)
library(inlabru)
library(ggplot2)
bru_options_set(control.compute = list(dic = TRUE, waic = TRUE))
```

We use a simple simulated Gaussian dataset throughout.

```{r simulate, eval=inla_available}
set.seed(42L)
n <- 100
x <- runif(n, 0, 5)
y <- 2 + 1.5 * x + rnorm(n, sd = 0.8)
dat <- data.frame(x = x, y = y)
```

## Fit a model

```{r fit, eval=inla_available, results="hide"}
fit <- bru(
  y ~ Intercept(1) + slope(x, model = "linear"),
  family = "gaussian",
  data = dat
)
```

## `tidy()` — fixed effects

`tidy()` returns a tibble with one row per fixed-effect term, including the
posterior mean, standard deviation, and 95 % credible interval.

```{r tidy-fixed, eval=inla_available}
library(generics)
tidy(fit)
```

The column names follow the broom convention so the output integrates with
standard tidy tools.  For example, a quick coefficient plot:

```{r coef-plot, eval=inla_available}
tidy(fit) |>
  ggplot(aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(title = "Fixed-effect posterior summaries", x = NULL, y = "Estimate")
```

## `tidy()` — hyperparameters

Pass `effects = "hyperpar"` to extract the precision (or other hyperparameters)
instead.

```{r tidy-hyperpar, eval=inla_available}
tidy(fit, effects = "hyperpar")
```

The Gaussian family has one hyperparameter — the precision of the likelihood.
More complex models (e.g. with SPDE random fields) will have additional rows.

## `glance()` — model-level summaries

`glance()` returns a single-row tibble with the DIC, WAIC, marginal log-likelihood,
the number of observations, and total wall-clock time.

```{r glance, eval=inla_available}
glance(fit)
```

This is useful when comparing competing models:

```{r compare, eval=inla_available, results="hide"}
fit_null <- bru(
  y ~ Intercept(1),
  family = "gaussian",
  data = dat
)
```

```{r compare-show, eval=inla_available}
library(dplyr)
bind_rows(
  glance(fit_null) |> mutate(model = "intercept only"),
  glance(fit)      |> mutate(model = "intercept + slope")
) |>
  select(model, dic, waic, marginal_loglik, nobs)
```

The model with `slope` should have a substantially lower DIC and WAIC.

## `augment()` — fitted values on new data

`augment()` adds posterior summaries of the linear predictor to a data frame.
Because inlabru prediction expressions are arbitrary R code, you must supply a
`pred_formula` that names what to predict.

```{r augment, eval=inla_available, results="hide"}
grid <- data.frame(x = seq(0, 5, length.out = 50))

augmented <- augment(
  fit,
  data = grid,
  pred_formula = ~ Intercept_latent + slope_latent,
  n_samples = 500L,
  seed = 1L
)
```

```{r augment-show, eval=inla_available}
head(augmented)
```

The four added columns are `.fitted` (posterior mean), `.fitted_low` and
`.fitted_high` (2.5 % and 97.5 % quantiles), and `.fitted_sd` (posterior
standard deviation).

Plot the fitted line with a credible band on top of the raw data:

```{r augment-plot, eval=inla_available}
ggplot() +
  geom_point(data = dat, aes(x, y), alpha = 0.4, size = 0.8) +
  geom_ribbon(
    data = augmented,
    aes(x, ymin = .fitted_low, ymax = .fitted_high),
    fill = "steelblue", alpha = 0.3
  ) +
  geom_line(data = augmented, aes(x, .fitted), colour = "steelblue", linewidth = 1) +
  labs(
    title = "Posterior fitted line with 95 % credible band",
    x = "x", y = "y"
  )
```

## Naming convention for `pred_formula`

The latent variables in `pred_formula` follow inlabru's naming rule: a
component named `foo` is available as `foo_latent` inside prediction
expressions.  Components with `model = "linear"` (including the implicit
intercept) appear in `summary.fixed`; components with a random-effect model
(iid, rw1, SPDE, …) appear in `summary.random`.

```{r names, eval=inla_available}
# fixed-effect component names → use <name>_latent in pred_formula
names(fit$summary.fixed)

# random-effect component names (none in this model)
names(fit$summary.random)
```

## Calling tidiers with the `broom` package

The tidiers are registered as S3 methods on `bru` objects through the
[generics](https://cran.r-project.org/package=generics) package, so they work
automatically when you load either `generics` or `broom`.

```{r broom, eval=inla_available && requireNamespace("broom", quietly = TRUE)}
library(broom)
tidy(fit)
glance(fit)
```
