---
title: "LGCPs - An example in space and time"
author: "Fabian E. Bachl and Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - An example in space and time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
```

Introduction
----------------
For this vignette we are going to be working with a dataset obtained from 
the `R` package `MRSea`. We will set up a LGCP with a spatio-temporal SPDE model to estimate species distribution.


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(ggplot2)
```

Get the data
-----------------------------------
Load the dataset, that has coordinates in UTM in kilometres:

```{r }
data(mrsea, package = "inlabru")
```

The points (representing animals) and the sampling regions of this dataset are associated with a season. Let's have a look at the observed points and sampling regions for all seasons:

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() +
  gg(mrsea$mesh) +
  gg(mrsea$boundary) +
  gg(mrsea$samplers) +
  gg(mrsea$points, size = 0.5) +
  coord_fixed() +
  facet_wrap(~season) +
  ggtitle("MRSea observation seasons")
```

Integration points
-----------------------------------

Because our model will take time (season) into account we have to construct the integration points for the LGCP accordingly. Using the ´group´ parameter we can let the ´ipoints´ function know that we would like to construct the integration points for each season independently. Note that omitting this step would simply aggregate all sampling regions over time.

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ips <- ipoints(mrsea$samplers, mrsea$mesh, group = "season")
```
Plot the integration points:
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ggplot() +
  gg(ips, aes(color = season)) +
  facet_wrap(~season) +
  coord_equal()
```

Fitting the model
-----------------------------------

Fit an LGCP model to the locations of the animals. In this example we will employ a spatio-temporal SPDE. Note how the `group` and `ngroup` parameters are employed to let the SPDE model know about the name of the time dimension (season) and the total number of distinct points in time. 

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
matern <- inla.spde2.pcmatern(mrsea$mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(10, 0.01)
)

cmp <- coordinates + season ~ mySmooth(coordinates,
  model = matern, group = season, ngroup = 4
) +
  Intercept

fit <- lgcp(cmp, mrsea$points, ips = ips)
```

Predict and plot the intensity for all seasons:

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ppxl <- pixels(mrsea$mesh, mask = mrsea$boundary)
ppxl_all <- cprod(ppxl, data.frame(season = seq_len(4)))

lambda1 <- predict(fit, ppxl_all, ~ data.frame(season = season, lambda = exp(mySmooth + Intercept)))
```
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
pl1 <- ggplot() +
  gg(lambda1, aes(fill = mean)) +
  gg(mrsea$points, size = 0.3) +
  facet_wrap(~season) +
  coord_equal()
pl1
```
