---
title: "LGCPs - An example in two dimensions"
author: "David Borchers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Introduction
----------------
For this vignette we are going to be working with the inlabru's ´gorillas´ dataset which was originally obtained from 
the `R` package `spatstat`. The data set contains two types of gorillas nests which are marked as either major or minor. We will set up a multi-likelihood model for these nests which creates two spatial LGCPs that share a common intercept but have employ different spatial smoothers.


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
```

Use empricial Bayes in order to speed things up a little bit:
```{r results="hide",warning=FALSE,message=FALSE}
init.tutorial()
```


Get the data
-----------------------------------
For the next few practicals we are going to be working with a dataset obtained from 
the `R` package `spatstat`, which contains the locations of 647 gorilla nests. We load the 
dataset like this:

```{r }
data(gorillas, package = "inlabru")
```

Plot the nests and visualize the group membership (major/minor) by color:

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() + gg(gorillas$mesh) + 
           gg(gorillas$nests, aes(color=group)) + 
           gg(gorillas$boundary) + 
           coord_fixed() +
           ggtitle("Gorillas nests and group membership")
```


Fiting the model
-----------------------------------

First, we define all components that enter the joint model. That is, the intercept that is common to both LGCPs and
the two different spatial smoothers, one for each nest group.

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
matern <- inla.spde2.pcmatern(gorillas$mesh, 
                              prior.sigma = c(0.1, 0.01), 
                              prior.range = c(5, 0.01))

cmp <- coordinates ~ mySmoothMajor(map = coordinates, model = matern) + 
                     mySmoothMinor(map = coordinates, model = matern) + 
                     Intercept

cmp <- coordinates ~ mySmoothMajor(model = "linear") + 
                     mySmoothMinor(model = "linear") + 
                     Intercept
```

Given these components we define the linear predictor for each of the likelihoods
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
fml.major = coordinates ~ mySmoothMajor + Intercept
fml.minor = coordinates ~ mySmoothMajor + Intercept
```

Setting up the cox process integration points is easy in this example. Both nest types were observed within the same window.
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ips = ipoints(gorillas$boundary)
```
Lastly, we define the two likelihoods...
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
lik_minor = like("cp", formula = fml.major, components = cmp, data = gorillas$nests, ips = ips)
lik_major = like("cp", formula = fml.minor, components = cmp, data = gorillas$nests, ips = ips)
```
... which we provide to the ´bru´ function.
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
jfit = bru(cmp, lik_major, lik_minor)
```
