---
title: "Residual Analysis of spatial point process models using Bayesian methods"
author: "Niharika Reddy Peddinenikalva"
date: "`r Sys.Date()`"
output: html_document
header-includes: 
  - \newcommand{\R}{\mathbb{R}}
  - \newcommand{\C}{\mathbb{C}}
  - \newcommand{\Z}{\mathbb{Z}}
  - \newcommand{\E}{\mathbb{E}}
  - \newcommand{\probP}{\text{I\kern-0.15em P}}
  - \newcommand{\abs}[1]{\lvert #1 \rvert}
  - \newcommand{\vb}[1]{\mathbf{#1}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  dev = "png",
  dev.args = list(type = "cairo-png")
)
library("INLA")
library("inlabru")
library("RColorBrewer")
library("ggplot2")
library("dplyr")
library("lwgeom")
library("patchwork")
library("terra")
library("data.table")
bru_safe_sp(force = TRUE)
```

```{r, eval=TRUE, echo=FALSE}
# Determine function location
code_location <- "2d_lgcp_residuals_functions.R"
if (!file.exists(code_location)) {
  code_location <- file.path("vignettes", "web", code_location)
}
```

```{r load-functions, file=code_location, eval=TRUE, echo=FALSE}
# Do not change this code chunk
# Load function definitions
source(code_location)
```
```{r , eval=FALSE, echo=TRUE}
# Load function definitions
source("2d_lgcp_residuals_functions.R")
```

# Introduction
Point processes are useful for spatial data analysis and have wide applications 
in ecology, epidemiology, seismology, computational neuroscience, etc. 
Residual analysis is an effective assessment method for spatial point processes, which commonly takes a frequentist approach.

In this vignette, we will calculate residuals of some of the models from https://inlabru-org.github.io/inlabru/articles/web/2d_lgcp_covars.html
using a Bayesian approach to the residual methods in "Residual Analysis for 
spatial point processes" (Baddeley et al). 

# Theory of Residuals for Spatial point process models

Consider a spatial point pattern $\mathbf{x} = \{x_1, \dots, x_n \}$ of $n$ 
points in a bounded region $W$ of $\R^2$. For a model of a spatial point 
process $\vb{X}$ with probability density $f_{\theta}$ and parameter $\theta$, 
the innovation process of the model is defined by
\begin{equation}\label{eq:innovations}
    I_{\theta}(B) = n(\vb{X} \cap B) - \int_B \lambda_{\theta} (u) \, du
\end{equation}
Here, $\int_B \lambda_{\theta}(u) \, du$ is the expected number of points of the
fitted model in the bounded subset of $B$. 
Innovations satisfy $\E_{\theta} \left[ I_{\theta}(B)\right] = 0$ .

Given data $\vb{x}$ of the model and that the parameter estimate is 
$\hat{\theta}$, the raw residuals are given by 
\begin{equation}\label{eq:raw residuals}
    R_{\hat{\theta}}(B) = n(\vb{x} \cap B) - \int_B \hat{\lambda}(u) \,du
\end{equation}

Increments of the innovation process $I$ and raw residuals of the Poisson 
processes $R$ are analogous to errors and raw residuals of linear models 
respectively. 

## Types of Residuals

We can scale the raw residual by scaling the increments of $R_{\hat{\theta}}$. \\
The $h$-weighted innovations are given by
\begin{equation}\label{eq:h innovations}
    I(B, h, \lambda) = \sum_{x_i \in \vb{X} \cap B} h(x_i) - 
    \int_B h(u) \lambda(u) \,du
\end{equation}
This leads to the $h$-weighted residuals
\begin{equation}
\label{eq:h residuals}
    R(B, \hat{h}, \hat{\theta}) = I(B, \hat{h}, \hat{\lambda}) = 
    \sum_{x_i \in \vb{x} \cap B} h(x_i) - \int_B \hat{h}(u) \hat{\lambda}(u)\,du
\end{equation}
Since the innovation has mean 0, a true model would yield 
\[
\E \left[R(B, \hat{h}, \hat{\theta})\right] \approx 0.
\]
Changing the choice of the weight function $h$, yields  different types of 
residuals. 

### Scaling Residuals
By taking $h(u) = \vb{1} \{ u \in B\}$, we get the raw residuals:
\begin{aligned}
    R(B, \hat{h}, \hat{\theta}) & = I(B, \hat{h}, \hat{\lambda}) = 
    \sum_{x_i \in \vb{x} \cap B}h(x_i) - \int_B \hat{h}(u)\hat{\lambda}(u)\,du\\
    & = n(\vb{x} \cap B) - \int_B \hat{\lambda}(u) \,du
\end{aligned}

### Inverse Residuals
By taking $h(u) = \frac{1}{\lambda (u)}$, we may encounter the case where 
$\lambda(u) = 0$. The GNZ-formula
\begin{equation}
\E \left[ \sum_{x_i \in \vb{X}} h(x_i)\right] = 
\E \left[ \int_W h(u) \lambda(u) \,du \right],
\end{equation}
would still hold for the first terms of the $h$-weighted residuals to exist. If
we define $h(u) \lambda(u) = 0$ for all $u$ such that $\lambda(u) = 0$, then the second term of $h$-weighted residuals will also exist and the inverse residual 
is given by
\begin{aligned}
    R\left(B, \frac{1}{\hat{\lambda}}, \hat{\theta}\right) & = 
    \sum_{x_i \in \vb{x} \cap B} \hat{h}(x_i) - 
    \int_B \hat{h}(u) \hat{\lambda}(u) \,du \\
    & = \sum_{x_i \in \vb{x} \cap B} \frac{1}{\hat{\lambda}(x_i)} - 
    \int_B \vb{1} \{ x_i \in \vb{x} \} \,du 
\end{aligned}
    
### Pearson Residuals
By using $h(u) = \frac{1}{\sqrt{\lambda(u)}}$, the Pearson residual is given by
\begin{aligned}
    R\left( B, \frac{1}{\sqrt{\hat{\lambda}}}, \hat{\theta} \right) 
    & = \sum_{x_i \in \vb{x} \cap B} \hat{h}(x_i) - 
    \int_B \hat{h}(u) \hat{\lambda}(u) \,du \\
    & = \sum_{x_i \in \vb{x} \cap B} \frac{1}{\sqrt{\hat{\lambda}(x_i)}} -
    \int_B \sqrt{\hat{\lambda}(u)} \,du 
\end{aligned}
Here, zero values of $\hat{\lambda}(u)$ do not cause any issues with calculating residuals as we set $\hat{h}(u) \hat{\lambda}(u) = \sqrt{\hat{\lambda}(u)}$ for 
all $u$.

So the three types of residuals are:
\begin{eqnarray}
\text{Scaled:} \qquad & R(B, \hat{h}, \hat{\theta}) & = n(\vb{x} \cap B) - 
\int_B \hat{\lambda}(u) \,du \\
\text{Inverse:} \qquad & R\left(B, \frac{1}{\hat{\lambda}}, \hat{\theta}\right) 
& = \sum_{x_i \in \vb{x} \cap B} \frac{1}{\hat{\lambda}(x_i)} - 
\int_B \vb{1} \{ x_i \in \vb{x} \} \,du \\
\text{Pearson:} \qquad &  R\left( B, \frac{1}{\sqrt{\hat{\lambda}}},\hat{\theta} \right) & = \sum_{x_i \in \vb{x} \cap B} \frac{1}{\sqrt{\hat{\lambda}(x_i)}} - 
\int_B \sqrt{\hat{\lambda}(u)} \,du 
\end{eqnarray}

Note that $\hat{\lambda}(u)$ and $\hat{h}(u)$ are estimates of $\lambda (u)$ and
$h(u)$ respectively

## Motivation for choice of residuals
The scaling residuals are nothing but the raw residuals and hence depict the 
residual data as is. However, the Pearson residuals use a weighted function 
$h(u) = \frac{1}{\lambda(u)}$ so this helps to obtain normalised residuals since
the denominator accounts for variance of the residuals and hence makes it 
reliable for any sample size. 

The use of the inverse residual in this project is less clear. However, we do 
know that it is easier to compute this residual since we have the GNZ formula to
estimate the value of $h(u) = \frac{1}{\lambda(u)} for any value $\lambda(u)$. 

For notation's sake in this discussion, let $h_{s}(u), h_{i}(u), h_p(u)$ 
indicate the weight function for scaling, inverse and Pearson residuals 
respectively. 
Then, $h_s(u) = \vb{1}\{u \in B\}, h_i(u) = \frac{1}{\lambda(u)}, h_p(u) = \frac{1}{\sqrt{\lambda(u)}}$. 
We can also see that $h_s(u) = \left(h_i(u)\right)^0$  and $h_p(u) = 
\left(h_i(u)\right)^{1/2}$, so the values of the Pearson residuals would lie 
somewhere between the other two residuals. 
So, in our case, the inverse residuals would not be very useful, but we do find 
that they are easier to compute by hand in terms of estimates.


# Computation of Residuals
In this vignette, the residuals of some models from 
https://inlabru-org.github.io/inlabru/articles/web/2d_lgcp_covars.html
are computed using a Bayesian approach of the residuals described above. 
This is done using the `inlabru` package. The models used are:

* Model 1 (Vegetation Model): 
    The model is defined such that the gorilla nests have vegetation type as a 
    fixed covariate.
    
* Model 2 (Elevation Model): 
    The model is defined such that the gorilla nests have elevation as a 
    continuous variable. 
    
* Model 3 (Intercept Model): 
    The model is defined such that the gorilla nests have a constant effect. 
    
* Model 4 (SPDE Smooth type model): 
    The model is defined such that the gorilla nests depend on an SPDE type 
    smooth function. 
    

The functions used to calculate the residuals of these models are given in the 
Code appendix at the end of this vignette.


## Loading gorillas data
This vignette uses the `gorillas` dataset from the "inlabru" package in R for 
modelling spatial Poisson processes. 
Here, the ``inlabru" package uses Latent Gaussian Cox Processes to model the 
data using different model definitions.

```{r load-data, echo=TRUE}
# Load data
data(gorillas, package = "inlabru")
nests <- gorillas$nests
mesh <- gorillas$mesh
boundary <- gorillas$boundary
gcov <- gorillas$gcov
```

Initially, the set $B$ is defines as $B=W$, where $W$ is the object `boundary`. 
The function `prepare_residual_calculations()` is defined to compute a data 
frame of observations $x_i$ and sample points $u \in W$ to compute the residual,
a matrix `A_sum` that helps to calculate the summation term of the residual and 
a matrix `A_integrate` that helps to calculate the integral term of the residual.

The function `residual_df` is defined to compute all three types of residuals
for a given model and choice of $B$ given a set of observations.

```{r preparatory-function, echo=TRUE}
# Define the subset B
B <- boundary

# Store the required matrices and data frames for residual computation
As <- prepare_residual_calculations(
  samplers = B, domain = mesh,
  observations = nests
)
```


## Assessing 4 models with residuals
### Vegetation Model
This model is taken from 
https://inlabru-org.github.io/inlabru/articles/web/2d_lgcp_covars.html .
The model is defined such that the gorilla nests have vegetation type as a 
fixed covariate.

The code chunk below shows how the model is defined by `lgcp()` and also 
displays the model in the form of a plot.
The residuals of the model are also computed where $B = W$.

```{r veg-mod, echo=TRUE}
# Define the vegetation model
comp1 <- coordinates ~
  vegetation(gcov$vegetation, model = "factor_contrast") + Intercept(x * 0 + 1)

fit1 <- lgcp(comp1, nests,
  samplers = boundary,
  domain = list(coordinates = mesh)
)

# Display the model
int1 <- predict(fit1,
  newdata = fm_pixels(mesh, mask = boundary, format = "sp"),
  ~ exp(vegetation + Intercept)
)
ggplot() +
  gg(int1) +
  gg(boundary, alpha = 0, lwd = 2) +
  gg(nests, color = "DarkGreen") +
  coord_equal()

# Calculate the residuals for the vegetation model
veg_res <- residual_df(
  fit1, As$df, expression(exp(vegetation + Intercept)),
  As$A_sum, As$A_integrate
)
knitr::kable(edit_df(veg_res, c("Type", "mean.mc_std_err", "sd.mc_std_err", "median")))
```

Note: The data frame produced by `residual_df()` originally contains the columns
`Type`, `mean.mc_std_err`, `sd.mc_std_err` and `median` which have been removed
in all the tables in the vignette to highlight only essential and non-recurring 
data. This editing of the data frames is done by the function `edit_df`.

### Elevation model
This model is also taken from 
https://inlabru-org.github.io/inlabru/articles/web/2d_lgcp_covars.html .
The model is defined such that the gorilla nests have elevation as a 
continuous variable. 

```{r elevation-model-setup, echo=FALSE}
# Setting up the elevation model
elev <- gcov$elevation
elev$elevation <- elev$elevation - mean(elev$elevation, na.rm = TRUE)
matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(5, 0.01)
)
```

The code chunk below shows how the model is defined by `lgcp()` and also 
displays the model in the form of a plot. 

```{r elev-mod, echo=TRUE}
# Define the Elevation model
comp2 <- coordinates ~ elev(elev, model = "linear") +
  mySmooth(coordinates, model = matern) + Intercept(1)

fit2 <- lgcp(comp2, nests,
  samplers = boundary,
  domain = list(coordinates = mesh)
)

# Display the model
int2 <- predict(
  fit2,
  fm_pixels(mesh, mask = boundary, format = "sp"),
  ~ exp(elev + mySmooth + Intercept)
)
ggplot() +
  gg(int2) +
  gg(boundary, alpha = 0) +
  gg(nests, shape = "+") +
  coord_equal()
```

The residuals of the model when $B = W$ are given below:

```{r elev-res, echo=FALSE}
# Calculate the residuals for the model
elev_res <- residual_df(
  model = fit2, df = As$df,
  expr = expression(exp(elev + mySmooth + Intercept)),
  A_sum = As$A_sum, A_integrate = As$A_integrate
)
knitr::kable(edit_df(
  elev_res,
  c("Type", "mean.mc_std_err", "sd.mc_std_err", "median")
))
```

### Intercept Model
This model is also taken from 
https://inlabru-org.github.io/inlabru/articles/web/2d_lgcp_covars.html .
The model is defined such that the gorilla nests have a constant effect. 

The code chunk below shows how the model is defined by `lgcp()` and also 
displays the model in the form of a plot. The residuals of the model are also 
computed where $B = W$.

```{r int-mod, echo=TRUE}
# Define the Intercept model
comp3 <- coordinates ~ Intercept(rep(1, nrow(.data.)))
fit3 <- lgcp(comp3, nests,
  samplers = boundary,
  domain = list(coordinates = mesh)
)

# Display the model
int3 <- predict(
  fit3,
  fm_pixels(mesh, mask = boundary, format = "sp"),
  ~ exp(Intercept)
)
ggplot() +
  gg(int3) +
  gg(boundary, alpha = 0) +
  gg(nests, shape = "+") +
  coord_equal()
```

The residuals of the model when $B = W$ are given below:

```{r int-res, echo=FALSE}
# Calculate the residuals for the model
int_res <- residual_df(
  model = fit3, df = As$df,
  expr = expression(exp(Intercept)),
  A_sum = As$A_sum, A_integrate = As$A_integrate
)
knitr::kable(edit_df(
  int_res,
  c("Type", "mean.mc_std_err", "sd.mc_std_err", "median")
))
```

### Smooth Model
This model is also taken from 
https://inlabru-org.github.io/inlabru/articles/web/2d_lgcp_covars.html .
The model is defined such that the gorilla nests depend on an SPDE type smooth function. 

The code chunk below shows how the model is defined by `lgcp()` and also 
displays the model in the form of a plot. The residuals of the model are also 
computed where $B = W$.

```{r smooth-mod, echo=TRUE}
# Define the Smooth model
comp4 <- coordinates ~ mySmooth(coordinates, model = matern) +
  Intercept(1)
fit4 <- lgcp(comp4, nests,
  samplers = boundary,
  domain = list(coordinates = mesh)
)

# Display the model
int4 <- predict(
  fit4,
  fm_pixels(mesh, mask = boundary, format = "sp"),
  ~ exp(mySmooth + Intercept)
)
ggplot() +
  gg(int4) +
  gg(boundary) +
  gg(nests, shape = "+") +
  coord_equal()
```

The residuals of the model when $B = W$ are given below:

```{r smooth-res, echo=FALSE}
# Calculate the residuals for the model
smooth_res <- residual_df(
  model = fit4, df = As$df,
  expr = expression(exp(mySmooth + Intercept)),
  A_sum = As$A_sum, A_integrate = As$A_integrate
)
knitr::kable(edit_df(
  smooth_res,
  c("Type", "mean.mc_std_err", "sd.mc_std_err", "median")
))
```
### Comparing models
Firstly, from the plots of the four models, we see that there are visual 
similarities in what the Elevation and smooth models appear to be. So, we would expect that these two models show similar trends in residuals. 


## Redefining the set B

Consider defining a new type of set $B$ which consists of two subpolygons within
the boundary. This is done by the `partition()` function which divides a polygon
into grids based on number of rows and columns or by a desired resolution. 
The code chunk below demonstrates a use of the function for dividing the polygon
and calculating the residuals for this choice of $B$.

```{r partitioning-B-two, echo=TRUE}
# Create a grid for B partitioned into two
B1 <- partition(samplers = boundary, nrows = 1, ncols = 2)
plot(B1, main = "Two partitions of B")

As1 <- prepare_residual_calculations(
  samplers = B1, domain = mesh,
  observations = nests
)
# Residuals for the vegetation model
veg_res2 <- residual_df(
  fit1, As1$df, expression(exp(vegetation + Intercept)),
  As1$A_sum, As1$A_integrate
)
knitr::kable(edit_df(veg_res2, c(
  "Type", "mean.mc_std_err",
  "sd.mc_std_err", "median"
)))
```
Here, it can be seen that there are two lines of residual data for each type of 
residual. This signifies that the `residual_df` function has calculated 
residuals for each partition. To compare these residuals effectively, a 
function `residual_plot()` is defined that plots the residuals for each 
corresponding polygon of `B`. This is discussed later in the vignette.

Another type of partitioning is considered with three sections of the region and 
its residuals are as displayed below:
```{r partitioning-B-three, echo=FALSE}
# Create a grid for B partitioned into three
B2 <- partition(samplers = boundary, nrows = 3, ncols = 1)
plot(B2, main = "Three partitions of B")

As2 <- prepare_residual_calculations(
  samplers = B2, domain = mesh,
  observations = nests
)
# Residuals for the vegetation model
veg_res3 <- residual_df(
  fit1, As2$df, expression(exp(vegetation + Intercept)),
  As2$A_sum, As2$A_integrate
)
knitr::kable(edit_df(veg_res3, c("Type", "mean.mc_std_err", "sd.mc_std_err", "median")))
```



## Residuals of models at different resolutions
Consider two more resolutions of $B$, given in the plots below.  Here we use the
`resolution` argument instead of `nrow` and `ncol`. Due to the way `terra::rast()`
interprets the arguments, using `resolution` doesn't guarantee complete coverage of the
`samplers` polygon.
```{r more-resolutions, echo=FALSE}
# Create two more partitions of B using two resolutions
B3 <- partition(samplers = boundary, resolution = c(0.5, 0.5))
plot(B3, main = "B3")

# Calculate preparatory data for residuals at this resolution
As3 <- prepare_residual_calculations(
  samplers = B3, domain = mesh,
  observations = nests
)


B4 <- partition(samplers = boundary, resolution = c(2, 2))
plot(B4, main = "B4")

# Calculate preparatory data for residuals at this resolution
As4 <- prepare_residual_calculations(
  samplers = B4, domain = mesh,
  observations = nests
)
```


Using these two new choices of $B$, the residuals are calculated for all four 
models at both resolutions using `residual_df()`.
```{r calc-res-all-models-two-resolutions, echo=FALSE}
# Vegetation model
Residual_fit1_B3 <- residual_df(
  model = fit1, df = As3$df,
  expr = expression(exp(vegetation + Intercept)), A_sum = As3$A_sum,
  A_integrate = As3$A_integrate
)

Residual_fit1_B4 <- residual_df(
  model = fit1, df = As4$df,
  expr = expression(exp(vegetation + Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)


# Elevation model
Residual_fit2_B3 <- residual_df(
  model = fit2, df = As3$df,
  expr = expression(exp(elev + mySmooth + Intercept)),
  A_sum = As3$A_sum, A_integrate = As3$A_integrate
)

Residual_fit2_B4 <- residual_df(
  model = fit2, df = As4$df,
  expr = expression(exp(elev + mySmooth + Intercept)),
  A_sum = As4$A_sum, A_integrate = As4$A_integrate
)


# Intercept model
Residual_fit3_B3 <- residual_df(
  model = fit3, df = As3$df,
  expr = expression(exp(Intercept)), A_sum = As3$A_sum,
  A_integrate = As3$A_integrate
)

Residual_fit3_B4 <- residual_df(
  model = fit3, df = As4$df,
  expr = expression(exp(Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)


# Smooth + Intercept
Residual_fit4_B3 <- residual_df(
  model = fit4, df = As3$df,
  expr = expression(exp(mySmooth + Intercept)), A_sum = As3$A_sum,
  A_integrate = As3$A_integrate
)

Residual_fit4_B4 <- residual_df(
  model = fit4, df = As4$df,
  expr = expression(exp(mySmooth + Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)
```

The functions `set_csc()` and `residual_plot()` are defined to plot the three 
types of residuals for the model for each partition in $B$. The code chunk below
demonstrates how this plotting is done for the vegetation model at resolution 
$B = B_3$. Here, it is useful for each type of residual to have the same colour
scale at both resolutions.

```{r store-res-plot-1, echo = FALSE}
# Colour scales for each model to cover both resolutions
fit1_csc <- set_csc(rbind(Residual_fit1_B3, Residual_fit1_B4), rep("RdBu", 3))
fit2_csc <- set_csc(rbind(Residual_fit2_B3, Residual_fit2_B4), rep("RdBu", 3))
fit3_csc <- set_csc(rbind(Residual_fit3_B3, Residual_fit3_B4), rep("RdBu", 3))
fit4_csc <- set_csc(rbind(Residual_fit4_B3, Residual_fit4_B4), rep("RdBu", 3))

# Plots for all 4 models with resolution B3
plotB3_fit1 <- residual_plot(B3, Residual_fit1_B3, fit1_csc, "Vegetation Model")
plotB3_fit2 <- residual_plot(B3, Residual_fit2_B3, fit2_csc, "Elevation Model")
plotB3_fit3 <- residual_plot(B3, Residual_fit3_B3, fit3_csc, "Intercept Model")
plotB3_fit4 <- residual_plot(B3, Residual_fit4_B3, fit4_csc, "Smooth Model")

# Plots for all 4 models with resolution B4
plotB4_fit1 <- residual_plot(B4, Residual_fit1_B4, fit1_csc, "Vegetation Model")
plotB4_fit2 <- residual_plot(B4, Residual_fit2_B4, fit2_csc, "Elevation Model")
plotB4_fit3 <- residual_plot(B4, Residual_fit3_B4, fit3_csc, "Intercept Model")
plotB4_fit4 <- residual_plot(B4, Residual_fit4_B4, fit4_csc, "Smooth Model")
```

```{r display-veg-plot-code, eval=FALSE,echo=TRUE, fig.height=2.5,fig.width=4.5}
# Residuals of Vegetation model
Residual_fit1_B3 <- residual_df(
  model = fit1, df = As3$df,
  expr = expression(exp(vegetation + Intercept)), A_sum = As3$A_sum,
  A_integrate = As3$A_integrate
)

Residual_fit1_B4 <- residual_df(
  model = fit1, df = As4$df,
  expr = expression(exp(vegetation + Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)

# Colour scales for each model to cover both resolutions
fit1_csc <- set_csc(rbind(Residual_fit1_B3, Residual_fit1_B4), rep("RdBu", 3))

# Store plots
plotB3_fit1 <- residual_plot(B3, Residual_fit1_B3, fit1_csc, "Vegetation Model")
plotB4_fit1 <- residual_plot(B4, Residual_fit1_B4, fit1_csc, "Vegetation Model")

# comparing the vegetation model
((plotB3_fit1$Scaling | plotB3_fit1$Inverse | plotB3_fit1$Pearson) /
  (plotB4_fit1$Scaling | plotB4_fit1$Inverse | plotB4_fit1$Pearson)) +
  plot_annotation(title = "Vegetation Model") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Comparing different types of residuals for each model
Now, the plots below help to compare the residuals for each model at two 
resolutions of $B$ for each type of residual.

```{r plot-all-1, echo=FALSE, fig.width=9, , fig.height=7}
# comparing the vegetation model
((plotB3_fit1$Scaling | plotB3_fit1$Inverse | plotB3_fit1$Pearson) /
  (plotB4_fit1$Scaling | plotB4_fit1$Inverse | plotB4_fit1$Pearson)) +
  plot_annotation(title = "Vegetation Model") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# comparing the elevation model
((plotB3_fit2$Scaling | plotB3_fit2$Inverse | plotB3_fit2$Pearson) /
  (plotB4_fit2$Scaling | plotB4_fit2$Inverse | plotB4_fit2$Pearson)) +
  plot_annotation(title = "Elevation Model") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# comparing the intercept model
((plotB3_fit3$Scaling | plotB3_fit3$Inverse | plotB3_fit3$Pearson) /
  (plotB4_fit3$Scaling | plotB4_fit3$Inverse | plotB4_fit3$Pearson)) +
  plot_annotation("Intercept model") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# comparing the smooth+intercept model
((plotB3_fit4$Scaling | plotB3_fit4$Inverse | plotB3_fit4$Pearson) /
  (plotB4_fit4$Scaling | plotB4_fit4$Inverse | plotB4_fit4$Pearson)) +
  plot_annotation("Smooth model") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

#### Discussion
In this section, we discuss some of the findings from the output in the code. 
A relevant point that would be used repeatedly in this discussion is that since 
the colour scales for the residuals are chosen so that negative and positive 
values are in hues of red and blue respectively, negative and positive residuals 
imply an overestimation and underestimation of gorilla nests by the model 
respectively. Thus, red regions in a plot would imply overestimation of gorilla 
nests by the model in that region and blue regions in a plot would imply 
underestimation of gorilla nests by the model in that region.

A common observation for all models is that the scaling residuals seem to have 
the highest range among the residuals since they are direct interpretations of 
residuals for all models.
Also, the Pearson residuals seem to take values between the scaling and inverse 
residuals for both resolutions of all four models.
However, for almost all types of residuals for each model, the positive and 
negative residual values are not very extreme for the $B_3$ resolution but these 
corresponding residual values are more extreme at the $B_4$ resolution for each 
type of residual. This could be because of the bias in the ``experimental" mode 
of \texttt{INLA} which is explored at other points in this project.

For the vegetation and intercept model, the regions which have negative and 
positive residuals remain the same for all residuals at different levels for 
each type of residual. Also, it seems that the north and north-west region of 
both models are underestimated by the model while the remaining regions are 
overestimated by the model at different levels.

For the elevation model, each type of residual has slight variations in negative
and positive residuals for corresponding regions in a fixed resolution. 
This means that different residuals suggest that a particular polygon in 
overestimated and underestimated by the model. However, this happens only at a 
few points and is not containing extremes of the residuals in the same regions. 
The Pearson residuals are supposedly the most reliable in this case as they are 
normalised residuals. 
All three types of residuals contain larger proportions of negative residuals 
which suggest that the model overestimates the gorilla population the region (at 
different amounts in different regions).

### Comparing different models for each type of residual
Another way to compare residuals, is to consider each type of residual 
separately and plot the residuals for all four models at each resolution. This 
helps to compare the models for a given type of residual and choice of $B$. 
Here, it is useful to have the same colour scale for all the 4 models.

```{r store-res-plot-2, echo=FALSE}
# Data frame of all models together for B3 and B4
B3res <- rbind(
  Residual_fit1_B3, Residual_fit2_B3,
  Residual_fit3_B3, Residual_fit4_B3
)
B4res <- rbind(
  Residual_fit1_B4, Residual_fit2_B4,
  Residual_fit3_B4, Residual_fit4_B4
)

# Colour scale for each resolution to cover all 4 models
B3csc <- set_csc(B3res, rep("RdBu", 3))
B4csc <- set_csc(B4res, rep("RdBu", 3))

# Plots for all 4 models with resolution B3
plotB3_veg <- residual_plot(B3, Residual_fit1_B3, B3csc, "Vegetation Model")
plotB3_elev <- residual_plot(B3, Residual_fit2_B3, B3csc, "Elevation Model")
plotB3_int <- residual_plot(B3, Residual_fit3_B3, B3csc, "Intercept Model")
plotB3_smooth <- residual_plot(B3, Residual_fit4_B3, B3csc, "Smooth Model")

# Plots for all 4 models with resolution B4
plotB4_veg <- residual_plot(B4, Residual_fit1_B4, B4csc, "Vegetation Model")
plotB4_elev <- residual_plot(B4, Residual_fit2_B4, B4csc, "Elevation Model")
plotB4_int <- residual_plot(B4, Residual_fit3_B4, B4csc, "Intercept Model")
plotB4_smooth <- residual_plot(B4, Residual_fit4_B4, B4csc, "Smooth Model")
```


The code chunk below demonstrates the code for plotting the "Pearson" residuals 
when $B = B_4$.

```{r disply-pearson-plot-code, eval = FALSE, echo = TRUE}
# Calculate the residuals for all models at B4
Residual_fit1_B4 <- residual_df(
  model = fit1, df = As4$df,
  expr = expression(exp(vegetation + Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)

Residual_fit2_B4 <- residual_df(
  model = fit2, df = As4$df,
  expr = expression(exp(elev + mySmooth + Intercept)),
  A_sum = As4$A_sum, A_integrate = As4$A_integrate
)

Residual_fit3_B4 <- residual_df(
  model = fit3, df = As4$df,
  expr = expression(exp(Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)

Residual_fit4_B4 <- residual_df(
  model = fit4, df = As4$df,
  expr = expression(exp(mySmooth + Intercept)), A_sum = As4$A_sum,
  A_integrate = As4$A_integrate
)

# Set the colour scale
B4res <- rbind(
  Residual_fit1_B4, Residual_fit2_B4,
  Residual_fit3_B4, Residual_fit4_B4
)
B4csc <- set_csc(B4res, rep("RdBu", 3))

# Plots for residuals of all 4 models with resolution B4
plotB4_veg <- residual_plot(B4, Residual_fit1_B4, B4csc, "Vegetation Model")
plotB4_elev <- residual_plot(B4, Residual_fit2_B4, B4csc, "Elevation Model")
plotB4_int <- residual_plot(B4, Residual_fit3_B4, B4csc, "Intercept Model")
plotB4_smooth <- residual_plot(B4, Residual_fit4_B4, B4csc, "Smooth Model")

# Comparing all models for B4 Pearson residuals
((plotB4_veg$Pearson | plotB4_elev$Pearson) /
  (plotB4_int$Pearson | plotB4_smooth$Pearson)) +
  plot_annotation("B4 Pearson Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Now, the plots below help to compare all 4 models at a particular choice of $B$
and for a chosen residual type.

```{r plot-all-2, fig.height=7, fig.width=7, echo=FALSE}
# Comparing all models for B3 Scaling residuals
((plotB3_veg$Scaling | plotB3_elev$Scaling) /
  (plotB3_int$Scaling | plotB3_smooth$Scaling)) +
  plot_annotation("B3 Scaling Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Comparing all models for B3 Inverse residuals
((plotB3_veg$Inverse | plotB3_elev$Inverse) /
  (plotB3_int$Inverse | plotB3_smooth$Inverse)) +
  plot_annotation("B3 Inverse Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Comparing all models for B3 Pearson residuals
((plotB3_veg$Pearson | plotB3_elev$Pearson) /
  (plotB3_int$Pearson | plotB3_smooth$Pearson)) +
  plot_annotation("B3 Pearson Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Comparing all models for B4 Scaling Residuals
((plotB4_veg$Scaling | plotB4_elev$Scaling) /
  (plotB4_int$Scaling | plotB4_smooth$Scaling)) +
  plot_annotation("B4 Scaling Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")


# Comparing all models for B4 Inverse residuals
((plotB4_veg$Inverse | plotB4_elev$Inverse) /
  (plotB4_int$Inverse | plotB4_smooth$Inverse)) +
  plot_annotation("B4 Inverse Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")


# Comparing all models for B4 Pearson residuals
((plotB4_veg$Pearson | plotB4_elev$Pearson) /
  (plotB4_int$Pearson | plotB4_smooth$Pearson)) +
  plot_annotation("B4 Pearson Residuals") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

#### Discussion
Firstly, a common feature in each of the three figures is that the residual 
plots for the vegetation and intercept models are relatively similar and those of the elevation and smooth models are relatively similar as well for both resolutions 
for any type of residual. 
Also, for a fixed type of residual and a chosen model, the $B_3$ residuals are 
less extreme compared to the $B_4$ residuals, plausibly owing to the bias in the 
``experimental" mode in \texttt{INLA}. 
Another feature in all three types of residuals is that the elevation and 
intercept model have less extreme residual values compared to the vegetation and 
intercept models for a given type of residual. This could suggest that the 
former two are better models for gorilla nesting in the given region. 
The residual values have a wider range when $B=B_4$ than when $B = B_3$. 

For the scaling residuals and the Pearson residuals, the elevation and smooth 
models have less extreme values of residuals relative the vegetation and 
intercept models. This suggests that the elevation and smooth models are more 
suited to estimating the gorilla nesting locations in $W$. Also, there is a 
significant underestimation in the north-west areas by the vegetation and 
intercept models while the same areas are not underestimated by the other two 
models. However, as explored in \cref{sec:comparing_inla_modes}, these residuals
are calculated in the ``experimental" mode of \texttt{INLA} which produces some
overestimates in these two models, so this must be kept in mind while reaching 
to conclusions about a suitable model for nesting locations.

For the inverse residuals, the same observations hold, except the residuals for 
the elevation and smooth models have residuals that are not significantly less
extreme relative to the vegetation and intercept models as seen in the case of 
the Pearson and scaling residuals for both resolutions $B_3$ and $B_4$.


# Code appendix

## Function definitions

```{r file=code_location, eval=FALSE, echo=TRUE}
# Do not change this code chunk
```
