---
title: "LGCPs - An example in space and time"
author: "Fabian E. Bachl"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Introduction
----------------
For this vignette we are going to be working with a dataset obtained from 
the `R` package `MRSea`. We will set up a LGCP with a spatio-temporal SPDE model to estimate species distribution.


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
```

Use empricial Bayes in order to speed things up a little bit:
```{r results="hide",warning=FALSE,message=FALSE}
init.tutorial()
```


Get the data
-----------------------------------
Load the dataset:

```{r }
data(mrsea, package = "inlabru")
```

The points (representing animals) and the sampling regions of this dataset are associated with a season. Let's have a look at the observed points and sampling regions for all seasons:

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() + gg(mrsea$mesh) +
           gg(mrsea$boundary) + 
           gg(mrsea$samplers) +
           gg(mrsea$points, size = 0.5) + 
           coord_fixed() + 
           facet_wrap(~ season) +
           ggtitle("MRSea observation seasons")
```

Integration points
-----------------------------------

Because our model will take time (season) into account we have to construct the integration points for the LGCP accordingly. Using the ´group´ parameter we can let the ´ipoints´ function know that we would like to construct the integration points for each season inteprendently. Note that omitting this step would simply aggregate all sampling regions over time.

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ips <- ipoints(mrsea$samplers, group = "season")
```
Plot the integration points:
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ggplot() + gg(ips, aes(color = season)) + facet_wrap(~season) + coord_equal()
```

Fitting the model
-----------------------------------

Fit an LGCP model to the locations of the animals. In this example we will employ a spatio-temporal SPDE. Note how the ´group´ and ´ngroup´ parameters are employed to let the SPDE model know about the name of the time dimension (season) and the total number of distinct points in time. 

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
matern <- inla.spde2.pcmatern(mrsea$mesh, 
                              prior.sigma = c(0.1, 0.01), 
                              prior.range = c(10000, 0.01))

season = mrsea$points$season # Workaround for known bug (fixed in new backend)

cmp <- coordinates + season ~ mySmooth(map = coordinates,
                              model = matern, group = season, ngroup = 4) +
                          Intercept

fit <- lgcp(cmp, mrsea$points, ips = ips)
```

Predict and plot the intensity for all seasons:

```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
ppxl <- pixels(mrsea$mesh, mask = mrsea$boundary)
ppxl1 <- SpatialPixelsDataFrame(ppxl, data = data.frame(season = rep(1, nrow(data.frame(ppxl)))))
ppxl2 <- SpatialPixelsDataFrame(ppxl, data = data.frame(season = rep(2, nrow(data.frame(ppxl)))))
ppxl3 <- SpatialPixelsDataFrame(ppxl, data = data.frame(season = rep(3, nrow(data.frame(ppxl)))))
ppxl4 <- SpatialPixelsDataFrame(ppxl, data = data.frame(season = rep(4, nrow(data.frame(ppxl)))))

lambda1 <- predict(fit, ppxl1, ~ exp(mySmooth + Intercept))
lambda2 <- predict(fit, ppxl2, ~ exp(mySmooth + Intercept))
lambda3 <- predict(fit, ppxl3, ~ exp(mySmooth + Intercept))
lambda4 <- predict(fit, ppxl4, ~ exp(mySmooth + Intercept))

pl1 <- ggplot() + gg(lambda1, aes(fill=mean)) + gg(mrsea$points[mrsea$points$season == 1,], size = 0.3) + coord_equal()
pl2 <- ggplot() + gg(lambda2, aes(fill=mean)) + gg(mrsea$points[mrsea$points$season == 2,], size = 0.3) + coord_equal()
pl3 <- ggplot() + gg(lambda3, aes(fill=mean)) + gg(mrsea$points[mrsea$points$season == 3,], size = 0.3) + coord_equal()
pl4 <- ggplot() + gg(lambda4, aes(fill=mean)) + gg(mrsea$points[mrsea$points$season == 4,], size = 0.3) + coord_equal()

multiplot(pl1, pl2, pl3, pl4, cols = 2)
```
