---
title: "LGCPs - Spatial covariates"
author: "David Borchers and Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Spatial covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
.cache_dir <- file.path("cache", "2d_lgcp_covars")
.cache_recompute <- !file.exists(.cache_dir)
if (.cache_recompute) {
  dir.create(.cache_dir, recursive = TRUE)
}
.vignette_cache <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```
Set things up

```{r results="hide",warning=FALSE,message=FALSE}
library(INLA)
library(inlabru)
library(RColorBrewer)
library(ggplot2)

bru_options_set(inla.mode = "experimental")
```

Introduction
----------------------------------

We are going to fit spatial models to the gorilla data, using factor and continuous explanatory
variables in this practical. We will fit one using the factor variable `vegetation`, the other using
the continuous covariate `elevation`

(Jump to the bottom of the practical if you want to start gently with a 1D example!)

Get the data
-----------------------------------

```{r }
data(gorillas, package = "inlabru")
```

This dataset is a list (see `help(gorillas)` for details. Extract the objects
you need from the list, for convenience:

```{r }
nests <- gorillas$nests
mesh <- gorillas$mesh
boundary <- gorillas$boundary
gcov <- gorillas$gcov
```

Factor covariates
-----------------------------------

Look at the vegetation type, nests and boundary:


```{r warning=FALSE,message=FALSE}
ggplot() +
  gg(gcov$vegetation) +
  gg(boundary) +
  gg(nests, color = "white", cex = 0.5) +
  coord_equal()
```

Or, with the mesh:


```{r warning=FALSE,message=FALSE}
ggplot() +
  gg(gcov$vegetation) +
  gg(mesh) +
  gg(boundary) +
  gg(nests, color = "white", cex = 0.5) +
  coord_equal()
```

#### A model with vegetation type only

It seems that vegetation type might be a good predictor because nearly all the nests fall in 
vegetation type `Primary`. So we construct a model with vegetation type as a fixed effect.
To do this, we need to tell 'lgcp' how to find the vegetation type at any point in
space, and we do this by creating model components with a fixed effect that we
call `vegetation` (we could call it
anything), as follows:


```{r warning=FALSE,message=FALSE}
comp1 <- coordinates ~ vegetation(gcov$vegetation, model = "factor_full") - 1
```

Notes:
* We need to tell 'lgcp' that this is a factor fixed effect, which we do 
with `model="factor_full"`, giving one coefficient for each factor level.
* We need to be careful about overparameterisation when using factors.
Unlike regression models like 'lm()', 'glm()' or 'gam()', 'lgcp()',
`inlabru` does not automatically remove the first level and absorb it into
an intercept. Instead, we can either use `model="factor_full"` without an intercept,
or `model="factor_contrast"`, which does remove the first level.

```{r warning=FALSE,message=FALSE}
comp1alt <- coordinates ~ vegetation(gcov$vegetation, model = "factor_contrast") + Intercept(1)
```

Fit the model as usual:
```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit1 <- lgcp(comp1, nests, samplers = boundary, domain = list(coordinates = mesh))
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit1.rds")
if ((!.cache_recompute) &&
    file.exists(.cache_file)) {
  fit1 <- readRDS(file = .cache_file)
} else {
  fit1 <- lgcp(comp1, nests, samplers = boundary, domain = list(coordinates = mesh))
  saveRDS(fit1, .cache_file)
}
```

Predict the intensity, and plot the median intensity surface. (In older versions, predicting takes some
time because we did not have vegetation values outside the mesh so 'inlabru' needed 
to predict these first. Since v2.0.0, the vegetation has been pre-extended.)

The `predidct` function of `inlabru` takes into its `data` argument a `SpatialPointsDataFrame`, 
a `SpatialPixelsDataFrame` or a `data.frame`. We can use the `inlabru` function `pixels` to generate
a `SpatialPixelsDataFrame` only within the boundary, using its `mask` argument, as shown below.


```{r warning=FALSE,message=FALSE}
df <- pixels(mesh, mask = boundary)
int1 <- predict(fit1, data = df, ~ exp(vegetation))

ggplot() +
  gg(int1) +
  gg(boundary, alpha = 0, lwd = 2) +
  gg(nests, color = "DarkGreen") +
  coord_equal()
```

Not surprisingly, given that most nests are in `Primary` vegetation, the high density 
is in this vegetation. But there are substantial patches of predicted high density 
that have no nests, and some areas of predicted low density that have nests. What 
about the estimated abundance (there are really 647 nests there):

```{r warning=FALSE,message=FALSE}
ips <- ipoints(boundary, mesh)
Lambda1 <- predict(fit1, ips, ~ sum(weight * exp(vegetation)))
Lambda1
```

#### A model with vegetation type and a SPDE type smoother
Lets try to `explain` the pattern in nest distribution that is not captured by 
the vegetation covariate, using an SPDE:


```{r results='hide',warning=FALSE,message=FALSE}
pcmatern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(5, 0.01)
)

comp2 <- coordinates ~
-1 +
  vegetation(gcov$vegetation, model = "factor_full") +
  mySmooth(coordinates, model = pcmatern)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
fit2 <- lgcp(comp2, nests, samplers = boundary, domain = list(coordinates = mesh))
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "fit2.rds")
if ((!.cache_recompute) &&
    file.exists(.cache_file)) {
  fit2 <- readRDS(file = .cache_file)
} else {
  fit2 <- lgcp(comp2, nests, samplers = boundary, domain = list(coordinates = mesh))
  saveRDS(fit2, .cache_file)
}
```

And plot the median intensity surface

```{r warning=FALSE,message=FALSE}
df <- pixels(mesh, mask = boundary)
int2 <- predict(fit2, df, ~ exp(mySmooth + vegetation), n.samples = 1000)

ggplot() +
  gg(int2, aes(fill = q0.025)) +
  gg(boundary, alpha = 0, lwd = 2) +
  gg(nests) +
  coord_equal()
```

... and the expected integrated intensity (mean of abundance)


```{r warning=FALSE,message=FALSE}
Lambda2 <- predict(
  fit2,
  ipoints(boundary, mesh),
  ~ sum(weight * exp(mySmooth + vegetation))
)
Lambda2
```

Look at the contributions to the linear predictor from the SPDE and from vegetation:
```{r warning=FALSE,message=FALSE}
lp2 <- predict(fit2, df, ~ list(
  smooth_veg = mySmooth + vegetation,
  smooth = mySmooth,
  veg = vegetation
))
```

The function `scale_fill_gradientn` sets the scale 
for the plot legend. Here we set it to span the range of the three linear predictor
components being plotted (medians are plotted by default).


```{r warning=FALSE,message=FALSE}
lprange <- range(lp2$smooth_veg$median, lp2$smooth$median, lp2$veg$median)
csc <- scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), limits = lprange)

plot.lp2 <- ggplot() +
  gg(lp2$smooth_veg) +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("mySmooth + vegetation") +
  coord_equal()

plot.lp2.spde <- ggplot() +
  gg(lp2$smooth) +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("mySmooth") +
  coord_equal()

plot.lp2.veg <- ggplot() +
  gg(lp2$veg) +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("vegetation") +
  coord_equal()

multiplot(plot.lp2, plot.lp2.spde, plot.lp2.veg, cols = 3)
```

#### A model with SPDE only

Do we need vegetation at all? Fit a model with only an SPDE + Intercept, and choose
between models on the basis of DIC, using 'deltaIC()'.


```{r results='hide',warning=FALSE,message=FALSE,eval=FALSE}
comp3 <- coordinates ~ mySmooth(coordinates, model = pcmatern) + Intercept(1)
fit3 <- lgcp(comp3,
  data = nests,
  samplers = boundary,
  domain = list(coordinates = mesh)
)
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
comp3 <- coordinates ~ mySmooth(coordinates, model = pcmatern) + Intercept(1)
.cache_file <- file.path(.cache_dir, "fit3.rds")
if ((!.cache_recompute) &&
    file.exists(.cache_file)) {
  fit3 <- readRDS(file = .cache_file)
} else {
  fit3 <- lgcp(comp3, nests, samplers = boundary, domain = list(coordinates = mesh))
  saveRDS(fit3, .cache_file)
}
```

```{r results='hide',warning=FALSE,message=FALSE}
int3 <- predict(fit3, df, ~ exp(mySmooth + Intercept))

ggplot() +
  gg(int3) +
  gg(boundary, alpha = 0) +
  gg(nests) +
  coord_equal()
```
```{r warning=FALSE,message=FALSE}
Lambda3 <- predict(
  fit3,
  ipoints(boundary, mesh),
  ~ sum(weight * exp(mySmooth + Intercept))
)
Lambda3

deltaIC(fit1, fit2, fit3)
```

#### CV and SPDE parameters for Model 2

We are going with Model `fit2`. Lets look at the spatial distribution of the 
coefficient of variation

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() +
  gg(int2["cv"]) +
  gg(boundary, alpha = 0) +
  gg(nests) +
  coord_fixed()
```

Plot the vegetation "fixed effect" posteriors. First get their names - from `$marginals.random$vegetation` of the fitted object, 
which contains the fixed effect marginal distribution data


```{r warning=FALSE,message=FALSE}
flist <- vector("list", NROW(fit2$summary.random$vegetation))
for (i in seq_along(flist)) flist[[i]] <- plot(fit2, "vegetation", index = i)
multiplot(plotlist = flist, cols = 3)
```

Use `spde.posterior( )` to obtain and then plot the SPDE parameter posteriors and the 
Matern correlation and covariance functions
for this model.


```{r results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
spde.range <- spde.posterior(fit2, "mySmooth", what = "range")
spde.logvar <- spde.posterior(fit2, "mySmooth", what = "log.variance")
range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)

multiplot(range.plot, var.plot)
corplot <- plot(spde.posterior(fit2, "mySmooth", what = "matern.correlation"))
covplot <- plot(spde.posterior(fit2, "mySmooth", what = "matern.covariance"))
multiplot(covplot, corplot)
```

Continuous covariates
-----------------------------------

Now lets try a model with elevation as a (continuous) explanatory variable. (First centre elevations
for more stable fitting.)


```{r results="hide",warning=FALSE,message=FALSE}
elev <- gcov$elevation
elev$elevation <- elev$elevation - mean(elev$elevation, na.rm = TRUE)

ggplot() +
  gg(elev) +
  gg(boundary, alpha = 0) +
  coord_fixed()
```

The elevation variable here is of class 'SpatialGridDataFrame', that can be
handled in the same way as the vegetation covariate. However, since in some
cases data may be stored differently, and other methods are needed to access
the stored values. In such cases, we can define a function that knows how to
evaluate the covariate at arbitrary points in the survey region, and call that
function in the component definition. In this case, we can use a powerful
method from the 'sp' package to do this. We use this to create the needed
function.


```{r results="hide",warning=FALSE,message=FALSE}
f.elev <- function(x, y) {
  # turn coordinates into SpatialPoints object:
  # with the appropriate coordinate reference system (CRS)
  spp <- SpatialPoints(data.frame(x = x, y = y), proj4string = fm_sp_get_crs(elev))
  proj4string(spp) <- fm_sp_get_crs(elev)
  # Extract elevation values at spp coords, from our elev SpatialGridDataFrame
  v <- over(spp, elev)
  if (any(is.na(v$elevation))) {
    v$elevation <- inlabru:::bru_fill_missing(elev, spp, v$elevation)
  }
  return(v$elevation)
}
```

For brevity we are not going to consider models with elevation only, with elevation
and a SPDE, and with SPDE only. We will just fit one with elevation and SPDE.
We create our model to pass to lgcp thus:


```{r results="hide",warning=FALSE,message=FALSE}
matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(5, 0.01)
)

ecomp <- coordinates ~ elev(f.elev(x, y), model = "linear") +
  mySmooth(coordinates, model = matern) + Intercept(1)
```

Note how the elevation effect is defined. When we used the `Spatial` grid
object directly we specified it like
```{r results="hide",warning=FALSE,message=FALSE,eval=FALSE}
vegetation(gcov$vegetation, model = "factor_full")
```
whereas with the function method we specify the covariate like this:
```{r results="hide",warning=FALSE,message=FALSE,eval=FALSE}
elev(f.elev(x, y), model = "linear")
```

We also now include an intercept term.

The model is fitted in the usual way:
```{r results="hide",warning=FALSE,message=FALSE,eval=FALSE}
efit <- lgcp(ecomp, nests, samplers = boundary, domain = list(coordinates = mesh))
```
```{r results='hide',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE}
.cache_file <- file.path(.cache_dir, "efit.rds")
if ((!.cache_recompute) &&
    file.exists(.cache_file)) {
  efit <- readRDS(file = .cache_file)
} else {
  efit <- lgcp(ecomp, nests, samplers = boundary, domain = list(coordinates = mesh))
  saveRDS(efit, .cache_file)
}
```

Summary and model selection


```{r warning=FALSE,message=FALSE}
summary(efit)
deltaIC(fit1, fit2, fit3, efit)
```

Predict and plot the density


```{r results="hide",warning=FALSE,message=FALSE}
e.int <- predict(efit, pixels(mesh, mask = boundary), ~ exp(mySmooth + elev + Intercept))

ggplot() +
  gg(e.int) +
  gg(boundary, alpha = 0) +
  gg(nests, shape = "+") +
  coord_equal()
```

Now look at the elevation and SPDE effects in space. Leave out the Intercept
because it swamps the spatial effects of elevation and the SPDE in the 
plots and we are interested in comparing the effects of elevation and the SPDE.

First we need to predict on the linear predictor scale.
```{r warning=FALSE,message=FALSE,echo=TRUE}
e.lp <- predict(
  efit, pixels(mesh, mask = boundary),
  ~ list(
    smooth_elev = mySmooth + elev,
    elev = elev,
    smooth = mySmooth
  )
)
```

The code below, which is very similar to that used for the vegetation factor
variable, produces the plots we want.


```{r warning=FALSE,message=FALSE, fig.height=5, fig.width=13}
lprange <- range(e.lp$smooth_elev$mean, e.lp$elev$mean, e.lp$smooth$mean)

library(RColorBrewer)
csc <- scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), limits = lprange)

plot.e.lp <- ggplot() +
  gg(e.lp$smooth_elev, mask = boundary) +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("SPDE + elevation") +
  coord_equal()

plot.e.lp.spde <- ggplot() +
  gg(e.lp$smooth, mask = boundary) +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("SPDE") +
  coord_equal()

plot.e.lp.elev <- ggplot() +
  gg(e.lp$elev, mask = boundary) +
  csc +
  theme(legend.position = "bottom") +
  gg(boundary, alpha = 0) +
  ggtitle("elevation") +
  coord_equal()

multiplot(plot.e.lp,
  plot.e.lp.spde,
  plot.e.lp.elev,
  cols = 3
)
```

You might also want to look at the posteriors of the fixed effects and of the SPDE.
Adapt the code used for the vegetation factor to do this.


```{r results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
flist <- vector("list", NROW(efit$summary.fixed))
for (i in seq_along(flist)) {
  flist[[i]] <- plot(efit, rownames(efit$summary.fixed)[i])
}
multiplot(plotlist = flist, cols = 2)
```

Plot the SPDE parameter posteriors and the Matern correlation and covariance functions
for this model.


```{r results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}

spde.range <- spde.posterior(efit, "mySmooth", what = "range")
spde.logvar <- spde.posterior(efit, "mySmooth", what = "log.variance")
range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)

multiplot(range.plot, var.plot)
corplot <- plot(spde.posterior(efit, "mySmooth", what = "matern.correlation"))
covplot <- plot(spde.posterior(efit, "mySmooth", what = "matern.covariance"))
multiplot(covplot, corplot)
```

Also estimate abundance. The `data.frame` in the second call leads to inclusion of `N`
in the prediction object, for easier plotting.

```{r fig.show="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
Lambda <- predict(
  efit, ipoints(boundary, mesh),
  ~ sum(weight * exp(mySmooth + elev + Intercept))
)
Lambda

Nest.e <- predict(
  efit,
  ipoints(boundary, mesh),
  ~ data.frame(
    N = 200:1000,
    density = dpois(200:1000,
      lambda = sum(weight * exp(mySmooth + elev + Intercept))
    )
  ),
  n.samples = 2000
)
```

Plot in the same way as in previous practicals

```{r }
Nest.e$plugin_estimate <- dpois(Nest.e$N, lambda = Lambda$median)
ggplot(data = Nest.e) +
  geom_line(aes(x = N, y = mean, colour = "Posterior")) +
  geom_line(aes(x = N, y = plugin_estimate, colour = "Plugin"))
```

### Non-spatial evaluation of the covariate effect

The previous examples of posterior prediction focused on spatial prediction.
From `inlabru` version 2.2.8, a feauture is available for overriding the
component input value specification from the component definition. Each model
component can be evaluated directly, for arbitrary values by functions named by
adding the suffix `_eval` to the end of the component name in the predictor
expression. Since the elevation effect in this model is linear, the resulting plot isn't very interesting, but the same method can be applied to non-linear effects as well, and combined into general R expressions: 

```{r}
elev.pred <- predict(
  efit,
  data = data.frame(elevation = seq(0, 100, length.out = 1000)),
  formula = ~ elev_eval(elevation)
)

ggplot(elev.pred) +
  geom_line(aes(elevation, mean)) +
  geom_ribbon(aes(elevation,
                  ymin = q0.025,
                  ymax = q0.975),
              alpha = 0.2) +
  geom_ribbon(aes(elevation,
                  ymin = mean - 1 * sd,
                  ymax = mean + 1 * sd),
              alpha = 0.2)
```


A 1D Example
----------------
Try fitting a 1-dimensional model to the point data in the `inlabru` dataset 
`Poisson2_1D`. This comes with a covariate function called `cov2_1D`. Try to reproduce 
the plot below (used in lectures) showing the effects of the `Intercept + z` and the 
`SPDE`. (You may find it helpful to build on the model you fitted in the previous 
practical, adding the covariate to the model specification.)


```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
data(Poisson2_1D)
ss <- seq(0, 55, length = 200)
z <- cov2_1D(ss)
x <- seq(1, 55, length = 100)
mesh <- inla.mesh.1d(x, degree = 1)

comp <- x ~
  beta_z(cov2_1D(x), model = "linear") +
  spde1D(x, model = inla.spde2.matern(mesh)) +
  Intercept(1)

fitcov1D <- lgcp(comp, pts2, domain = list(x = mesh))
pr.df <- data.frame(x = x)
prcov1D <- predict(
  fitcov1D, pr.df,
  ~ list(
    total = exp(beta_z + spde1D + Intercept),
    fx = exp(beta_z + Intercept),
    spde = exp(spde1D)
  )
)

ggplot() +
  gg(prcov1D$total, color = "red") +
  geom_line(aes(x = prcov1D$spde$x, y = prcov1D$spde$median), col = "blue", lwd = 1.25) +
  geom_line(aes(x = prcov1D$fx$x, y = prcov1D$fx$median), col = "green", lwd = 1.25) +
  geom_point(data = pts2, aes(x = x), y = 0.2, shape = "|", cex = 4) +
  xlab(expression(bold(s))) +
  ylab(expression(hat(lambda)(bold(s)) ~ ~"and its components")) +
  annotate(geom = "text", x = 40, y = 6, label = "Intensity", color = "red") +
  annotate(geom = "text", x = 40, y = 5.5, label = "z-effect", color = "green") +
  annotate(geom = "text", x = 40, y = 5, label = "SPDE", color = "blue")
```
