---
title: "LGCPs - Distance sampling"
author: "David Borchers and Finn Lindgren"
date: "Generated on `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LGCPs - Distance sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
```



Introduction
----------------
We're going to estimate distribution and abundance from a line transect survey of dolphins in the Gulf of Mexico. These data are 
also available in the `R` package `dsm` (where they go under the name `mexdolphins`). In `inlabru` the data are called `mexdolphin`.

Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(ggplot2)
```

Get the data
-----------------------------------------------------------------
 
We'll start by loading the data and extracting the mesh (for convenience).

```{r results="hide",warning=FALSE,message=FALSE}
data(mexdolphin, package = "inlabru")
mesh <- mexdolphin$mesh
```

Plot the data (the initial code below is just to get rid of tick marks)


```{r results="hide",warning=FALSE,message=FALSE}
noyticks <- theme(
  axis.text.y = element_blank(),
  axis.ticks = element_blank()
)

noxticks <- theme(
  axis.text.x = element_blank(),
  axis.ticks = element_blank()
)

ggplot() +
  gg(mexdolphin$ppoly) +
  gg(mexdolphin$samplers, color = "grey") +
  gg(mexdolphin$points, size = 0.2, alpha = 1) +
  noyticks +
  noxticks +
  theme(legend.key.width = unit(x = 0.2, "cm"), legend.key.height = unit(x = 0.3, "cm")) +
  theme(legend.text = element_text(size = 6)) +
  guides(fill = FALSE) +
  coord_equal()
```

Spatial model with a half-normal detection function
-----------------------------------------------------------------
The `samplers` in this dataset are lines, not polygons, so we need to tell `inlabru` about the
strip half-width, `W`, which in the case of these data is 8.
We start by plotting the distances and histogram of frequencies in distance intervals:

```{r fig.width=10, fig.heigth=4}
W <- 8
ggplot(data.frame(mexdolphin$points)) +
  geom_histogram(aes(x = distance),
    breaks = seq(0, W, length = 9),
    boundary = 0, fill = NA, color = "black"
  ) +
  geom_point(aes(x = distance), y = 0, pch = "|", cex = 4)
```

We need to define a half-normal detection probability function. This must take distance as its first
argument and the linear predictor of the sigma parameter (which we will call `lsig`) as its second:

```{r results="hide",warning=FALSE,message=FALSE, eval = TRUE}
hn <- function(distance, lsig) {
  exp(-0.5 * (distance / exp(lsig))^2)
}
```

Specify and fit an SPDE model to these data using a half-normal detection function form.
We need to define a (Matern) covariance function for the SPDE

```{r results="hide",warning=FALSE,message=FALSE, echo = TRUE}
matern <- inla.spde2.pcmatern(mexdolphin$mesh,
  prior.sigma = c(2, 0.01),
  prior.range = c(50, 0.01)
)
```

We need to now separately define the components of the model 
(the SPDE, the Intercept and the detection function parameter `lsig`)

```{r results="hide",warning=FALSE,message=FALSE, echo = TRUE}
cmp <- ~ mySPDE(main = coordinates, model = matern) +
  lsig(1) + Intercept(1)
```

... and the formula, which describes how these components are combined to 
form the linear predictor (remembering that we need an offset due to the unknown
direction of the detections!):

```{r results="hide",warning=FALSE,message=FALSE, echo = TRUE}
form <- coordinates + distance ~ mySPDE +
  log(hn(distance, lsig)) +
  Intercept + log(2)
```

Then fit the model, passing both the components and the formula 
(previously the formula was constructed invisibly by `inlabru`),
and specify integration domains for the spatial and distance dimensions:

```{r results="hide",warning=FALSE,message=FALSE, echo = TRUE}
fit <- lgcp(
  components = cmp,
  mexdolphin$points,
  samplers = mexdolphin$samplers,
  domain = list(
    coordinates = mesh,
    distance = INLA::inla.mesh.1d(seq(0, 8, length.out = 30))
  ),
  formula = form
)
```

Look at the SPDE parameter posteriors
 

```{r }
spde.range <- spde.posterior(fit, "mySPDE", what = "range")
plot(spde.range)
spde.logvar <- spde.posterior(fit, "mySPDE", what = "log.variance")
plot(spde.logvar)
```

Predict spatial intensity, and plot it:

```{r results="hide",warning=FALSE,message=FALSE, echo = TRUE, fig.width=10, fig.heigth=4}
pxl <- fm_pixels(mesh, nx = 100, ny = 50, mask = mexdolphin$ppoly, format = "sp")
pr.int <- predict(fit, pxl, ~ exp(mySPDE + Intercept))

ggplot() +
  gg(pr.int) +
  gg(mexdolphin$ppoly) +
  gg(mexdolphin$samplers, color = "grey") +
  gg(mexdolphin$points, size = 0.2, alpha = 1) +
  noyticks +
  noxticks +
  theme(legend.key.width = unit(x = 0.2, "cm"), legend.key.height = unit(x = 0.3, "cm")) +
  theme(legend.text = element_text(size = 6)) +
  guides(fill = FALSE) +
  coord_equal()
```

Predict the detection function and plot it, to generate a plot like the one below.
Here, we should make sure that it doesn't try to evaluate the effects of
components that can't be evaluated using the given input data. Here, we're only
providing distances and no spatial coordinates, so we cannot evaluate the
spatial random field in this `predict()` call. We can specify this by providing
a vector of component names to include in the prediction calculations, here only
"lsig", with `include = "lsig"`. See `?predict.bru` for more information.

```{r results="hide",warning=FALSE,message=FALSE, echo = TRUE, eval=TRUE, fig.width=10, fig.heigth=4}
distdf <- data.frame(distance = seq(0, 8, length = 100))
dfun <- predict(fit, distdf, ~ hn(distance, lsig), include = "lsig")
plot(dfun)
```
The average detection probability within the maximum detection distance is estimated to be
`r mean(dfun$mean)`.


We can look at the posterior for expected number of dolphins as usual:

```{r warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
predpts <- fm_int(mexdolphin$mesh, mexdolphin$ppoly)
Lambda <- predict(fit, predpts, ~ sum(weight * exp(mySPDE + Intercept)))
Lambda
```

and including the randomness about the expected number. In this case, it turns
out that you need lots of posterior samples, e.g. 2,000 to smooth out the Monte Carlo
error in the posterior, and this takes a little while to compute:

```{r results="hide",warning=FALSE,message=FALSE, echo=TRUE}
Ns <- seq(50, 450, by = 1)
Nest <- predict(fit, predpts,
  ~ data.frame(
    N = Ns,
    density = dpois(Ns,
      lambda = sum(weight * exp(mySPDE + Intercept))
    )
  ),
  n.samples = 2000
)

Nest$plugin_estimate <- dpois(Nest$N, lambda = Lambda$mean)
ggplot(data = Nest) +
  geom_line(aes(x = N, y = mean, colour = "Posterior")) +
  geom_ribbon(
    aes(
      x = N,
      ymin = mean - 2 * mean.mc_std_err,
      ymax = mean + 2 * mean.mc_std_err,
      colour = NULL, fill = "Posterior"
    ),
    alpha = 0.2
  ) +
  geom_line(aes(x = N, y = plugin_estimate, colour = "Plugin", fill = "Plugin"))
```

Hazard-rate Detection Function
-----------------------------------------------------------------
Try doing this all again, but use this hazard-rate detection function model:

```{r results="hide",warning=FALSE,message=FALSE, echo=TRUE}
hr <- function(distance, lsig) {
  1 - exp(-(distance / exp(lsig))^-1)
}
```

Solution:

```{r, results="hide", warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
formula1 <- coordinates + distance ~ mySPDE +
  log(hr(distance, lsig)) +
  Intercept + log(2)

fit1 <- lgcp(
  components = cmp,
  mexdolphin$points,
  samplers = mexdolphin$samplers,
  domain = list(
    coordinates = mesh,
    distance = INLA::inla.mesh.1d(seq(0, 8, length.out = 30))
  ),
  formula = formula1
)
```
Plots:
```{r, warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
spde.range <- spde.posterior(fit1, "mySPDE", what = "range")
plot(spde.range)
spde.logvar <- spde.posterior(fit1, "mySPDE", what = "log.variance")
plot(spde.logvar)

pr.int1 <- predict(fit1, pxl, ~ exp(mySPDE + Intercept))

ggplot() +
  gg(pr.int1) +
  gg(mexdolphin$ppoly) +
  gg(mexdolphin$samplers, color = "grey") +
  gg(mexdolphin$points, size = 0.2, alpha = 1) +
  noyticks +
  noxticks +
  theme(legend.key.width = unit(x = 0.2, "cm"), legend.key.height = unit(x = 0.3, "cm")) +
  theme(legend.text = element_text(size = 6)) +
  guides(fill = FALSE) +
  coord_equal()

distdf <- data.frame(distance = seq(0, 8, length = 100))
dfun1 <- predict(fit1, distdf, ~ hr(distance, lsig), include = "lsig")
plot(dfun1)

predpts <- fm_int(mexdolphin$mesh, mexdolphin$ppoly)
Lambda1 <- predict(fit1, predpts, ~ sum(weight * exp(mySPDE + Intercept)))
Lambda1

Ns <- seq(50, 650, by = 1)
Nest1 <- predict(
  fit1,
  predpts,
  ~ data.frame(
    N = Ns,
    density = dpois(Ns,
      lambda = sum(weight * exp(mySPDE + Intercept))
    )
  ),
  n.samples = 2000
)

Nest1$plugin_estimate <- dpois(Nest1$N, lambda = Lambda1$mean)
ggplot(data = Nest1) +
  geom_line(aes(x = N, y = mean, colour = "Posterior")) +
  geom_ribbon(
    aes(
      x = N,
      ymin = mean - 2 * mean.mc_std_err,
      ymax = mean + 2 * mean.mc_std_err,
      colour = NULL, fill = "Posterior"
    ),
    alpha = 0.2
  ) +
  geom_line(aes(x = N, y = plugin_estimate, colour = "Plugin", fill = "Plugin"))
```

Comparing the models
-----------------------


Look at the goodness-of-fit of the two models in the distance dimension
```{r,warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
bc <- bincount(
  result = fit,
  observations = mexdolphin$points$distance,
  breaks = seq(0, max(mexdolphin$points$distance), length = 9),
  predictor = distance ~ hn(distance, lsig)
)
attributes(bc)$ggp

bc1 <- bincount(
  result = fit1,
  observations = mexdolphin$points$distance,
  breaks = seq(0, max(mexdolphin$points$distance), length = 9),
  predictor = distance ~ hn(distance, lsig)
)
attributes(bc1)$ggp
```

Fit Models only to the distance sampling data
-----------------------------------------------------------------
Half-normal first

```{r,results="hide",warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
formula <- distance ~ log(hn(distance, lsig)) + Intercept
cmp <- ~ lsig(1) + Intercept(1)
dfit <- lgcp(
  components = cmp,
  mexdolphin$points,
  domain = list(distance = INLA::inla.mesh.1d(seq(0, 8, length.out = 30))),
  formula = formula,
  options = list(bru_initial = list(lsig = 1, Intercept = 3))
)
detfun <- predict(dfit, distdf, ~ hn(distance, lsig), include = "lsig")
```

Half-normal next

```{r,results="hide", warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
formula1 <- distance ~ log(hr(distance, lsig)) + Intercept
cmp <- ~ lsig(1) + Intercept(1)
dfit1 <- lgcp(
  components = cmp,
  mexdolphin$points,
  domain = list(distance = INLA::inla.mesh.1d(seq(0, 8, length.out = 30))),
  formula = formula1
)
detfun1 <- predict(dfit1, distdf, ~ hr(distance, lsig), include = "lsig")
```


Plot both lines on histogram of observations
First scale lines to have same area as that of histogram
Half-normal:

```{r,warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
hnline <- data.frame(distance = detfun$distance, p = detfun$mean, lower = detfun$q0.025, upper = detfun$q0.975)
wts <- diff(hnline$distance)
wts[1] <- wts[1] / 2
wts <- c(wts, wts[1])
hnarea <- sum(wts * hnline$p)
n <- length(mexdolphin$points$distance)
scale <- n / hnarea
hnline$En <- hnline$p * scale
hnline$En.lower <- hnline$lower * scale
hnline$En.upper <- hnline$upper * scale
```

Hazard-rate:

```{r,warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
hrline <- data.frame(distance = detfun1$distance, p = detfun1$mean, lower = detfun1$q0.025, upper = detfun1$q0.975)
wts <- diff(hrline$distance)
wts[1] <- wts[1] / 2
wts <- c(wts, wts[1])
hrarea <- sum(wts * hrline$p)
n <- length(mexdolphin$points$distance)
scale <- n / hrarea
hrline$En <- hrline$p * scale
hrline$En.lower <- hrline$lower * scale
hrline$En.upper <- hrline$upper * scale
```

Combine lines in a single object for plotting

```{r,warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
dlines <- rbind(
  cbind(hnline, model = "Half-normal"),
  cbind(hrline, model = "Hazard-rate")
)
```

Plot without the 95% credible intervals

```{r,warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
ggplot(data.frame(mexdolphin$points)) +
  geom_histogram(aes(x = distance), breaks = seq(0, 8, length = 9), alpha = 0.3) +
  geom_point(aes(x = distance), y = 0.2, shape = "|", size = 3) +
  geom_line(data = dlines, aes(x = distance, y = En, group = model, col = model))
```

Plot with the 95% credible intervals (without taking the count rescaling into account)
```{r,warning=FALSE,message=FALSE, eval=TRUE, echo=TRUE}
ggplot(data.frame(mexdolphin$points)) +
  geom_histogram(aes(x = distance), breaks = seq(0, 8, length = 9), alpha = 0.3) +
  geom_point(aes(x = distance), y = 0.2, shape = "|", size = 3) +
  geom_line(data = dlines, aes(x = distance, y = En, group = model, col = model)) +
  geom_ribbon(
    data = dlines, aes(x = distance, ymin = En.lower, ymax = En.upper, group = model, col = model, fill = model),
    alpha = 0.2, lty = 2
  )
```
