---
title: "LGCPs - Plot sampling"
author: "David Borchers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Introduction
----------

This practical demonstrates use of the `samplers` argument in `lgcp`, which you
need to use when you have observed points from only a sample of plots in the survey 
region. 


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(mgcv)
```

Use empricial Bayes in order to speed things up a little bit:
```{r results="hide",warning=FALSE,message=FALSE}
init.tutorial()
```

Get the data
-----------------------------------

```{r }
data(gorillas, package = "inlabru")
```

This dataset is a list (see `help(gorillas)` for details. Extract the the objects 
you need from the list, for convenience:

```{r }
nests <- gorillas$nests
mesh <- gorillas$mesh
boundary <- gorillas$boundary
gcov <- gorillas$gcov
```

The `gorillas` data also contains a plot sample subset which covers 60% of the survey region.

```{r }
sample <- gorillas$plotsample

```{r results="hide",warning=FALSE,message=FALSE}
plotdets <- ggplot() + gg(boundary) + gg(sample$plots) + 
  gg(sample$nests, pch = "+", cex = 4, color = "red") +
  geom_text(aes(label = sample$counts$count, x = sample$counts$x, y = sample$counts$y)) + 
  coord_fixed() + labs(x = "Easting", y = "Northing")
plot(plotdets)
```

On this plot survey, only points within the rectangles are detected, but it is also 
informative to plot all the points here (which if it was a real plot survey you could
not do, because you would not have seen them all).


```{r results="hide",warning=FALSE,message=FALSE}

plotwithall <- ggplot() + gg(boundary) + gg(sample$plots) + 
  gg(nests, pch="+", cex = 4, color = "blue") +
  geom_text(aes(label = sample$counts$count, x = sample$counts$x, y = sample$counts$y)) + 
  gg(sample$nests, pch = "+", cex = 4, color = "red") +
  coord_fixed() + labs(x = "Easting", y = "Northing")
plot(plotwithall)
```

Inference
------------------------------------
The observed nest locations are in the SpatialPointsDataFrame `sample$nests`, and the 
plots are in the SpatialPolygonsDataFrame `sample$plots`. Again, we are using the following SPDE
setup:


```{r }
matern <- inla.spde2.pcmatern(mesh, 
                              prior.sigma = c(0.1, 0.01), 
                              prior.range = c(5, 0.01))
```


Fit an LGCP model with SPDE only to these data by using the `samplers=` argument of 
the function `lgcp( )`:


```{r results="hide",warning=FALSE,message=FALSE}
cmp <- coordinates ~ my.spde(map = coordinates, 
                             model = matern) 

fit <- lgcp(cmp, sample$nests, samplers = sample$plots)
```


Plot the density surface from your fitted model

```{r results="hide",warning=FALSE,message=FALSE}
lambda.sample <- predict(fit, pixels(mesh), ~ exp(my.spde + Intercept))
```

```{r warning=FALSE,message=FALSE}
lambda.sample.plot <- ggplot() + 
                        gg(lambda.sample) + 
                        gg(sample$plots) + 
                        gg(boundary, col = "yellow") + 
                        coord_fixed() + 
                        theme(legend.position = "bottom")

lambda.sample.plot
```


Estimate the integrated intensity lambda.
```{r results="hide",warning=FALSE,message=FALSE}
Lambda <- predict(fit, ipoints(boundary, mesh), ~ sum(weight * exp(my.spde + Intercept)))
Lambda
```


Fit the same model to the full dataset (the points in `gorillas$nests`), or get your previous 
fit, if you kept it. Plot the intensity surface and estimate the integrated intensity

```{r results="hide",warning=FALSE,message=FALSE}

fit.all <- lgcp(cmp, gorillas$nests, samplers = gorillas$boundary)
lambda.all <- predict(fit.all, pixels(mesh), ~ exp(my.spde + Intercept))
Lambda.all <- predict(fit.all,  ipoints(boundary, mesh), ~ sum(weight * exp(my.spde + Intercept)))
```

Your plot should look like this:


```{r results="hide",warning=FALSE,message=FALSE,fig.height=5,fig.width=11, echo=FALSE}
lambda.all.plot <- ggplot() + gg(lambda.all) + 
                                gg(sample$plots) + 
                                gg(boundary, col = "yellow") + 
                                coord_fixed() + 
                                theme(legend.position = "bottom")
lambda.all.plot
```

And `Lambda.all` should be close to

```{r }
Lambda
```

Now, let's compare the results


```{r results="hide",warning=FALSE,message=FALSE,fig.height=5,fig.width=11}
multiplot(lambda.sample.plot, lambda.all.plot, cols = 2)
```

Do you understand the reason for the differences in the posteriors of the two abundance estimates?
