---
title: "LGCPs - An example in two dimensions"
author: "David Borchers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
```

Introduction
----------------
For this vignette we are going to be working with a dataset obtained from 
the `R` package `spatstat`. We will set up a two-dimensional LGCP to estimate Gorilla abundance.


Setting things up
----------------

Load libraries
```{r results="hide",warning=FALSE,message=FALSE}
library(inlabru)
library(INLA)
library(mgcv)
```

Get the data
-----------------------------------
For the next few practicals we are going to be working with a dataset obtained from 
the `R` package `spatstat`, which contains the locations of 647 gorilla nests. We load the 
dataset like this:

```{r }
data(gorillas, package = "inlabru")
```

This dataset is a list containing a number of `R` objects, including the locations of the 
nests, the boundary of the survey area and an `INLA` mesh - see `help(gorillas)` for 
details. Extract the the objects we need from the list, into other objects, so that we 
don't have to keep typing '`gorillas$`':

```{r }
nests <- gorillas$nests
mesh <- gorillas$mesh
boundary <- gorillas$boundary
```

Plot the points (the nests(. (The `ggplot2` function `coord_fixed()` sets the aspect ratio, 
which defaults to 1.)

```{r results="hide",warning=FALSE,message=FALSE}
ggplot() +
  gg(mesh) +
  gg(nests) +
  gg(boundary) +
  coord_fixed() +
  ggtitle("Points")
```


Fiting the model
-----------------------------------

Fit an LGCP model to the locations of the gorilla nests, predict on the survey region,
and produce a plot of the estimated density - which should look like the 
plot shown below.

Recall that the steps to specifying, fitting and predicting are:

1. Specify a model, comprising (for 2D models) `coordinates` on the left of `~` and 
an SPDE `+ Intercept` on the right. Please use the SPDE prior specification stated below.

2. Call `lgcp( )`, passing it (with 2D models) the model components, the `SpatialPointsDataFrame` 
containing the observed points and the `SpatialPolygonsDataFrame` defining the survey boundary
using the `samplers` argument.

3. Call `predict( )`, passing it the fitted model from 2., locations at which to predict and
an appropriate predictor spcification. The locations at which to predict should be a 
`SpatialPixelsDataFrame` covering the mesh obtained by calling  `pixels(mesh)`.


```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = c(0.1, 0.01),
  prior.range = c(5, 0.01)
)

cmp <- coordinates ~ mySmooth(coordinates,
  model = matern
) +
  Intercept

fit <- lgcp(cmp, nests, samplers = boundary, domain = list(coordinates = mesh))
```

Predicting intensity
-----------------------------------

You should get a plot like that below (the command below assumes that the prediction
is in an object called `lambda`):

```{r results="hide",warning=FALSE,message=FALSE, fig.height=4,fig.width=9}
pred <- predict(
  fit,
  pixels(mesh, mask = gorillas$boundary),
  ~ data.frame(
    lambda = exp(mySmooth + Intercept),
    loglambda = mySmooth + Intercept
  )
)

pl1 <- ggplot() +
  gg(pred$lambda) +
  gg(boundary) +
  ggtitle("LGCP fit to Points", subtitle = "(Response Scale)") +
  coord_fixed()

pl2 <- ggplot() +
  gg(pred$loglambda) +
  gg(boundary) +
  ggtitle("LGCP fit to Points", subtitle = "(Linear Predictor Scale)") +
  coord_fixed()

multiplot(pl1, pl2, cols = 2)
```

You can plot the median, lower 95% and upper 95% density surfaces as follows (assuming
that the predicted intensity is in object `lambda`).

```{r warning=FALSE,message=FALSE,fig.width=9,fig.height=4}
ggplot() +
  gg(cbind(pred$lambda, data.frame(property = "q0.500")), aes(fill = median)) +
  gg(cbind(pred$lambda, data.frame(property = "q0.025")), aes(fill = q0.025)) +
  gg(cbind(pred$lambda, data.frame(property = "q0.975")), aes(fill = q0.975)) +
  coord_equal() +
  facet_wrap(~property)
```


SPDE parameters 
-----------------------------------

Plot the SPDE parameter and fixed effect parameter posteriors.

```{r results="hide",warning=FALSE,message=FALSE}
int.plot <- plot(fit, "Intercept")
spde.range <- spde.posterior(fit, "mySmooth", what = "range")
spde.logvar <- spde.posterior(fit, "mySmooth", what = "log.variance")
range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)

multiplot(range.plot, var.plot, int.plot)
```

Look at the correlation function if you want to:

```{r eval=TRUE,warning=FALSE,message=FALSE}
corplot <- plot(spde.posterior(fit, "mySmooth", what = "matern.correlation"))
covplot <- plot(spde.posterior(fit, "mySmooth", what = "matern.covariance"))
multiplot(covplot, corplot)
```


Estimating Abundance
-----------------------------------

Finally, estimate abundance using the `predict` function. As a first step we need an estimate
 for the integrated lambda. The integration `weight` values are contained in the `ipoints`
 output. 
 

```{r eval=TRUE,warning=FALSE,message=FALSE,echo=TRUE}
Lambda <- predict(
  fit,
  ipoints(boundary, mesh),
  ~ sum(weight * exp(mySmooth + Intercept))
)
Lambda
```

Given some generous interval boundaries (500, 800) for lambda we can estimate the posterior abundance distribution via
```{r,warning=FALSE,message=FALSE,echo=TRUE}
Nest <- predict(
  fit, ipoints(boundary, mesh),
  ~ data.frame(
    N = 500:800,
    dpois(500:800,
      lambda = sum(weight * exp(mySmooth + Intercept))
    )
  )
)
```

Get its quantiles via

```{r,warning=FALSE,message=FALSE,echo=TRUE}
inla.qmarginal(c(0.025, 0.5, 0.975), marginal = list(x = Nest$N, y = Nest$mean))
```
... the mean via
```{r,warning=FALSE,message=FALSE,echo=TRUE}
inla.emarginal(identity, marginal = list(x = Nest$N, y = Nest$mean))
```
and plot posteriors:
```{r results="hide",warning=FALSE,message=FALSE,echo=TRUE}
Nest$plugin_estimate <- dpois(Nest$N, lambda = Lambda$mean)
ggplot(data = Nest) +
  geom_line(aes(x = N, y = mean, colour = "Posterior")) +
  geom_line(aes(x = N, y = plugin_estimate, colour = "Plugin"))
```

The true number of nests in 647; the mean and median of the posterior 
distribution of abundance should be close to this if you have not done anything 
wrong!
