---
title: "Devel: Model evaluation flowchart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Devel: Model evaluation flowchart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

(Vignette under construction!)

## Mapping from component inputs and latent states to component effects

```{r, eval=TRUE,echo=FALSE}
DiagrammeR::mermaid('
flowchart TD

subgraph "main, group, replicate"
  input([input])
  mapper([mapper])
end
weights(["weights"])
input_comb(["combined inputs:
  list(mapper = list(main, group, replicate),
  scale = weights)"])
multi_mapper(["mapper_multi<br/>list(main, group, replicate)"])
scale_mapper([mapper_scale])
pipe_mapper([mapper_pipe])
state([state])
lin(["mapper_linearised<br/>(offset, jacobian, state0)"])
effect([effect])

ibm_do_lin{linearise}
ibm_lin[ibm_linear]
ibm_eval[ibm_eval]
ibm_eval_lin[ibm_eval]

%% edge definitions with the node IDs
input --> input_comb
weights --> input_comb
input_comb --> ibm_eval
input_comb --> ibm_lin
pipe_mapper --> ibm_do_lin
ibm_do_lin -->|Yes| ibm_lin
ibm_do_lin -->|No| ibm_eval
state0([state0]) --> ibm_lin
ibm_lin --> lin
state --> ibm_eval_lin
lin --> ibm_eval_lin
state --> ibm_eval
mapper --> multi_mapper --> pipe_mapper
scale_mapper --> pipe_mapper
ibm_eval --> effect
ibm_eval_lin --> effect
',
  width = 600,
  height = 800
)
```

## Linearising a mapping

```{r, eval=TRUE,echo=FALSE}
DiagrammeR::mermaid('
flowchart TD

mapper([mapper])
input([input])
state0([state0])
off(["offset"])
jac(["jacobian"])
lin(["mapper_linearised<br/>(offset, jacobian, state0)"])

ibm_A[ibm_amatrix]
style ibm_A color:#f00
ibm_lin[ibm_linear]
ibm_jac[ibm_jacobian]

%% edge definitions with the node IDs
mapper-->ibm_lin
input-->ibm_lin
state0-->ibm_lin
ibm_lin --> ibm_eval
ibm_eval --- ibm_jac
ibm_lin --> ibm_jac
ibm_eval --> off
ibm_jac --> jac
ibm_jac --- ibm_A
ibm_lin --> lin
%%off-->lin
%%jac-->lin
%%state-->lin
',
  width = 600,
  height = 400
)
```

## Component input evaluation

For each `<label>` of `main`, `group`, `replicate`, and `weights`, the given expression `expr`
is evaluated in the data context, producing the `input` to the component `mapper`.
For spatial covariate inputs, the corresponding `<label>_layer` expression is also evaluated.

Red nodes indicate deprecated behaviour retained for backwards compatibility.

```{r, eval=TRUE,echo=FALSE}
DiagrammeR::mermaid('
flowchart TD
  expr([`label`_expr])
  data([data])
  _envir([.envir])
  input(["`label`_input"])
  input0([input = NULL])
  input1([input = 1])
  style input1 color:#f00

  eval[eval]
  eval_function["value=fun(.data.)"]
  eval_formula["model.Matrix"]
  eval_spatial["eval_spatial"]
  eval_coordinates["SpatialPoints(value, crs)"]
  style eval_coordinates color:#f00
  eval_crs["crs=fm_CRS(.data.)"]

  res{"Type of<br/>result"}
  null_on_fail{"null_on_fail?"}
  is_coordinates_fun{"Is expr =<br/>coordinates?"}

  expr --> eval
  eval --> res
  res -->|error| null_on_fail
  data --> eval
  _envir --> eval

  null_on_fail -->|TRUE| input0
  null_on_fail -->|FALSE| input1
  input0 --> input
  input1 --> input

  res -->|function| eval_function
  eval_function --> is_coordinates_fun

  is_coordinates_fun -->|Yes| eval_coordinates
  is_coordinates_fun -->|No| input
  eval_coordinates --> input
  eval_crs --> eval_coordinates

  res -->|"vector,<br/>matrix,<br/>data.frame<br/>list"| input

  res -->|formula| eval_formula
  eval_formula --> input

  res -->|"spatial<br/>covariate"| eval_spatial
  eval_spatial --> input

%% spatial covariate layer input
  layer_expr(["`label`_layer_expr"])
  layer(["`label`_layer"])
  layer_eval["eval"]
  data --> layer_eval
  _envir --> layer_eval
  layer_expr --> layer_eval --> layer --> eval_spatial
',
  width = 900,
  height = 700
)
```
