---
title: "Converting inla.spde.make.A calls into the bru_mapper system"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting inla.spde.make.A calls into the bru_mapper system}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

(Vignette under construction!)

```{r setup, eval = TRUE, echo=FALSE, include=FALSE}
library(inlabru)
library(INLA)
library(ggplot2)
set.seed(12345L)
```

## Introduction


In traditional `INLA` code involving `inla.mesh` objects and `inla.spde` models,
the `inla.spde.make.A()` function is used to construct the component design matrix that
maps between spatial/spatio-temporal locations and the latent variables associated
with mesh basis functions. For 2-manifold meshes, such as flat and spherical meshes,
the implementation has one latent variable per mesh node, linked to piecewise
linear basis functions on the mesh triangles.  For 1-manifolds such as intervals and
cyclic domains, both piecewise linear and piecewise quadratic basis functions are
supported.

The `inla.spde.make.A()` interface supports a variety of features, that can
be broken down in more simple building blocks. With the `inlabru` `bru_mapper`
system, these building blocks are more easily customised for specific uses, some of
which aren't necessarily connected to spde models, such as the `block` feature
that is used to aggregate rows of a design matrix, e.g. to construct numerical
integration schemes.

If you haven't already, go read the [bru_mapper](http://inlabru-org.github.io/inlabru/articles/bru_mapper.html)
vignette for information about the various `bru_mapper` classes and methods.
Then come back here to continue.

## Converting the basic inla.spde.make.A calls into mappers

The most basic `inla.spde.make.A` call is to map purely spatial points to a mesh:
```{r}
mesh <- inla.mesh.2d(cbind(0, 0), offset = 2, max.edge = 10)
loc <- matrix(runif(10) * 2 - 1, 5, 2)
ggplot() +
  gg(mesh) +
  geom_point(aes(loc[, 1], loc[, 2])) +
  coord_equal()
```

```{r}
A.loc <- inla.spde.make.A(mesh, loc = loc)
A.loc
```

A basic conversion of this becomes
```{r}
A.loc <- fm_evaluator(mesh, loc = loc)$proj$A
A.loc
```
but this is limited to just the basic case of evaluating on only a mesh.

With a `bru_mapper`, this becomes the more generally useful
```{r}
mapper <- bru_mapper(mesh)
A.loc <- ibm_jacobian(mapper, input = loc)
A.loc
```

### Mapping with a precomputed location mapping

```{r}
index <- c(1, 3, 5, 2, 1, 2)
inla.spde.make.A(A.loc = A.loc, index = index)

mapper <- bru_mapper_taylor(jacobian = A.loc[index, , drop = FALSE])
ibm_jacobian(mapper)

# For run-time indexing:
mapper <-
  bru_mapper_pipe(
    list(
      matrix = bru_mapper_taylor(jacobian = A.loc),
      index = bru_mapper_index(nrow(A.loc))
    )
  )
ibm_jacobian(mapper, input = list(index = index))
```

### Group mapping with a group mesh

```{r, eval=FALSE}
inla.spde.make.A(..., group = group.values, group.mesh = group.mesh)

mapper <- bru_mapper_multi(list(
  main = bru_mapper(mesh),
  group = bru_mapper(group.mesh)
))
ibm_jacobian(mapper, input = list(main = loc, group = group.values))
```
### Blockwise aggregation

```{r, eval=FALSE}
block_rescale <- # one of "none", "count", "weights", "sum"
  inla.spde.make.A(...,
    weights = weights,
    block = block,
    block.rescale = block_rescale,
    n.block = n_block
  )

mapper <- bru_mapper_pipe(
  list(
    main = bru_mapper_multi(list(main = bru_mapper(mesh), ...)),
    block = bru_mapper_aggregate(
      rescale = (block_rescale != "none"),
      n_block = n_block
    )
  )
)
ibm_jacobian(mapper,
  input = list(
    main = list(main = loc),
    block = list(block = block, weights = weights)
  )
)
```



## inla.spde.make.index

```{r}
ngroup <- 2
nrepl <- 3

summary(
  as.data.frame(
    inla.spde.make.index("field",
      n.spde = mesh$n,
      n.group = ngroup,
      n.repl = nrepl
    )
  )
)

mapper <- bru_mapper_multi(list(
  field.main = bru_mapper(mesh),
  field.group = bru_mapper_index(ngroup),
  field.replicate = bru_mapper_index(nrepl)
))
summary(ibm_values(mapper, multi = TRUE, inla_f = TRUE))
```
The benefit of the mapper approach here is that it encapsulates all the information,
so that only the mapper needs to be carried around to code that needs it, and
that it doesn't restrict the group and replicate mappings to integer indices;
the index mappers can be replaced by other mappers, e.g. to allow interpolation
between group indices, with a 1d mesh mapper.
