---
title: "Random Fields in 2D"
author: "Finn Lindgren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random Fields in 2D}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5
)
```

Setting things up
----------------

```{r results="hide",warning=FALSE,message=FALSE}
library(INLA)
library(inlabru)
library(mgcv)
```

Make a shortcut to a nicer colour scale:

```{r results="hide",warning=FALSE,message=FALSE}
colsc <- function(...) {
  scale_fill_gradientn(
    colours = rev(RColorBrewer::brewer.pal(11, "RdYlBu")),
    limits = range(..., na.rm = TRUE)
  )
}
```


Modelling on 2D domains
----------------
We will now construct a 2D model, generate a sample of a
random field, and attempt to recover the field from observations at a few locations.
Tomorrow, we will look into more general mesh constructions that adapt to irregular domains.

First, we build a high resolution mesh for the true field, using low level INLA functions

```{r }
bnd <- spoly(data.frame(lon = c(0, 10, 10, 0), lat = c(0, 0, 10, 10)),
  crs = inla.CRS("longlat")
)
mesh_fine <- inla.mesh.2d(boundary = bnd, max.edge = 0.2)
ggplot() +
  gg(mesh_fine) +
  coord_equal()

# Note: the priors here will not be used in estimation
matern_fine <-
  inla.spde2.pcmatern(mesh_fine,
    prior.sigma = c(1, 0.01),
    prior.range = c(1, 0.01)
  )
true_range <- 4
true_sigma <- 1
true_Q <- inla.spde.precision(matern_fine, theta = log(c(true_range, true_sigma)))
```
What is the pointwise standard deviation of the field? Along straight boundaries,
the variance is twice the target variance. At corners the variance is 4 times as large.

```{r results="hide"}
true_sd <- diag(inla.qinv(true_Q))^0.5
ggplot() +
  gg(mesh_fine, col = true_sd) +
  coord_equal()
```

Generate a sample from the model:

```{r }
true_field <- inla.qsample(1, true_Q)[, 1]

truth <- expand.grid(
  lon = seq(0, 10, length = 100),
  lat = seq(0, 10, length = 100)
)
truth$field <- inla.mesh.project(mesh_fine,
  loc = as.matrix(truth),
  field = true_field
)
coordinates(truth) <- c("lon", "lat")
truth <- as(truth, "SpatialPixelsDataFrame")

pl_truth <- ggplot() +
  gg(truth, mapping = aes_string("lon", "lat", fill = "field")) +
  coord_equal() +
  ggtitle("True field")
pl_truth

## Or with another colour scale:
csc <- colsc(truth$field)
multiplot(pl_truth, pl_truth + csc, cols = 2)
```

Extract observations from some random locations:

```{r }
n <- 200
mydata <- data.frame(lon = runif(n, 0, 10), lat = runif(n, 0, 10))
mydata$observed <- inla.mesh.project(mesh_fine,
  loc = as.matrix(mydata),
  field = true_field
) + rnorm(n, sd = 0.4)
coordinates(mydata) <- c("lon", "lat")
ggplot() +
  gg(mydata)
```

Estimating the field
----------

Construct a mesh covering the data:

```{r }
mesh <- inla.mesh.2d(boundary = bnd, max.edge = 0.5)
ggplot() +
  gg(mesh) +
  coord_equal()
```

Construct an SPDE model object for a Matern model:

```{r }
matern <-
  inla.spde2.pcmatern(mesh,
    prior.sigma = c(10, 0.01),
    prior.range = c(1, 0.01)
  )
```

Specify the model components:

```{r }
cmp <- observed ~ field(
  map = coordinates,
  model = matern
) +
  Intercept
```

Fit the model and inspect the results:

```{r warning=FALSE,message=FALSE,results="hide"}
fit <- bru(cmp, mydata, family = "gaussian")
summary(fit)
INLA:::summary.inla(fit)
```

Predict the field on a lattice, and generate
a single realisation from the posterior distribution:

```{r warning=FALSE,message=FALSE}
pix <- pixels(mesh, nx = 200, ny = 200)
pred <- predict(
  fit, pix,
  ~ field + Intercept
)
samp <- generate(fit, pix,
  ~ field + Intercept,
  n.samples = 1
)
pred$sample <- samp[[1]]
```

Compare the truth to the estimated field
(posterior mean and a sample from the posterior distribution):


```{r results="hide",warning=FALSE,message=FALSE, pfig.height=5,fig.width=12}
pl_posterior_mean <- ggplot() +
  gg(pred) +
  gg(bnd) +
  ggtitle("Posterior mean") +
  coord_fixed()
pl_posterior_sample <- ggplot() +
  gg(pred, mapping = aes_string(x = "x", y = "y", fill = "sample")) +
  gg(bnd) +
  ggtitle("Posterior sample") +
  coord_fixed()

# Common colour scale for the truth and estimate:
csc <- colsc(truth$field, pred$mean, pred$sample)
multiplot(pl_truth + csc,
  pl_posterior_mean + csc,
  pl_posterior_sample + csc,
  cols = 3
)
```

Plot the SPDE parameter and fixed effect parameter posteriors.


```{r results="hide",warning=FALSE,message=FALSE}
int.plot <- plot(fit, "Intercept")
spde.range <- spde.posterior(fit, "field", what = "range")
spde.logvar <- spde.posterior(fit, "field", what = "log.variance")
range.plot <- plot(spde.range)
var.plot <- plot(spde.logvar)

multiplot(range.plot, var.plot, int.plot)
```

Look at the correlation function if you want to:


```{r eval=TRUE,warning=FALSE,message=FALSE}
corplot <- plot(spde.posterior(fit, "field", what = "matern.correlation"))
covplot <- plot(spde.posterior(fit, "field", what = "matern.covariance"))
multiplot(covplot, corplot)
```

You can plot the median, lower 95% and upper 95% density surfaces as follows (assuming
that the predicted intensity is in object `pred`).


```{r results="hide",warning=FALSE,message=FALSE}
csc <- colsc(
  pred@data["median"],
  pred@data["q0.025"],
  pred@data["q0.975"]
) ## Common colour scale from SpatialPixelsDataFrame

gmedian <- ggplot() +
  gg(pred["median"]) +
  coord_equal() +
  csc
glower95 <- ggplot() +
  gg(pred["q0.025"]) +
  coord_equal() +
  csc +
  theme(legend.position = "none")
gupper95 <- ggplot() +
  gg(pred["q0.975"]) +
  coord_equal() +
  csc +
  theme(legend.position = "none")

multiplot(gmedian, glower95, gupper95,
  layout = matrix(c(1, 1, 2, 3), byrow = TRUE, ncol = 2)
)
```
